### 2016-04-09 paper.tex
Add Wind Energy template files

% wedoc.tex V1.0, 17 June 2010

\documentclass[times]{weauth}

\usepackage{moreverb}

\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2010}

\begin{document}

\runningheads{A.~N.~Other}{A demonstration of the \journalabb\
class file}

\articletype{RESEARCH ARTICLE}

\title{A demonstration of the \LaTeXe\ class file for
\itshape{\journalnamelc}\footnotemark[2]}

\author{A.~N.~Other}

\address{John Wiley \& Sons, Ltd, The Atrium, Southern Gate, Chichester,
West Sussex, PO19~8SQ, UK}

\corraddr{Journals Production Department, John Wiley \& Sons, Ltd,
The Atrium, Southern Gate, Chichester, West Sussex, PO19~8SQ, UK.}

\begin{abstract}
This paper describes the use of the \LaTeXe\
\textsf{\journalclass} class file for setting papers for
\emph{\journalnamelc}.
\end{abstract}

\keywords{class file; \LaTeXe; \emph{\journalabb}}

\maketitle

\footnotetext[2]{Please ensure that you use the most up to date
class file,
available from the WE Home Page at\\
\href{http://www3.interscience.wiley.com/journal/6276/home}{\texttt{http://www3.interscience.wiley.com/journal/6276/home}}}

\section{Introduction}
Many authors submitting to research journals use \LaTeXe\ to
prepare their papers. This paper describes the
\textsf{\journalclass} class file which can be used to convert
articles produced with other \LaTeXe\ class files into the correct
form for publication in \emph{\journalnamelc}.

The \textsf{\journalclass} class file preserves much of the
standard \LaTeXe\ interface so that any document which was
produced using the standard \LaTeXe\ \textsf{article} style can
easily be converted to work with the \textsf{\journalclassshort}
style. However, the width of text and typesize will vary from that
of \textsf{article.cls}; therefore, \emph{line breaks will change}
and it is likely that displayed mathematics and tabular material
will need re-setting.

In the following sections we describe how to lay out your code to
use \textsf{\journalclass} to reproduce the typographical look of
\emph{\journalnamelc}. However, this paper is not a guide to
using \LaTeXe\ and we would refer you to any of the many books
available (see, for example, \cite{R1,R2,R3}).

\section{The Three Golden Rules}

Before we proceed, we would like to stress \emph{three golden
rules} that need to be followed to enable the most efficient use
of your code at the typesetting stage:
\begin{enumerate}
\item[(i)] keep your own macros to an absolute minimum;

\item[(ii)] as \TeX\ is designed to make sensible spacing
decisions by itself, do \emph{not} use explicit horizontal or
vertical spacing commands, except in a few accepted (mostly
mathematical) situations, such as \verb"\," before a
differential~d, or \verb"\quad" to separate an equation from its
qualifier;

\item[(iii)] follow the \emph{\journalnamelc} reference style.
\end{enumerate}

\section{Getting Started} The \textsf{\journalclassshort} class file should run
on any standard \LaTeXe\ installation. If any of the fonts, style
files or packages it requires are missing from your installation,
they can be found on the \emph{\TeX\ Collection} DVDs or from
CTAN.

\emph{\journalnamelc} is published using Times fonts and this is
achieved by using the \verb"times"
option as\\
\verb"\documentclass[times]{weauth}".

\noindent If for any reason you have a problem using Times you can
easily resort to Computer Modern fonts by removing the
\verb"times" option.

\begin{figure}
\setlength{\fboxsep}{0pt}%
\setlength{\fboxrule}{0pt}%
\begin{center}
\begin{boxedverbatim}
\documentclass[times]{weauth}
%\documentclass[times,doublespace]{weauth}%For paper submission

\begin{document}

\runningheads{<Initials and Surnames>}{<Short title>}

\articletype{<Article type>}

\title{<Initial cap, lower case>}

\author{<An Author\affil{1},
Someone Else\affil{2}\ and Perhaps Another\affil{1}>}

\address{<\affilnum{1}First author's address
(in this example it is the same as the third author)
\affilnum{2}Second author's address>}

\corraddr{<Corresponding author's address>.
E-mail: <corresponding author's email address>}

%\cgs{<Contract/grant sponsor name (no number)>}
%\cgsn{<Contract/grant sponsor name>}{<number>}

\begin{abstract}
<Text>
\end{abstract}

\keywords{<List keywords>}

\maketitle

\section{Introduction}
.
.
.
\end{boxedverbatim}
\end{center}
\caption{Example header text.\label{F1}}
\end{figure}

\section{The Article Header Information}
The heading for any file using \textsf{\journalclass} is shown in
Figure~\ref{F1}.

\subsection{Remarks}
\begin{enumerate}
\item[(i)] In \verb"\runningheads" use `\emph{et~al.}' if there
are three or more authors.

\item[(ii)] Note the use of \verb"\affil" and \verb"\affilnum" to
link names and addresses. The details for the corresponding
author should be included in the \verb"\corraddr" command.

\item[(iii)] For submitting a double-spaced manuscript, add
\verb"doublespace" as an option to the documentclass line.

\item[(iv)] Use \verb"\cgs" for giving details of financial
sponsors; alternatively use \verb"\cgsn" if the grant number is
also to be included.
%These details will be printed as a footnote,
%with `Contract/grant sponsor:' and `contract/grant number:'
%inserted in the appropriate places.

\item[(v)] The abstract should be capable of standing by itself,
in the absence of the body of the article and of the bibliography.
Therefore, it must not contain any reference citations.

\item[(vi)] Keywords are separated by semicolons.
\end{enumerate}

\begin{figure}
\setlength{\fboxsep}{0pt}%
\setlength{\fboxrule}{0pt}%
\begin{center}
\begin{boxedverbatim}
\begin{table}
\caption{<Table caption>}
\centering
\small
\begin{tabular}{<table alignment>}
\toprule
<column headings>\\
\midrule
<table entries
(separated by & as usual)>\\
<table entries>\\
.
.
.\\
\bottomrule
\end{tabular}
\end{table}
\end{boxedverbatim}
\end{center}
\caption{Example table layout.\label{F2}}
\end{figure}

\section{The Body of the Article}

\subsection{Mathematics} \textsf{\journalclass} makes the full
functionality of \AmS\/\TeX\ available. We encourage the use of
the \verb"align", \verb"gather" and \verb"multline" environments
for displayed mathematics. \textsf{amsthm} is used for setting
theorem-like and proof environments. The usual \verb"\newtheorem"
command needs to be used to set up the environments for your
particular document.

\subsection{Figures and Tables} \textsf{\journalclass} includes the
\textsf{graphicx} package for handling figures.

Figures are called in as follows:
\begin{verbatim}
\begin{figure}
\centering
\includegraphics{<figure name>}
\caption{<Figure caption>}
\end{figure}
\end{verbatim}

For further details on how to size figures, etc., with the
\textsf{graphicx} package see, for example, \cite{R1}
or \cite{R3}. If figures are available in an
acceptable format (for example, .eps, .ps) they will be used but a
printed version should always be provided. \medbreak

The standard coding for a table is shown in Figure~\ref{F2}.

\subsection{Cross-referencing}
The use of the \LaTeX\ cross-reference system
for figures, tables, equations, etc., is encouraged
(using \verb"\ref{<name>}" and \verb"\label{<name>}").

\subsection{Acknowledgements} An Acknowledgements section is started with \verb"\ack" or
\verb"\acks" for \textbf{ACKNOWLEDGEMENT} or
\textbf{ACKNOWLEDGEMENTS}, respectively. It must be placed just
before the References.

\subsection{Bibliography}
The normal commands for producing the reference list are:
\begin{verbatim}
\begin{thebibliography}{99}
\bibitem{<x-ref label>}
         <Reference details>
.
.
.
\end{thebibliography}
\end{verbatim}
where \verb"\bibitem{x-ref label}"
corresponds to \verb"\cite{x-ref label}" in the body of the article
and \verb"{99}" is the widest such number expected and determines
the width of the number column in the reference list.

Please note that the file \textsf{wileyj.bst} is available from
the same download page for those authors using \BibTeX.

\subsection{Double Spacing}
If you need to double space your document for submission please
use the \verb+doublespace+ option as shown in the sample layout in
Figure~\ref{F1}.

\section{Support for \textsf{\journalclass}}
We offer on-line support to participating authors. Please contact
us via e-mail at\\
\href{mailto:weauth-cls@wiley.co.uk}{\texttt{weauth-cls@wiley.co.uk}}.

We would welcome any feedback, positive or otherwise, on your
experiences of using \textsf{\journalclass}.

\section{Copyright Statement}
Please  be  aware that the use of  this \LaTeXe\ class file is
governed by the following conditions.

\subsection{Copyright}
Copyright \copyright\ \volumeyear\ John Wiley \& Sons, Ltd, The
Atrium, Southern Gate, Chichester, West Sussex, PO19~8SQ, UK. All
rights reserved.

\subsection{Rules of Use}
This class file is made available for use by authors who wish to
prepare an article for publication in \emph{\journalnamelc}
published by John Wiley \& Sons, Ltd. The user may not exploit any
part of the class file commercially.

This class file is provided on an \emph{as is}  basis, without
warranties of any kind, either express or implied, including but
not limited to warranties of title, or implied  warranties of
merchantablility or fitness for a particular purpose. There will
be no duty on the author[s] of the software or  John Wiley \&
Sons, Ltd to correct any errors or defects in the software. Any
statutory  rights you may have remain unaffected by your
acceptance of these rules of use.

\ack This class file was developed by Sunrise Setting Ltd,
Torquay, Devon, UK. Website:\\
\href{http://www.sunrise-setting.co.uk}{\texttt{www.sunrise-setting.co.uk}}

\begin{thebibliography}{9}

\bibitem{R1} Kopka~H, Daly~PW. 2003. \emph{A Guide to \LaTeX} (4th~edn).
Addison-Wesley.

\bibitem{R2} Lamport~L. 1994. \emph{\LaTeX: a Document Preparation System} (2nd~edn).
Addison-Wesley.

\bibitem{R3} Mittelbach~F, Goossens~M. 2004. \emph{The \LaTeX\ Companion}
(2nd~edn). Addison-Wesley.
\end{thebibliography}
\end{document}

### 2016-04-11 scripts/getbib.py
Add scripts and Makefile targets

@@ -1,0 +1,10 @@
+#!/usr/bin/env python
+"""This script copies my personal BibTeX file to the current directory."""
+
+import os
+import shutil
+
+bibpath = os.path.join(os.path.expanduser("~"), "Google Drive",
+                       "Library", "Library.bib")
+
+shutil.copy2(bibpath, "library.bib")

### 2016-04-11 scripts/getfigs.py

@@ -1,0 +1,87 @@
+#!/usr/bin/env python
+"""This script copies figures into the local figures directory."""
+
+import os
+import shutil
+from os.path import join
+
+homedir = os.path.expanduser("~")
+expdir = join(homedir, "Research", "Experiments")
+foamrun_23x = join(homedir, "OpenFOAM", "pete-2.3.x", "run")
+foamrun_24x = foamrun_23x.replace("2.3.x", "2.4.x")
+foamrun_ext1 = join("/media", "pete", "Data1", "OpenFOAM", "pete-2.3.x", "run")
+foamrun_ext2 = join("/media", "Data2", "OpenFOAM", "pete-2.3.x", "run")
+gdrive = join(homedir, "Google Drive")
+foamrun_30x = foamrun_23x.replace("2.3.x", "3.0.x")
+foamrun_dev_ext2 = foamrun_ext2.replace("2.3.x", "dev")
+
+
+figdirs = {"RVAT-baseline": join(expdir, "RVAT baseline", "Figures"),
+           "RVAT-Re-dep": join(expdir, "RVAT Re dep", "Figures"),
+           "AD": join(foamrun_23x, "actuatorSurface", "figures"),
+           "CFD-pop": join(homedir, "Google Drive", "Research",
+                           "CFD popularity", "figures"),
+           "CFT-vectors": join(gdrive, "Research", "CFT-vectors", "figures"),
+           "NACAXX20-XFOIL": join(gdrive, "Research", "Foils", "Data",
+                                  "NACAXX20_QBlade", "figures"),
+           "BR-CFD": join(gdrive, "Research", "Papers", "CFT wake modeling",
+                          "figures"),
+           "RVAT-ALM": join(foamrun_30x, "UNH-RVAT-turbinesFoam", "figures"),
+           "RM2-ALM": join(foamrun_30x, "RM2-turbinesFoam", "figures"),
+           "RVAT-ALM-LES": join(foamrun_dev_ext2, "UNH-RVAT-turbinesFoam",
+                                "figures"),
+           "end-effects": join(gdrive, "Research", "OpenFOAM", "turbinesFoam",
+                               "End effects", "figures"),
+           "RM2-tow-tank": join(expdir, "RM2 tow tank", "Figures")}
+
+figlists = {"RVAT-baseline": ["perf.pdf", "meancontquiv.pdf", "kcont.pdf",
[…]

### 2016-04-11 scripts/zip.py

@@ -1,0 +1,20 @@
+#!/usr/bin/env python
+"""This script zips up all the necessary source files to compile the paper."""
+
+from zipfile import ZipFile
+from subprocess import check_output
+import os
+
+# Describe version via git
+version = check_output(["git", "describe"]).strip().decode()
+
+figures = ["figures/" + f for f in os.listdir("figures") \
+           if f.endswith(".pdf") or f.endswith(".png")]
+
+files = ["paper.tex", "paper.bbl"]
+
+with ZipFile("archive/paper-{}.zip".format(version), "w") as f:
+    for fig in figures:
+        f.write(fig)
+    for file in files:
+        f.write(file)

### 2016-04-11 scripts/getbib.py
Add submodule for bib file

@@ -1,10 +1,0 @@
-#!/usr/bin/env python
-"""This script copies my personal BibTeX file to the current directory."""
-
-import os
-import shutil
-
-bibpath = os.path.join(os.path.expanduser("~"), "Google Drive",
-                       "Library", "Library.bib")
-
-shutil.copy2(bibpath, "library.bib")

### 2016-04-11 paper.tex
Start cleaning out template stuff

@@
 
-\def\volumeyear{2010}
+\def\volumeyear{2016}
 
@@
 
-\title{A demonstration of the \LaTeXe\ class file for
-\itshape{\journalnamelc}\footnotemark[2]}
+\title{Actuator line modeling of vertical-axis turbines}
 
-\author{A.~N.~Other}
+\author{P.~Bachant and M.~Wosnik}
 
-\address{John Wiley \& Sons, Ltd, The Atrium, Southern Gate, Chichester,
-West Sussex, PO19~8SQ, UK}
+\address{Center for Ocean Renewable Energy, University of New Hampshire, 24
+Colovos Rd., Durham, NH, 03824, USA}
 
-\corraddr{Journals Production Department, John Wiley \& Sons, Ltd,
-The Atrium, Southern Gate, Chichester, West Sussex, PO19~8SQ, UK.}
+\corraddr{M.~Wosnik, Chase Ocean Engineering Laboratory, University of New
+Hampshire, 24 Colovos Rd., Durham, NH, 03824, USA. \\
+E-mail: martin.wosnik@unh.edu}
 
 \begin{abstract}
-This paper describes the use of the \LaTeXe\
-\textsf{\journalclass} class file for setting papers for
-\emph{\journalnamelc}.
+    % TODO add abstract here
 \end{abstract}
 
-\keywords{class file; \LaTeXe; \emph{\journalabb}}
+\keywords{wind turbine; VAWT; cross-flow turbine; blade element theory; dynamic
+stall; lifting line}
 
@@
 
-\footnotetext[2]{Please ensure that you use the most up to date
-class file,
-available from the WE Home Page at\\
-\href{http://www3.interscience.wiley.com/journal/6276/home}{\texttt{http://www3.interscience.wiley.com/journal/6276/home}}}
+\section{Introduction}
 
-\section{Introduction}
 Many authors submitting to research journals use \LaTeXe\ to

### 2016-04-11 paper.tex
Add Goude as 2nd author
Clean out more template stuff

@@
 
-\runningheads{A.~N.~Other}{A demonstration of the \journalabb\
-class file}
+\runningheads{P.~Bachant \emph{et al.}}{Actuator line modeling of vertical-axis
+turbines}
 
@@
 
-\author{P.~Bachant and M.~Wosnik}
+\author{P.~Bachant, A.~Goude, and M.~Wosnik}
 
@@
 
-Many authors submitting to research journals use \LaTeXe\ to
-prepare their papers. This paper describes the
-\textsf{\journalclass} class file which can be used to convert
-articles produced with other \LaTeXe\ class files into the correct
-form for publication in \emph{\journalnamelc}.
-
-The \textsf{\journalclass} class file preserves much of the
-standard \LaTeXe\ interface so that any document which was
-produced using the standard \LaTeXe\ \textsf{article} style can
-easily be converted to work with the \textsf{\journalclassshort}
-style. However, the width of text and typesize will vary from that
-of \textsf{article.cls}; therefore, \emph{line breaks will change}
-and it is likely that displayed mathematics and tabular material
-will need re-setting.
-
-In the following sections we describe how to lay out your code to
-use \textsf{\journalclass} to reproduce the typographical look of
-\emph{\journalnamelc}. However, this paper is not a guide to
-using \LaTeXe\ and we would refer you to any of the many books
-available (see, for example, \cite{R1,R2,R3}).
-
 \section{The Three Golden Rules}
@@
 
-\section{Getting Started} The \textsf{\journalclassshort} class file should run
-on any standard \LaTeXe\ installation. If any of the fonts, style
-files or packages it requires are missing from your installation,
-they can be found on the \emph{\TeX\ Collection} DVDs or from
-CTAN.
-
-\emph{\journalnamelc} is published using Times fonts and this is
-achieved by using the \verb"times"
-option as\\
-\verb"\documentclass[times]{weauth}".
-
-\noindent If for any reason you have a problem using Times you can
-easily resort to Computer Modern fonts by removing the
-\verb"times" option.
-
-\begin{figure}
-\setlength{\fboxsep}{0pt}%
-\setlength{\fboxrule}{0pt}%
-\begin{center}
-\begin{boxedverbatim}
-\documentclass[times]{weauth}
-%\documentclass[times,doublespace]{weauth}%For paper submission
-
-\begin{document}
-
-\runningheads{<Initials and Surnames>}{<Short title>}
-
-\articletype{<Article type>}
-
-\title{<Initial cap, lower case>}
-
-\author{<An Author\affil{1},
-Someone Else\affil{2}\ and Perhaps Another\affil{1}>}
-
-\address{<\affilnum{1}First author's address
-(in this example it is the same as the third author)
-\affilnum{2}Second author's address>}
-
-\corraddr{<Corresponding author's address>.
-E-mail: <corresponding author's email address>}
-
-%\cgs{<Contract/grant sponsor name (no number)>}
-%\cgsn{<Contract/grant sponsor name>}{<number>}
-
-\begin{abstract}
-<Text>
-\end{abstract}
-
-\keywords{<List keywords>}
-
-\maketitle
-
-\section{Introduction}
-.
-.
-.
-\end{boxedverbatim}
-\end{center}
-\caption{Example header text.\label{F1}}
-\end{figure}
-
-\section{The Article Header Information}
-The heading for any file using \textsf{\journalclass} is shown in
-Figure~\ref{F1}.
-
-\subsection{Remarks}
-\begin{enumerate}
-\item[(i)] In \verb"\runningheads" use `\emph{et~al.}' if there
-are three or more authors.
-
-\item[(ii)] Note the use of \verb"\affil" and \verb"\affilnum" to
-link names and addresses. The details for the corresponding
-author should be included in the \verb"\corraddr" command.
-
-\item[(iii)] For submitting a double-spaced manuscript, add
-\verb"doublespace" as an option to the documentclass line.
-
-\item[(iv)] Use \verb"\cgs" for giving details of financial
-sponsors; alternatively use \verb"\cgsn" if the grant number is
-also to be included.
-%These details will be printed as a footnote,
-%with `Contract/grant sponsor:' and `contract/grant number:'
-%inserted in the appropriate places.
-
-\item[(v)] The abstract should be capable of standing by itself,
-in the absence of the body of the article and of the bibliography.
-Therefore, it must not contain any reference citations.
-
-\item[(vi)] Keywords are separated by semicolons.
-\end{enumerate}
-
-\begin{figure}
-\setlength{\fboxsep}{0pt}%
-\setlength{\fboxrule}{0pt}%
-\begin{center}
-\begin{boxedverbatim}
 \begin{table}
-\caption{<Table caption>}
+    \caption{This is a sample table}
     \centering
     \small
-\begin{tabular}{<table alignment>}
+    \begin{tabular}{c|c|c}
     \toprule
-<column headings>\\
+    This & that & other\\
     \midrule
-<table entries
-(separated by & as usual)>\\
-<table entries>\\
-.
-.
-.\\
+    yo & yo & yo\\
     \bottomrule
@@
 \end{table}
-\end{boxedverbatim}
-\end{center}
-\caption{Example table layout.\label{F2}}
-\end{figure}
 
-\section{The Body of the Article}
+\acks These are the acknowledgments
 
-\subsection{Mathematics} \textsf{\journalclass} makes the full
-functionality of \AmS\/\TeX\ available. We encourage the use of
-the \verb"align", \verb"gather" and \verb"multline" environments
-for displayed mathematics. \textsf{amsthm} is used for setting
-theorem-like and proof environments. The usual \verb"\newtheorem"
-command needs to be used to set up the environments for your
-particular document.
-
-\subsection{Figures and Tables} \textsf{\journalclass} includes the
-\textsf{graphicx} package for handling figures.
-
-Figures are called in as follows:
-\begin{verbatim}
-\begin{figure}
-\centering
-\includegraphics{<figure name>}
-\caption{<Figure caption>}
-\end{figure}
-\end{verbatim}
-
-For further details on how to size figures, etc., with the
-\textsf{graphicx} package see, for example, \cite{R1}
-or \cite{R3}. If figures are available in an
-acceptable format (for example, .eps, .ps) they will be used but a
-printed version should always be provided. \medbreak
-
-The standard coding for a table is shown in Figure~\ref{F2}.
-
-\subsection{Cross-referencing}
-The use of the \LaTeX\ cross-reference system
-for figures, tables, equations, etc., is encouraged
-(using \verb"\ref{<name>}" and \verb"\label{<name>}").
-
-\subsection{Acknowledgements} An Acknowledgements section is started with \verb"\ack" or
-\verb"\acks" for \textbf{ACKNOWLEDGEMENT} or
-\textbf{ACKNOWLEDGEMENTS}, respectively. It must be placed just
-before the References.
-
-\subsection{Bibliography}
-The normal commands for producing the reference list are:
-\begin{verbatim}
-\begin{thebibliography}{99}
-\bibitem{<x-ref label>}
-         <Reference details>
-.
-.
-.
-\end{thebibliography}
-\end{verbatim}
-where \verb"\bibitem{x-ref label}"
-corresponds to \verb"\cite{x-ref label}" in the body of the article
-and \verb"{99}" is the widest such number expected and determines
-the width of the number column in the reference list.
-
-Please note that the file \textsf{wileyj.bst} is available from
-the same download page for those authors using \BibTeX.
-
-\subsection{Double Spacing}
-If you need to double space your document for submission please
-use the \verb+doublespace+ option as shown in the sample layout in
-Figure~\ref{F1}.
-
-\section{Support for \textsf{\journalclass}}
-We offer on-line support to participating authors. Please contact
-us via e-mail at\\
-\href{mailto:weauth-cls@wiley.co.uk}{\texttt{weauth-cls@wiley.co.uk}}.
-
-We would welcome any feedback, positive or otherwise, on your
-experiences of using \textsf{\journalclass}.
-
-\section{Copyright Statement}
-Please  be  aware that the use of  this \LaTeXe\ class file is
-governed by the following conditions.
-
-\subsection{Copyright}
-Copyright \copyright\ \volumeyear\ John Wiley \& Sons, Ltd, The
-Atrium, Southern Gate, Chichester, West Sussex, PO19~8SQ, UK. All
-rights reserved.
-
-\subsection{Rules of Use}
-This class file is made available for use by authors who wish to
-prepare an article for publication in \emph{\journalnamelc}
-published by John Wiley \& Sons, Ltd. The user may not exploit any
-part of the class file commercially.
-
-This class file is provided on an \emph{as is}  basis, without
-warranties of any kind, either express or implied, including but
-not limited to warranties of title, or implied  warranties of
-merchantablility or fitness for a particular purpose. There will
-be no duty on the author[s] of the software or  John Wiley \&
-Sons, Ltd to correct any errors or defects in the software. Any
-statutory  rights you may have remain unaffected by your
-acceptance of these rules of use.
-
-\ack This class file was developed by Sunrise Setting Ltd,
-Torquay, Devon, UK. Website:\\
-\href{http://www.sunrise-setting.co.uk}{\texttt{www.sunrise-setting.co.uk}}
-
-\begin{thebibliography}{9}
-
-\bibitem{R1} Kopka~H, Daly~PW. 2003. \emph{A Guide to \LaTeX} (4th~edn).
-Addison-Wesley.
-
-\bibitem{R2} Lamport~L. 1994. \emph{\LaTeX: a Document Preparation System} (2nd~edn).
-Addison-Wesley.
-
-\bibitem{R3} Mittelbach~F, Goossens~M. 2004. \emph{The \LaTeX\ Companion}
-(2nd~edn). Addison-Wesley.
-\end{thebibliography}
+\bibliographystyle{wileyj.bst}
+\bibliography{library}
 \end{document}

### 2016-04-28 scripts/getfigs.py
Remove figs from getfigs script

@@ -34,32 +34,8 @@ figdirs = {"RVAT-baseline": join(expdir, "RVAT baseline", "Figures"),
                                "End effects", "figures"),
            "RM2-tow-tank": join(expdir, "RM2 tow tank", "Figures")}
 
-figlists = {"RVAT-baseline": ["perf.pdf", "meancontquiv.pdf", "kcont.pdf",
-                              "meanu_2tsrs.pdf", "meanv_2tsrs.pdf",
-                              "meanw_2tsrs.pdf", "xvorticitycont.pdf",
-                              "meanupvpcont.pdf", "meanupwpcont.pdf",
-                              "mean_vel_vs_tsr.pdf", "multispec.pdf",
-                              "mombargraph.pdf", "Kbargraph.pdf",
-                              "Kturbtranscont.pdf", "fpeak_vcont.pdf",
-                              "fstrength_vcont.pdf", "k_2tsrs.pdf"],
-            "RVAT-Re-dep": ["cp_curves.pdf", "cd_curves.pdf",
-                            "perf_re_dep.pdf", "meancontquiv_10.pdf",
-                            "k_contours_10.pdf", "mean_u_k_profiles.pdf",
-                            "wake_spectra.pdf", "mom_bar_graph.pdf",
-                            "K_trans_bar_graph.pdf", "wake_trans_totals.pdf"],
-            "AD": ["streamwise.pdf", "meancontquiv.pdf"],
+figlists = {"RVAT-baseline": ["meancontquiv.pdf", "kcont.pdf"],
             "CFT-vectors": ["cft-vectors.pdf", "alpha_deg_urel_geom.pdf"],
-            "NACAXX20-XFOIL": ["all_foils_re_dep.pdf",
-                               "foil_kinematics_ct.pdf",
-                               "cft_re_dep_foils.pdf"],
-            "BR-CFD": ["CAD.pdf", "2D_blade_mesh_closeup.png", "2D_mesh.png",
-                       "2D_vorticity_SA_964.png",
-                       "3D_vorticity_SA_964_10-threshold.png",
-                       "kcont_kOmegaSST.pdf", "kcont_SpalartAllmaras.pdf",
-                       "meancontquiv_kOmegaSST.pdf",
-                       "meancontquiv_SpalartAllmaras.pdf", "mom_bar_graph.pdf",
-                       "perf_bar_chart.pdf", "profiles.pdf",
-                       "verification.pdf"],
[…]

### 2016-04-28 scripts/getfigs.py
Fix typo in getfigs.py

@@ -49,7 +49,7 @@ figlists = {"RVAT-baseline": ["meancontquiv.pdf", "kcont.pdf"],
             "RM2-tow-tank": ["meancontquiv.pdf", "k_contours.pdf"]}
 
 
-if not os.path.isdir("figure"):
+if not os.path.isdir("figures"):
     os.mkdir("figures")
 
 

### 2016-04-28 paper.tex
Remove more template content

@@
 
-\section{The Three Golden Rules}
-
-Before we proceed, we would like to stress \emph{three golden
-rules} that need to be followed to enable the most efficient use
-of your code at the typesetting stage:
-\begin{enumerate}
-\item[(i)] keep your own macros to an absolute minimum;
-
-\item[(ii)] as \TeX\ is designed to make sensible spacing
-decisions by itself, do \emph{not} use explicit horizontal or
-vertical spacing commands, except in a few accepted (mostly
-mathematical) situations, such as \verb"\," before a
-differential~d, or \verb"\quad" to separate an equation from its
-qualifier;
-
-\item[(iii)] follow the \emph{\journalnamelc} reference style.
-\end{enumerate}
-
-\begin{table}
-    \caption{This is a sample table}
-    \centering
-    \small
-    \begin{tabular}{c|c|c}
-    \toprule
-    This & that & other\\
-    \midrule
-    yo & yo & yo\\
-    \bottomrule
-    \end{tabular}
-\end{table}
-
 \acks These are the acknowledgments
 
-\bibliographystyle{wileyj.bst}
+\bibliographystyle{wileyj}
 \bibliography{library}

### 2016-04-28 paper.tex
Add content from thesis and modify just enough to build

@@
 \usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+
+%% Code syntax highlighting
+
+\usepackage{listings}
+\usepackage{color}
+
+\definecolor{mygreen}{rgb}{0,0.6,0}
+\definecolor{mygray}{rgb}{0.5,0.5,0.5}
+\definecolor{mymauve}{rgb}{0.58,0,0.82}
+\definecolor{mylightgray}{rgb}{0.95,0.95,0.95}
+
+\lstset{frame=single,
+    % language=C++,
+    aboveskip=3mm,
+    belowskip=3mm,
+    showstringspaces=false,
+    columns=flexible,
+    basicstyle=\ttfamily,
+    numbers=none,
+    numberstyle=\tiny\color{mygray},
+    keywordstyle=\bfseries\color{blue!40!black},
+    commentstyle=\color{gray},
+    stringstyle=\color{mygreen},
+    % otherkeywords={for,if,else},
+    breaklines=true,
+    breakatwhitespace=true,
+    tabsize=4,
+    escapeinside={\%*}{*)},
+    keepspaces=true,
+    captionpos=b,
+    backgroundcolor=\color{mylightgray}
+}
+
+\graphicspath{{figures/}}
 
@@
 
+Despite its apparent---but not completely certain---effectiveness for predicting
+the performance and wake of a single turbine, blade-resolved CFD presents a huge
+computational expense since it must resolve fine details of the blade boundary
+layers, and must be performed in three dimensions, which will preclude its use
+for array analysis until the availability of computing power increases
+sufficiently. It is therefore necessary to explore simpler models that can
+predict the turbine loading and flow field with acceptable fidelity, but that
+are economical enough to not require high performance computing, at least for
+individual devices.
+
+Very fast solution times can be obtained with blade element momentum (BEM)
+models, such as the double multiple streamtube (DMST) approached developed by
+Paraschivoiu~\cite{Para1988}. However, their poor performance for high solidity
+turbines \cite{Joo2015}, makes them less attractive, particular for marine
+hydrokinetic applications. Their description of the flow field is very crude as
+well, simply computing the momemtum contained within discretized streamtubes,
+which cannot resolve any of the complicated flow structures shed by the rotor.
+
+Blade element based vortex line methods can be solved with reasonable
+computational effort---ranging from an expense close to BEM up to that of 2-D
+blade-resolved CFD. The cost is a function of the number of vortex elements,
+which for free vortex methods increases at each time step, slowing down the
+calculation as it marches forward in time. Vortex methods can resolve more flow
+details than BEM. However, since the flow is governed by the Laplace equation,
+nonlinear effects such as the turbulent transport will not be included, which we
+have determined in experiments to be the same order of magnitude as the mean
+vertical advection.
+
+It is therefore desirable to retain a Navier--Stokes description of the flow
+field and use an actuator-type model for parameterizing the turbine loading,
+which dramatically drives down computational expense by removing the need for
+complicated meshes with many fine cells to resolve the boundary layers near the
+blade surfaces. As shown in Chapter~\ref{chap:RVAT-baseline}, the conventional
+uniform actuator disk is not a good candidate for a cross-flow turbine wake
+generator, never mind the fact that it does not typically compute performance
+predictions. There are then two actuator methods that use blade element theory
+to compute blade loading and therefore generate a wake that more closely
+resembles that of an actual turbine. These are the actuator cylinder or
+swept-surface model (ASSM) and the actuator line model (ALM). The ASSM solves
+for the average blade loading along its path and applies this as a constant body
+force term in the momentum equation. The ALM takes a similar approach but is an
+unsteady method, resolving the blade element locations in time.
+
+The ALM, originally developed by Sorensen and Shen \cite{Sorensen2002}, has
+become popular for modeling axial-flow or horizontal-axis turbines, and has been
+shown in blind tests to be competitive with blade-resolved
+CFD~\cite{Krogstad2013, Pierella2014}. The ALM combined with large eddy
+simulation has become the state-of-the-art for modeling entire wind
+farms~\cite{Archer2013, Churchfield2012, Sorensen2015, Fleming2013,
+    Fleming2014}. Like other blade element techniques, the effectiveness of the ALM
+for AFTs is in part due to the quasi-steady nature of the flow in the blade
+reference frame, and the relatively rare occurrence of stall.
+
+The ASSM and ALM were implemented to model a very low Reynolds number 2-D
+cross-flow turbine experiment in a flume using large eddy simulation
+(LES)~\cite{Shamsoddin2014}. Performance predictions for this case were not
+reported, but the ALM was shown to be more effective at postdicting the wake
+characteristics measured in the experiments by Brochier et
+al.~\cite{Brochier1986}.
+
+It is therefore proposed that an actuator line model may be the optimal
+combination of high-fidelity flow modeling that includes performance
+predictions, but with reduced computational expense. Here we will develop and
+evaluate the effectiveness of an ALM for cross-flow turbines inside both RANS
+and LES simulations, implemented as an extension for the OpenFOAM open-source
+CFD library.
+
+\section{Theory}
+
+The actuator line model is based on the classical blade element theory combined
+with a Navier--Stokes description of the flow field. The ALM treats turbine
+blades as lines of blade elements, for which 2-D profile lift and drag
+coefficients are given. For each blade element, relative flow velocity and angle
+of attack are computed by adding the vectors of relative blade motion and the
+local fluid velocity. The blade lift and drag forces are calculated as
+\begin{equation}
+    F_l = \frac{1}{2} \rho A_\mathrm{elem} C_l |\vec{U}_\mathrm{rel}|^2,
+\end{equation}
+and
+\begin{equation}
+    F_d = \frac{1}{2} \rho A_\mathrm{elem} C_d |\vec{U}_\mathrm{rel}|^2,
+\end{equation}
+where $\rho$ is the fluid density, $A_\mathrm{elem}$ is the blade element
+planform area (span $\times$ chord), $\vec{U}_\mathrm{rel}$ is the local
+relative velocity, and $C_l$ and $C_d$ are the sectional lift and drag
+coefficients, linearly interpolated from a table per the local angle of attack.
+The forces are then projected onto the rotor coordinate system to calculate
+torque, overall drag, etc. Forces from the turbine shaft and blade support
+struts are computed in a similar way. After the force on the actuator lines from
+the flow is computed, it is then added to the Navier--Stokes equations as a body
+force or momentum source (per unit density, assuming incompressible flow):
+\begin{equation}
+    \frac{\mathrm{D} \vec{u}}{\mathrm{D} t} = - \frac{1}{\rho} \nabla p + \nu
+    \nabla^2 \vec{u} + F_\mathrm{turbine}.
+\end{equation}
+
+\section{Blade element discretization}
+
+In the ALM, a turbine is a collection of actuator lines, which themselves are
+collections of actuator line elements (ALEs). The position of each ALE is a
+point in space indicating its quarter-chord location. The element is further
+defined by its chord direction vector, chord length, span direction vector, span
+length, and velocity vector.
+
+An actuator line is created from defined geometry points, between which ALE
+parameters are interpolated linearly. This way, an actuator line can be defined
+by fewer geometry points than element locations. For example, an AL with
+straight planform boundaries---e.g. a straight or tapered wing---only needs two
+geometry points to be fully defined. Figure~\ref{fig:AL-geom} shows and example
+schematic of a tapered actuator line with three geometry points at the
+half-chord locations and six total elements. Note that the middle geometry point
+is technically redundant, but is shown for illustrative purposes. For actuator
+lines that represent bluff bodies, e.g., shafts, the chord mount offset is set
+to 1/4, such that the element location is centered along the line of geometry
+points.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.9\textwidth]{thesis_alm-geometry}
+
+    \caption{Actuator line geometry. Filled circles indicate geometry points
+        whereas squares indicate actuator line element locations.}
+
+    \label{fig:AL-geom}
+\end{figure}
+
+\section{Determining inflow velocity}
+
+In momentum methods, the inflow velocity is determined by solving for the axial
+and angular induction factors \cite{Manwell2002}. However, using Navier--Stokes
+methods, it is somewhat unclear how to calculate the velocity vector used to
+compute the angle of attack and relative velocity, though we have access to much
+more information about the flow. Sorensen and Shen used an actuator line
+element's position to determine the inflow velocity for an axial-flow turbine
+\cite{Sorensen2002}. Similarly, Shamsoddin and Porte-Agel use the velocity at a
+blade element's location in their actuator line simulation of a vertical-axis
+turbine using LES \cite{Shamsoddin2014}. NREL's SOWFA ALM for OpenFOAM also uses
+the velocity at the actuator line element location for computing inflow velocity
+and local angle of attack with no corrections \cite{Churchfield2013}.
+
+Schito and Zasso developed an effective velocity model (EVM) for computing
+actuator forces in Navier--Stokes simulations \cite{Schito2014}. Their EVM
+proposes that inflow velocity should be sampled along a line perpendicular to
+the mean relative flow direction. They ultimately determined that the sampling
+line should be 1.5 chord lengths upstream of the actuator point (i.e.
+quarter-chord location). A sampling line was chosen to be 5 times the local mesh
+cell length. Finally, an angle of attack correction is proposed
+\begin{equation}
+    \Delta \alpha = \frac{c}{M} (1.2553 - 0.0552 C_d) C_l,
+    \label{eq:EVM-dalpha}
+\end{equation}
+where $c$ is the chord length, $M$ is the mesh size, and $C_l$ and $C_d$ are the
+lift and drag coefficients, respectively. Note that the constants in
+Equation~\ref{eq:EVM-dalpha} were determined from a calibration with 2-D
+blade-resolved CFD for a NACA 0012 foil and are not assumed to be universal.
+
+The EVM, despite showing robustness for its chosen validation case,
+unfortunately involves determination of two unknown tuning parameters. To avoid
+the additional effort and uncertainty in determining these, the inflow velocity
+was sampled at the element quarter-chord location using OpenFOAM's
+\texttt{interpolationCellPoint} class, which provides a linear weighted
+interpolation using cell values. This algorithm helps keep the sampled velocity
+``smooth'' compared with using the cell values themselves, especially when
+elements are moving in space as they are in a turbine, since meshes will likely
+have a cell size on the same order as the chord length, and will move on the
+order of one cell length per time step.
+
+\section{Static foil coefficient data}
+
+Static input foil coefficient data were taken from Sheldahl and
+Klimas~\cite{Sheldahl1981}---a popular database developed for CFTs, which
+contains values over a wide range of Reynolds numbers. The Sheldahl and Klimas
+dataset has some limitations, namely that data for some foil and/or Reynolds
+numbers were ``synthesized'' numerically from other measurements. Despite its
+flaws, this dataset is the likely most comprehensive available with respect to
+variety of profiles and ranges of Reynolds numbers. Surprisingly, considering
+the maturity and popularity of NACA foils, data remains scarce, especially for
+$Re_c \sim 10^5$.
+
+NACA 0021 coefficients were used for both turbines, despite the fact that the
+UNH-RVAT is constructed from NACA 0020 foils, as a NACA 0020 dataset is not
+available---it is assumed the small difference in foil thickness is negligible.
+Since pitching moment data were only available at limited Reynolds numbers, two
+datasets were used: The lowest for $Re_c \leq 3.6 \times 10^5$ and highest $Re_c
+\geq 6.8 \times 10^5$. For each actuator line element, blade chord Reynolds
+number is computed based on the sampled inflow velocity, and the static
+coefficients are then interpolated linearly within the database.
+
+\section{Force projection}
+
+After the force on the ALE from the flow is calculated, it is then projected
+back onto the flow field as a source term in the momentum equation. To avoid
+instability due to steep gradients, the source term is tapered from its maximum
+value away from the element location by means of a spherical Gaussian function.
+The width of this function $\eta$ is controlled by a single parameter
+$\epsilon$, which is then multiplied by the actuator line element force and
+imparted on a cell with distance $| \vec{r} |$ from the actuator line element
+quarter chord location:
+\begin{equation}
+    \eta = \frac{1}{\epsilon^3 \pi^{3/2}} \exp
+    \left[ - \left( \frac{| \vec{r} |}{\epsilon} \right)^2 \right].
+    \label{eq:projection}
+\end{equation}
+
+Troldborg~\cite{Troldborg2008} proposed that the Gaussian width should be set to
+twice the local cell length $\Delta x$ in order to maintain numerical stability.
+
+Jha et al.~\cite{Jha2014} provides guidelines for choosing a projection
+width.
+
+Schito and Zasso \cite{Schito2014} found that a projection $\epsilon$ equal to
+the local mesh length was optimal.
+
+Martinez-Tossas and Meneveau \cite{Martinez-Tossas2015b} used a 2-D potential
+flow analysis to determine that the optimal projection width for a lifting
+surface is 14--25\% of the chord length. The width due to the wake caused by the
+foil drag force was recommended to be on the order of the momentum thickness
+$\theta$, which for a bluff body or foil at large angle of attack is related to
+the drag coefficient ($O(1)$) by \cite{TennekesAndLumley}
+\begin{equation}
+    C_d = 2 \theta / l,
+    \label{eq:mom-thickness}
+\end{equation}
+where $l$ is a reference length, e.g., diameter for a cylinder or chord length
+for a foil.
+
+Using these guidelines, three Gaussian width values were determined: one
+relative to the chord length, one to the mesh size, and one to the momentum
+thickness due to drag force. Each three were computed for all elements at each
+time step, and the largest was chosen for the force projection algorithm. Using
+this adaptive strategy, fine meshes could benefit from the increased accuracy of
+more concentrated momentum sources, and coarse meshes would be protected from
+numerical instability.
+
+The Gaussian width due to mesh size $\epsilon_{\mathrm{mesh}}$ was determined
+locally on an element-wise basis by estimating the size of the cell containing
+the element as
+\begin{equation}
+    \Delta x \approx \sqrt[3]{V_\mathrm{cell}},
+\end{equation}
+where $V_\mathrm{cell}$ is the cell volume. To account for the possibility of
+non-unity aspect ratio cells, an additional factor $C_\mathrm{mesh}$ was
+introduced, giving
+\begin{equation}
+    \epsilon_{\mathrm{mesh}} = 2C_\mathrm{mesh} \Delta x.
+\end{equation}
+$C_{\mathrm{mesh}}$ was set to 2.0 for the simulations presented
+here---determined by trial-and-error to provide stability on the finest grids.
+However, in the ALM code $C_{\mathrm{mesh}}$ is selectable at run time for each
+profile used.
+
+\section{Unsteady effects}
+
+In the context of a turbine---especially a cross-flow turbine---the actuator
+lines will encounter unsteady conditions, both in their angle of attack and
+relative velocity. These conditions necessitate the use of unsteady aerodynamic
+models to augment the static foil characteristics, both to capture the time
+resolved response of the attached flow loading and effects of flow acceleration,
+also know as added mass. Furthermore, the angles of attack encountered by a CFT
+blade will often be high enough to encounter dynamic stall (DS). It is therefore
+necessary to model both unsteady attached and detached flow to obtain accurate
+loading predictions.
+
+\subsection{Dynamic stall}
+
+Dynamic stall is encountered when the blade angle of attack changes rapidly in
+time and exceeds a certain threshold, often near the static stall angle
+\cite{McCroskey1981}. The stall is characterized by an initial increase in lift
+beyond static values as a vortex is shed from the foil's leading edge, after
+which a drop in lift and large nose-down pitching moment occurs as the vortex is
+advected downstream. As the angle of attack drops below the critical value, flow
+reattaches, closing the so-called hysteresis loop. Dynamic stall has been shown
+to be a significant positive contributor to performance in CFTs \cite{Para2002,
+    Urbina2013}, therefore an accurate model is key.
+
+DS models were first developed to improve predictive capability for helicopter
+rotors, on which DS has significant effects on maneuverability and operational
+envelope \cite{Bousman2000}. A summary of dynamic stall models developed for
+helicopter rotors is presented in \cite{Leishman2006}. The simplest dynamic
+stall models rely on semi-empirical correlations, e.g., the Gormont model
+\cite{Gormont1973}, developed at the Boeing--Vertol Company. Several variants of
+the Gormont model were developed for vertical-axis wind turbines, with varying
+degrees of success; a summary is presented in \cite{Para2002}.
+
+Leishman and Beddoes (LB) \cite{Leishman1989} developed a semi-empirical model
+for unsteady aerodynamics and dynamic stall, which is derived from the
+phenomenology of the physics instead rather than pure empiricism. Beddoes then
+updated the model to the so-called third generation or ``3G'' version
+\cite{Beddoes1993}. The LB DS models can be summarized conceptually based on the
+following principles:
+\begin{itemize}
+    \item Dynamic conditions cause a time lag in effective angle of attack and
+    lift force.
+
+    \item Separation is determined by the Kirchoff flow approximation, which is
+    also used to parameterize the normal force coefficient table based on the
+    trailing edge separation point. This separation point also encounters a time
+    lag.
+
+    \item The separation initiates a vortex shedding cycle that causes an
+    overshoot and subsequent undershoot in lift before returning to an attached
+    flow condition.
+\end{itemize}
+
+Sheng et al. \cite{Sheng2008} developed an LB DS model variant targeted at low
+Mach numbers. This model, along with the original and 3G LB DS model variants,
+was tested for its effectiveness in cross-flow turbine conditions by Dyachuk et
+al.~\cite{Dyachuk2014}, who concluded that the Sheng et al. variant results
+matched most closely with experiments. In a similar study, the Sheng et al.
+model also faired better than the Gormont model \cite{Dyachuk2015}, which
+inspired its adoption here for the ALM.
+
+Before the dynamic stall subroutine is executed, the static profile data for
+each element is interpolated linearly based on local chord Reynolds number. The
+profile data characteristics---static stall angle, zero-lift drag coefficient,
+and separation point curve fit parameters---are then recomputed each time step
+such that the effects of Reynolds number on the static data are included.
+
+Inside the ALM, angle of attack is sampled from the flow field rather than
+calculated based on the geometric angle of attack. Therefore, the implementation
+of the LB DS model was such that the equivalent angle of attack
+$\alpha_\mathrm{equiv}$ was taken as the sampled rather than the lagged
+geometric value. A similar implementation was used by Dyachuk et al.
+\cite{Dyachuk2015a} inside a vortex model.
+
+\subsection{Added mass}
+
+A correction for added mass effects, or the effects due to accelerating the
+fluid, was taken from Strickland et al.~\cite{Strickland1981}, which was
+derived by considering a pitching flat plate in potential flow. In the blade
+element coordinate system, the normal and chordwise (pointing from trailing to
+leading edge, which is opposite the $x$-direction used by Strickland \emph{et
+    al.}) coefficients due to added mass are
+\begin{equation}
+    C_{n_\mathrm{AM}} = -\frac{\pi c \dot{U_n}}{8 | U_\mathrm{rel} |^2},
+\end{equation}
+and
+\begin{equation}
+    C_{c_\mathrm{AM}} = \frac{\pi c \dot{\alpha} U_n }{8 | U_\mathrm{rel} |^2},
+\end{equation}
+respectively, where $U_n$ is the normal component of the relative velocity, and
+dotted variables indicate time derivatives, which were calculated using a simple
+first order backward finite difference. Similarly, the quarter-chord moment
+coefficient due to added mass was calculated as
+\begin{equation}
+    C_{m_\mathrm{AM}} = -\frac{C_{n_\mathrm{AM}}}{4}
+        - \frac{U_n U_c}{8 | U_\mathrm{rel} |^2},
+\end{equation}
+where $U_c$ is the chordwise component of relative velocity. Note that the
+direction of positive moment is ``nose-up,'' which is opposite that used by
+Strickland et al..
+
+The normal and chordwise added mass coefficients translate to lift and drag
+coefficients by
+\begin{equation}
+    C_{l_\mathrm{AM}} = C_{n_\mathrm{AM}} \cos \alpha + C_{c_\mathrm{AM}} \sin
+    \alpha,
+\end{equation}
+and
+\begin{equation}
+    C_{d_\mathrm{AM}} = C_{n_\mathrm{AM}} \sin \alpha - C_{c_\mathrm{AM}} \cos
+    \alpha,
+\end{equation}
+respectively. The added mass coefficients were then added to those calculated by
+the dynamic stall model.
+
+\section{Flow curvature corrections}
+
+The rotating blades of a cross-flow turbine will have non-constant chordwise
+angle of attack distributions due to their circular paths---producing so-called
+flow curvature effects~\cite{Migliore1980}. This makes it difficult to define a
+single angle of attack for use in the static coefficient lookup tables.
+Furthermore, this effect is more pronounced for high solidity ($c/R$) turbines.
+Two different flow curvature corrections were considered: one by Goude
+\cite{Goude2012} and one by Mandal and Burton \cite{Mandal1994}.
+
+The Goude correction is derived by considering a flat plat moving along a
+circular path in potential flow, for which the effective angle of attack
+including flow curvature effects is given by
+\begin{equation}
+    \alpha = \delta + \arctan \frac{V_\mathrm{abs} \cos(\theta_b -
+        \beta)}{V_\mathrm{abs} \sin(\theta_b - \beta) + \Omega R} - \frac{\Omega
+        x_{0r}c}{V_\mathrm{ref}} - \frac{\Omega c}{4 V_\mathrm{ref}},
+    \label{eq:Goude-curvature}
+\end{equation}
+where $\delta$ is the blade pitch angle, $V_\mathrm{abs}$ is the magnitude of
+the local inflow velocity at the blade, $\theta_b$ is the blade azimuthal
+position, $\beta$ is the direction of the inflow velocity, $\Omega$ is the
+turbine's angular velocity, $R$ is the blade element radius, $x_{0r}$ is a
+normalized blade attachment point along the chord (or fractional chord distance
+of the mounting point from the quarter-chord), $c$ is the blade chord length,
+and $V_\mathrm{ref}$ is the reference flow velocity for calculating angle of
+attack.
+
+In the actuator line model, each element's angle of attack is calculated using
+vector operations, which means the first two terms in
+Equation~\ref{eq:Goude-curvature} are taken care of automatically since each
+element's inflow velocity, chord direction, and element velocity vectors are
+tracked. Therefore, the last two terms in Equation~\ref{eq:Goude-curvature} were
+simply added to the scalar angle of attack value. Note that for a cross-flow
+turbine, this correction effectively offsets the angle of attack, which
+therefore increases its magnitude on the upstream half of the blade path, and
+decreases its magnitude on the downstream half, where the angle of attack is
+negative.
+
+The Mandal--Burton flow curvature correction assumes that since the blade is
+encountering a curvilinear flow, it can be treated as having virtual camber.
+They introduce a factor to describe the variation of angle of attack from the
+leading to trailing edge
+\begin{equation}
+    \Delta \alpha = \alpha_\mathrm{TE} - \alpha_\mathrm{LE},
+    \label{eq:Mandal-Burton-alpha-diff}
+\end{equation}
+where TE and LE subscripts denote the values of angle of attack at the trailing
+and leading edge, respectively. Calculating these values for an actuator line
+element can be done by tracking the leading and trailing edge locations and
+velocities, then performing the same vector arithmetic used to calculate the
+quarter-chord angle of attack.
+
+An incidence correction factor
+\begin{equation}
+    \alpha_c = \arctan \left( \frac{1 - \cos (\Delta \alpha / 2)}{\sin (\Delta
+        \alpha / 2)} \right)
+    \label{eq:Mandal-Burton-alpha-corr}
+\end{equation}
+is introduced and added to the uncorrected angle of attack. Like the Goude
+model, $\alpha_c$ is positive on the upstream half of the turbine rotation and
+negative on the downstream half.
+
+It was determined that the Goude model performed best with respect to matching
+the near-wake characteristics measured in the experiments, while the performance
+predictions from both flow curvature models were relatively close to each other.
+Thus, the Goude model was used for all simulations reported here.
+
+\section{End effects}
+
+Helmholtz's second vortex theorem states that vortex lines may not end in a
+fluid, but must either form closed loops or extend to boundaries. Consequently
+the lift distribution due to the bound vortex from foils of finite span must
+drop to zero at the tips.
+
+Glauert~\cite{Glauert1935} used Prandtl's lifting line theory \cite{Prandtl1927}
+to develop a tip loss correction factor $F$ to be applied to the local lift of
+an axial-flow rotor blade element:
+\begin{equation}
+    F = \frac{2}{\pi} \cos^{-1} \left[ \exp \left( - \frac{N (R-r)}{2R \sin
+        \phi_R} \right) \right],
+\end{equation}
+where $F$ is the correction factor applied to the local lift, $N$ is number of
+blades, $R$ is the rotor radius, $r$ is the local blade element radius, and
+$\phi_R$ is the flow angle at the tip.
+
+The Glauert correction was further refined for horizontal-axis wind turbines by
+Shen et al. \cite{Shen2005a} as
+\begin{equation}
+    F_1 = \frac{2}{\pi} \cos^{-1} \left[ \exp \left( -g \frac{N (R-r)}{2R \sin
+    \phi_R} \right) \right],
+\end{equation}
+where $g$ is an additional function depending on the tip speed ratio and two
+tuning constants:
+\begin{equation}
+    g = \exp [ -c_1 (B \lambda - c_2) ].
+\end{equation}
+
+Despite their success for the blade element analysis of axial-flow rotors, these
+corrections both depend on rotor parameters---tip speed ratio, number of blades,
+element radius, tip flow angle---that do no necessarily translate directly to
+the geometry and flow environment of a cross-flow rotor. Therefore, a more
+general end effects model was sought.
+
+From Prandtl's lifting line theory, the geometric angle of attack $\alpha$ of a
+foil with an arbitrary circulation distribution can be expressed as a function
+of nondimensional span $\theta$ as \cite{Anderson2001}
+\begin{equation}
+    \alpha (\theta) = \frac{2S}{\pi c (\theta)}
+    \sum_1^N A_n \sin \theta
+    + \sum_1^N n A_n \frac{\sin n \theta}{\sin \theta}
+    + \alpha_{L = 0}(\theta),
+    \label{eq:lifting-line}
+\end{equation}
+where $S$ is the span length, $c(\theta)$ is the chord length, and $N$ is the
+number of locations or elements sampled along the foil. This relationship can be
+rearranged into a matrix equation to solve for the unknown Fourier coefficients
+$A_n$,
+\begin{equation}
+    [\alpha_m ] - \alpha_{L=0} = [D_{mn}][A_n],
+\end{equation}
+where
+\begin{equation}
+    D_{mn} = \sum_1^N \left[ \frac{2b}{\pi c_m} \sin n \theta_m + n \frac{\sin n
+        \theta_m}{\sin \theta_m} \right].
+\end{equation}
+
+With the Fourier coefficients, the circulation distribution can be calculated as
+\begin{equation}
+    \Gamma (\theta) = 2SU_\infty \sum_1^N A_n \sin n \theta,
+\end{equation}
+which, via the Kutta--Joukowski theorem, provides the lift coefficient
+distribution
+\begin{equation}
+    C_l(\theta) = \frac{-\Gamma (\theta)}{\frac{1}{2} c U_\infty}.
+\end{equation}
+
+We can therefore compute a correction function based on the normalized spanwise
+lift coefficient distribution
+\begin{equation}
+    F = C_l(\theta)/C_l(\theta)_{\max},
+\end{equation}
+which will be in the range $[0, 1]$, similar to the Glauert corrections, but
+does not contain rotor parameters.
+
+The correction functions for the standard Glauert and lifting line methods are
+plotted for comparison in Figure~\ref{fig:end-effects}. Function values were
+computed using the UNH-RVAT geometry, $\lambda = \lambda_0 = 1.9$, and a 20
+degree tip angle of attack. Both end effects corrections produce similar results
+for this specific case, which gives confidence that the lifting line method will
+produce reasonable results without the uncertainty of translating axial-flow to
+cross-flow rotor parameters.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.75\textwidth]{end-effects_rvat-20-deg}
+
+    \caption{End effect correction function values for the Glauert and lifting
+        line methods using UNH-RVAT parameters at 20 degrees tip angle of attack and
+        $\lambda=1.9$.}
+
+    \label{fig:end-effects}
+\end{figure}
+
+\section{Software implementation}
+
+The US National Renewable Energy Laboratory (NREL) has developed and released an
+actuator line modeling library, SOWFA~\cite{Churchfield2014b}, for simulating
+horizontal-axis wind turbine arrays using the OpenFOAM finite volume CFD
+library. OpenFOAM is free, open-source, widely used throughout industry and
+academia, and has grown a very active support community around itself
+\cite{Bachant2015-CFD-online}. Though SOWFA is also open-source, and has the
+ability to be coupled to NREL's FAST HAWT BEM code, its mostly procedural style
+would have required significant effort and duplicate code to adapt for
+cross-flow turbines. Thus, a new and more general ALM library was developed from
+the ground up that could model both cross- and axial-flow turbines, as well as
+standalone actuator lines. The actuator line model developed here, dubbed
+turbinesFoam \cite{Bachant2016-turbinesFoam-v0.0.7}, was also written as an
+extension library for OpenFOAM, and was developed freely and openly from its
+inception in order to increase community engagement and research efficiency.
+
+turbinesFoam was written in OpenFOAM's style, using OpenFOAM's
+\texttt{fvOptions} framework for adding source terms to equations at
+runtime---see Listing~\ref{lst:fvOptions} for an example implementation within
+the Navier--Stokes' momentum equation. Using the \texttt{fvOptions} framework
+allows the CFT-ALM to be added to many of the solvers included in OpenFOAM,
+meaning it can be readily used with RANS or LES, multiphase models (e.g., for
+simulating the free surface in MHK installations), and even with heat transfer.
+This is in contrast to SOWFA's implementation, which requires custom flow
+solvers to be developed and maintained.
+
+\begin{lstlisting}[float,language=C++,caption=Adding source terms to the momentum equation in OpenFOAM.,label=lst:fvOptions]
+    tmp<fvVectorMatrix> UEqn
+    (
+        fvm::ddt(U)
+        + fvm::div(phi, U)
+        + turbulence->divDevReff(U)
+        ==
+        fvOptions(U)
+    );
+\end{lstlisting}
+
+OpenFOAM and turbinesFoam are written in the C++ programming language, which
+follows the object oriented programming paradigm. This characteristic helped
+modularize the ALM code for increased readability and reuse. In turbinesFoam, a
+turbine is a software object that is composed of actuator line objects, which
+themselves are composed of actuator line element objects. Structuring the code
+this way allows isolation and reuse of the functionality of individual
+components. For example, the actuator line object was written such that is could
+be used outside the turbine context to ensure it produces the correct forcing,
+without adding the complexity of rotation, other actuator lines, etc. that would
+be present in a turbine rotor. The very same actuator line objects can be used
+in both axial-flow and cross-flow rotors, without having to copy code from one
+to the other. In contrast, the actuator line model in SOWFA uses a single
+software object to represent an entire array of turbines, which necessitates
+iterating through many nested lists down to the element level, which can be
+confusing to read.
+
+OpenFOAM's data structures are designed to be inherently parallel via message
+passing interface (MPI). By working within the library infrastructure the ALM
+code was easily parallelized, which will facilitate its deployment on high
+performance computing clusters for large flow simulations.
+
+Since all applications are run from a command line and all input data is text
+based, automation and integration with other tools is relatively
+straightforward. Future enhancements could include coupling with software for
+generating static foil data, e.g., XFOIL~\cite{Drela1989} or other OpenFOAM
+solvers, turbine controller models, structural analysis codes, and optimization
+tools, e.g., SNL's DAKOTA \cite{AdamsBaumanBohnhoffEtAl2009} or
+OpenMDAO~\cite{Gray2010}, for both individual turbines and array layouts.
+
+\section{Results}
+
+Both the high solidity UNH-RVAT and medium-low solidity RM2 turbines were
+modeled using the ALM inside a Reynolds-averaged Navier--Stokes (RANS)
+simulation, closed with the standard $k$--$\epsilon$ turbulence model. These
+rotors provide diverse parameters, which helped evaluate the robustness of the
+ALM. The simulations were performed inside a domain similar in size to that used
+in Chapter~\ref{chap:CFD} to mimic the tow tank (3.66 m wide, 2.44 m tall, 1.52
+m upstream and 2.16 m downstream), with similar boundary conditions (1 m/s
+inflow, no-slip walls and bottom, and a rigid slip for the top). Simulations
+were run for a total of 6 seconds, with the latter half used to calculate
+performance and wake statistics. Pressure--velocity coupling for the momentum
+equation was achieved using the PISO (pressure implicit splitting of operators)
+method. A slice of the mesh in the $x$--$y$ plane is shown in
+Figure~\ref{fig:ALM-mesh}.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.8\textwidth]{thesis_alm-mesh}
+
+    \caption{$x$--$y$ planar slice of the mesh used for the ALM RANS
+        simulations.}
+
+    \label{fig:ALM-mesh}
+\end{figure}
+
+Similar numerical settings were used for each turbine as well. The Sheng et al.
+DS model was used with the default coefficients given in \cite{Sheng2008}, and
+the Goude flow curvature correction was employed. A second order backward
+difference was used for advancing the simulation in time, and second order
+linear schemes were used for the majority of the terms' spatial discretizations.
+The only major difference between the two simulation configurations was that the
+end effects model was deactivated for the RM2, since it reduced $C_P$ far below
+the experimental measurements. This modification is consistent with the RM2
+blades' higher aspect ratio (15 versus the UNH-RVAT's 7.1) and tapered planform,
+though will need to be investigated further. The number of elements per actuator
+line was set to be approximately equal to the total span divided by the Gaussian
+force projection width $\epsilon$. Case files for running all the simulations
+presented here with OpenFOAM 3.0.x are available from
+\cite{Bachant2016-UNH-RVAT-turbinesFoam-v1.0.0,
+Bachant2016-RM2-turbinesFoam-v1.0.0,
+Bachant2016-UNH-RVAT-turbinesFoam-v1.0.0-LES}.
+
+The same foil coefficient data were used for all simulations---those for a NACA
+0021 as reported by Sheldahl and Klimas \cite{Sheldahl1981}. Each rotor's shaft
+was assumed to have a drag coefficient $C_d = 1.1$, and the blade support strut
+end element drag coefficients were set to 0.05, to approximate the effects of
+separation in the corners of the blade--strut connections.
+
+Since the ALM is intended to be an engineering tool when coupled with RANS, it
+was assumed that information about tip speed ratio due to control details---that
+is, sinusoidally oscillating $\lambda$---would not be known a priori, and was
+excluded, unlike the 3-D blade-resolved cases in Chapter~\ref{chap:CFD}. Note
+that a systematic investigation of the effects of sinusoidal $\lambda$ was not
+undertaken, but for the UNH-RVAT a one to two percentage point increase in $C_P$
+was observed when running with similar parameters as the blade-resolved
+simulation.
+
+Compared to 3-D blade-resolved RANS, the ALM can solve a standalone turbine case
+on the order of CPU minutes per second of simulated time versus 1,000 CPU hours
+per second---a savings of 4 orders of magnitude.
+
+\subsection{Verification}
+
+Verification for sensitivity to spatial and temporal grid resolution was
+performed for both the UNH-RVAT and RM2 RANS cases at their optimal tip speed
+rations, the results from which are plotted in
+Figure~\ref{fig:RVAT-ALM-verification} and
+Figure~\ref{fig:RM2-ALM-verification}, respectively. Similar to the verification
+strategy employed in Chapter~\ref{chap:CFD}, the mesh topology was kept
+constant, and the resolution was scaled proportional to the number of cells in
+the $x$-direction $N_x$ for the base hexahedral mesh. Both models displayed low
+sensitivity to the number of time steps per revolution. Spatial grid dependence,
+however, was more important.
+
+Final spatial grid resolutions were chosen as $N_x=48$ for both the UNH-RVAT and
+RM2 cases. Time steps were chosen as $\Delta t = 0.01$ and $\Delta t = 0.005$
+seconds for the UNH-RVAT and RM2 respectively, which correspond to approximately
+200 steps per revolution. The chosen values should provide $C_P$ predictions
+within one percentage point of the true solution, which is on the order of the
+expanded uncertainty of the experimental measurements. Note that for computing
+performance curves, the number of steps per revolution was kept constant, i.e.,
+the time step was adjusted to $\Delta t = \Delta t_0 \lambda_0 / \lambda$.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RVAT-ALM_verification}
+
+    \caption{Temporal (left) and spatial (right) grid resolution sensitivity
+        results for the UNH-RVAT ALM RANS model.}
+
+    \label{fig:RVAT-ALM-verification}
+\end{figure}
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RM2-ALM_verification}
+
+    \caption{Temporal (left) and spatial (right) grid resolution sensitivity
+        results for the RM2 ALM RANS model.}
+
+    \label{fig:RM2-ALM-verification}
+\end{figure}
+
+\subsection{UNH-RVAT RANS}
+
+Power and drag coefficient curves are plotted for the UNH-RVAT in
+Figure~\ref{fig:RVAT-ALM-perf-curves}. The ALM was successful at predicting the
+performance tip speed ratios up to $\lambda_0$, which suggests that dynamic
+stall was being modeled accurately, but $C_P$ was overpredicted at high
+$\lambda$. This may have been caused by the omission of additional parasitic
+drag sources such as roughness from exposed bolt heads located far enough from
+the axis to have a large effect at high rotation rates, or an underestimation of
+the blade--strut connection corner drag coefficient. In Chapter~\ref{chap:RM2}
+we showed how these losses can be significant even with carefully smoothed
+struts and strut-blade connections. Overprediction of performance at high tip
+speed ratio could also be a consequence of the Leishman--Beddoes dynamic stall
+model, which can also be seen in the Darrieus VAWT momentum model results shown
+in Figure 6.70 of \cite{Para2002}.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RVAT-ALM_perf-curves}
+
+    \caption{Power and drag coefficient curves computed for the UNH-RVAT using
+        the actuator line model with RANS.}
+
+    \label{fig:RVAT-ALM-perf-curves}
+\end{figure}
+
+Figure~\ref{fig:RVAT-ALM-meancontquiv} shows mean velocity field for the
+UNH-RVAT computed by the ALM RANS model. The asymmetry was captured well, along
+with some of the vertical flow due to blade tip vortex shedding, though the flow
+structure is missing the detail present in the experiments and blade-resolved
+RANS simulations. Overall, the wake appears to be over-diffused, which could be
+a consequence of the relatively coarse mesh. Note that with the DS and flow
+curvature corrections turned off, the direction of the mean swirling motion
+reverses, which highlights the importance of resolving the correct azimuthal
+location of blade loading.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.9\textwidth]{RVAT-ALM_meancontquiv}
+
+    \caption{UNH-RVAT mean velocity field at $x/D=1$ computed with the ALM
+        coupled with a $k$--$\epsilon$ RANS model.}
+
+    \label{fig:RVAT-ALM-meancontquiv}
+\end{figure}
+
+Turbulence kinetic energy contours (including resolved and modeled energy) are
+shown in Figure~\ref{fig:RVAT-ALM-kcont}. The ALM was able to resolve the
+concentrated area of $k$ on the $+y$ side of the turbine, but the turbulence
+generated by the dynamic stall vortex shedding process is absent. This makes
+sense since in the ALM, the DS model only modulates the body force term in the
+momentum equation, which does not provide a mechanism for mimicking shed
+vortices or turbulence.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RVAT-ALM_kcont}
+
+    \caption{UNH-RVAT turbulence kinetic energy contours at $x/D=1$ predicted by
+        the ALM inside a $k$--$\epsilon$ RANS model.}
+
+    \label{fig:RVAT-ALM-kcont}
+\end{figure}
+
+Profiles of mean streamwise velocity and turbulence kinetic energy are shown in
+Figure~\ref{fig:RVAT-ALM-profiles}. Here the over-diffused or over-recovered
+characteristic of the mean velocity deficit seen in
+Figure~\ref{fig:RVAT-ALM-meancontquiv} is more apparent. This effect is also
+seen in the profile of $k$, where energy is smeared over the center region of
+the rotor.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RVAT-ALM_wake-profiles}
+
+    \caption{Mean streamwise velocity (left) and turbulence kinetic energy
+        (right) profiles at $z/H=0$ for the UNH-RVAT ALM.}
+
+    \label{fig:RVAT-ALM-profiles}
+\end{figure}
+
+Weighted averages for the momentum recovery terms were computed identically as
+they were in Chapter~\ref{chap:CFD}, and are plotted in
+Figure~\ref{fig:RVAT-ALM-recovery} along with the actuator disk (AD) results
+from Chapter~\ref{chap:RVAT-baseline}, 3-D blade-resolved RANS results from
+Chapter~\ref{chap:CFD} and experiments. The most glaring discrepancy is the
+ALM's prediction of positive cross-stream advection, which is caused by the lack
+of detail in the tip vortex shedding. The total for vertical advection, however,
+is close to that predicted by the 3-D blade-resolved Spalart--Allmaras model.
+Levels of turbulent transport due to eddy viscosity and deceleration due to the
+adverse pressure gradient are between those predicted by the 3-D blade-resolved
+$k$--$\omega$ SST and SA models. Overall, however, one might expect the total
+wake recovery rate to be comparable between all models except the actuator disk,
+which induces negative vertical advection, very little turbulent transport, and
+has a positive pressure gradient contribution. These results suggests the ALM
+would be an effective tool---much better than an actuator disk---for assessing
+downstream spacing of subsequent CFTs, though blade--vortex interaction for very
+tightly spaced rotors may not be captured, at least on relatively coarse meshes
+as used here.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RVAT-ALM_recovery-bar-chart}
+
+    \caption{Weighted average momentum recovery terms for the actuator disk (AD)
+    simulation from Chapter~\ref{chap:RVAT-baseline}, UNH-RVAT actuator line
+    model with a $k$--$\epsilon$ RANS closure, the two 3-D blade resolved RANS
+    models described in Chapter~\ref{chap:CFD}, and the experiments in
+    Chapter~\ref{chap:RVAT-baseline}.}
+
+    \label{fig:RVAT-ALM-recovery}
+\end{figure}
+
+\subsection{RM2 RANS}
+
+Figure~\ref{fig:RM2-ALM-perf-curves} shows the performance curves computed for
+the RM2 by the ALM, and those from the tow tank experiments. As with the high
+solidity RVAT, $C_P$ is overpredicted at high $\lambda$. However, $\lambda_0$,
+the tip speed ratio of peak power coefficient, is also shifted to the right.
+This is indicative of inaccurate dynamic stall modeling, which could possibly be
+attributed to one of the models' tuning constants, e.g., the time constant for
+the lagged angle of attack. Limited ad hoc testing revealed that $C_P$ at
+$\lambda_0$ was more accurately predicted with a lagged angle of attack time
+constant roughly double the default value. This could be investigated further by
+looking at the phase of, e.g., maximum $C_P$ peaks compared to the experimental
+values (or possibly those from blade-resolved CFD), but for the present study
+only mean values were considered.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RM2-ALM_perf-curves}
+
+    \caption{Power and drag coefficient curves computed for the RM2 using the
+        ALM.}
+
+    \label{fig:RM2-ALM-perf-curves}
+\end{figure}
+
+Figure~\ref{fig:RM2-ALM-meancontquiv} shows the mean velocity field at 1 m
+downstream or $x/D=0.93$ computed by the ALM for the RM2. The mean near-wake
+structure looks similar to the RVAT ALM RANS case, but for the RM2, the lack of
+detail from blade tip vortex shedding matches more closely with experiments.
+However, the vertical flow towards the $x$--$y$ center plane was captured, which
+is an important qualitative feature of both CFT near-wakes.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.9\textwidth]{RM2-ALM_meancontquiv}
+
+    \caption{Mean velocity field at $x/D=0.93$ for the RM2 predicted by the
+        ALM.}
+
+    \label{fig:RM2-ALM-meancontquiv}
+\end{figure}
+
+Figure~\ref{fig:RM2-ALM-kcont} shows the ALM's turbulence kinetic energy
+predictions in the near-wake of the RM2. Like for the UNH-RVAT, $k$ appears to
+be concentrated on the $+y$ side of the rotor. However, overall levels of
+turbulence are lower than for the UNH-RVAT, which is consistent with the
+experimental results. However, turbulence generated by the RM2's blade tip
+vortex shedding was not captured by the ALM.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.9\textwidth]{RM2-ALM_kcont}
+
+    \caption{Turbulence kinetic energy contours at $x/D=0.93$ behind the RM2
+        predicted by the ALM.}
+
+    \label{fig:RM2-ALM-kcont}
+\end{figure}
+
+Wake profiles at turbine mid-height in the $x$--$y$ center plane are shown in
+Figure~\ref{fig:RM2-ALM-profiles}. Like the UNH-RVAT case, the mean velocity
+deficit appears to be recovering too quickly, which may similarly be due to
+coarseness of the grid or overprediction of the eddy viscosity. The peak in
+turbulence kinetic energy on the $-y$ side of the turbine was also
+underpredicted, though was much lower in magnitude compared to the UNH-RVAT,
+even in the experimental measurements.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RM2-ALM_wake-profiles}
+
+    \caption{Mean streamwise velocity (left) and turbulence kinetic energy
+    (right) profiles at $x/D=0.93$ and $z/H=0$ computed for the RM2 using the
+    ALM inside a $k$--$\epsilon$ RANS model.}
+
+    \label{fig:RM2-ALM-profiles}
+\end{figure}
+
+Finally, a similar mean streamwise momentum transport analysis was undertaken by
+computing weighted sums of each term across the entire domain in the $y$--$z$
+directions. Similar results as for the UNH-RVAT were obtained, i.e.,
+cross-stream advection was predicted to be positive where it should have been
+negative, vertical advection was predicted reasonably well, and turbulent
+transport due to the eddy viscosity was also relatively large.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RM2-ALM_recovery-bar-chart}
+
+    \caption{Weighted average momentum recovery terms at $x/D=0.93$ for the RM2
+        actuator line model with a $k$--$\epsilon$ RANS closure and the experiments
+        described in Chapter~\ref{chap:RM2}}.
+
+    \label{fig:RM2-ALM-recovery}
+\end{figure}
+
+\subsection{UNH-RVAT LES}
+
+The state-of-the-art in high fidelity turbine array modeling uses the actuator
+line method coupled with large eddy simulation (LES), which allows more of the
+turbulent energy spectrum to be directly resolved, only requiring the dynamics
+of the smallest scales---where dissipation occurs---to be computed by the
+so-called subgrid-scale (SGS) model. Since the ALM LES approach has only been
+reported in the literature for a very low Reynolds number 2-D CFT
+\cite{Shamsoddin2014}, and CFTs may provide unique opportunities to array
+optimization, which could be explored with LES, it was of interest to determine
+how well the ALM coupled with LES might predict wake dynamics of a higher $Re$
+3-D CFT rotor.
+
+Thus, the UNH-RVAT baseline case was simulated using the Smagorinsky LES
+turbulence model \cite{Smagorinsky1963}, which was the first of its kind, and
+serves as a good standard for LES modeling since its behavior is well-reported
+in the literature. Default Smagorinsky model coefficients were used, LES filter
+width was set as the cube root of the local cell volume. The tip speed ratio was
+set to oscillate sinusoidally about $\lambda_0$ with a 0.19 magnitude and the
+angle of the first peak at 1.4 radians---similar to the rotation presribed in
+the blade-resolved RANS simulations discussed in Chapter~\ref{chap:CFD}.
+
+Since the computational cost of LES is significantly higher than RANS,
+verification with respect to grid dependence was not performed. Instead, mesh
+resolution was chosen relative to similar studies of turbine wake ALM LES. Of
+the studies surveyed
+\cite{Shamsoddin2014,Archer2013,Martinez-Tossas2015a,Troldborg2007}, the mesh
+resolution ranged from 18--64 points per turbine diameter. The mesh here was set
+accordingly by using a 16 point per meter base mesh, and refining twice in a
+region containing the turbine to produce a 64 point per turbine diameter/height
+resolution. The solver was run with a 0.002 second time step, which is
+significantly within the limit described by \cite{Martinez-Tossas2015}, where an
+actuator line element may not pass through more than one cell per time step.
+With these resolutions computation times were $O(10)$ CPU hours per second of
+simulated time, which is approximately two orders of magnitude lower than for
+blade-resolved RANS.
+
+Mean power coefficient predictions for the UNH-RVAT at its optimal mean tip
+speed ratio dropped to 0.20 using the ALM within the large eddy simulation.
+However, the amount of information regarding the wake dynamics was greatly
+increased, even beyond that of the 3-D blade-resolved RANS.
+Figure~\ref{RVAT-ALM-LES-vorticity} shows an instantaneous snapshot of
+isosurfaces of vorticity produced by the actuator lines.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=\textwidth]{RVAT-ALM-LES_vorticity-snapshot}
+
+    \caption{Snapshot of vorticity isosurfaces (colored by their streamwise
+        component) at $t=6$ s for the UNH-RVAT LES case.}
+
+    \label{RVAT-ALM-LES-vorticity}
+\end{figure}
+
+The near-wake's mean velocity field at $x/D=1$ is shown in
+Figure~\ref{RVAT-ALM-LES-vorticity}. Compared with the RANS ALM results, the LES
+looks much more like the blade-resolved and experimental results, showing the
+clockwise and counterclockwise mean swirling motion on the $-y$ and $+y$ sides
+of the rotor, respectively.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.9\textwidth]{RVAT-ALM-LES_meancontquiv}
+
+    \caption{Mean velocity field in the UNH-RVAT near-wake at $x/D=1$ computed
+        with the Smagorinsky LES model.}
+
+    \label{fig:RVAT-ALM-LES-meancontquiv}
+\end{figure}
+
+Contours of turbulence kinetic energy sampled at $x/D=1$ from the large eddy
+simulation are plotted in Figure~\ref{fig:RVAT-ALM-LES-kcont}. Compared with
+RANS, LES is more able to predict the turbulence generated by the blade tip
+vortex shedding and dynamic stall effects, though the overall level of
+unsteadiness generated was much lower, especially on the $+y$ side of the rotor.
+This is likely a consequence of the SGS modeling, where the vortical structures
+generated by the blades remain stable further downstream. Similar effects were
+seen in \cite{Martinez-Tossas2015a, Shamsoddin2014}, where higher levels of the
+Smagorinsky coefficient delayed vortex breakdown and subsequent higher levels of
+turbulence. Figure~\ref{RVAT-ALM-LES-vorticity} shows evidence of these effects,
+where the blade bound and tip vortices are still relatively coherent at $x/D=1$.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RVAT-ALM-LES_kcont}
+
+    \caption{Turbulence kinetic energy in the UNH-RVAT near-wake at $x/D=1$
+        computed with the Smagorinsky LES model.}
+
+    \label{fig:RVAT-ALM-LES-kcont}
+\end{figure}
+
+Mean velocity profiles at the turbine center plane, plotted in
+Figure~\ref{fig:RVAT-ALM-LES-profiles}, were predicted more accurately using LES
+versus RANS, and rival those of the 3-D blade-resolved models. However,
+turbulence kinetic energy profiles did not match as closely with experiments.
+Though the qualitative shape was resolved better than that by the RANS ALM
+simulation, notably the asymmetric peaks around $y/R = \pm 1$, the turbulence
+generated in the large eddy simulation was approximately an order of magnitude
+too low. Once again this was probably a result of the subgrid-scale modeling and
+its effect on the stability of vortex structures.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RVAT-ALM-LES_wake-profiles}
+
+    \caption{Mean velocity profiles in the UNH-RVAT near-wake at $x/D=1$ and
+        $z/H=0$ computed with the Smagorinsky LES model.}
+
+    \label{fig:RVAT-ALM-LES-profiles}
+\end{figure}
+
+The planar weighted sums of streamwise momentum recovery terms were computed in
+the same way as for the RANS cases with the exception of the turbulent transport
+term, which for the LES was computed from the $x$-components of the divergence
+of the resolved and SGS Reynolds stress tensors:
+\begin{equation}
+    \text{Turb. trans.} = - \left( \frac{\partial}{\partial x_j}
+    \overline{u^\prime_x u^\prime_j}
+    + \frac{\partial}{\partial x_j} R_{xj}
+    \right),
+\end{equation}
+where $u$ indicates the resolved or filtered velocity, and $R$ is the
+subgrid-scale Reynolds stress.
+
+Transport term weighted sums computed from the LES results are shown in
+Figure~\ref{fig:RVAT-ALM-LES-recovery}. Unlike the RANS ALM cases, the
+cross-stream advection contributions are negative, as they are in the experiment
+and blade-resolved CFD models. The vertical advection term is positive as
+expected, though smaller than in other cases. Interestingly, the turbulent
+transport is negative in the LES, meaning the combined effects of the resolved
+and SGS stressed are transferring momentum out of the wake. The low turbulent
+transport appears to be partially balanced by higher levels of viscous
+diffusion---about an order of magnitude larger than the 3-D blade-resolved RANS
+models and experiments. These discrepancies highlight the difficulty of
+predicting the near-wake dynamics, the importance of the SGS model in LES, and
+the need for data further downstream to test and refine predictions for wake
+evolution. For example, setting the Smagorinsky coefficient higher may induce
+vortex breakdown earlier, which would raise the turbulence levels significantly.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.85\textwidth]{RVAT-ALM-LES_recovery-bar-chart}
+
+    \caption{Weighted average momentum recovery terms at $x/D=1$ for the RVAT
+        ALM LES using the Smagorinsky SGS model}.
+
+    \label{fig:RVAT-ALM-LES-recovery}
+\end{figure}
+
+\section{Conclusions}
+
+An actuator line model for cross-flow turbines, including a Leishman--Beddoes
+type dynamic stall model, flow curvature, added mass, and lifting-line based end
+effects corrections, was developed and validated against experimental datasets
+acquired for high and medium-low solidity rotors at scales where the performance
+and near-wake dynamics were essentially Reynolds number independent. When
+coupled to a $k$--$\epsilon$ RANS solver ALM simulations took $O(0.1)$ CPU hours
+per second of simulated time, while when coupled with a Smagorinsky LES model
+the computing time was $O(10)$ hours per second, which represent a four and two
+order of magnitude decrease in computational expense versus 3-D blade-resolved
+RANS, respectively.
+
+The RANS ALM predicted the UNH-RVAT performance well at tip speed ratios up to
+and including that of max power coefficient. The RM2 power coefficient on the
+other hand was underpredicted at lower $\lambda$. Both models overestimated
+$C_P$ at the highest tip speed ratios, which has been observed in other
+simulations using Leishman--Beddoes type dynamic stall models. Possible
+explanations include underestimation of added mass effects or blade--strut
+connection corner drag, incorrect time constants in the LB DS model, and/or
+inaccuracy due to the virtual camber effect. In the present flow curvature
+model, the angle of attack is corrected, but the foil coefficient data is not
+transformed, meaning the LB DS separation point curve fit parameters are equal
+for both positive and negative angles of attack. A foil data transformation
+algorithm based on virtual camber should be investigated for future improvement
+of the ALM.
+
+The RANS ALM cases were able to match some important qualitative near-wake flow
+features, e.g., the mean vertical advection velocity towards the mid-rotor
+plane. However, the mean flow structure and turbulence generation due to blade
+tip and dynamic stall vortex shedding shows some discrepancy with experimental
+and blade-resolved CFD. Extensions to the ALM to deal with these shortcomings
+should be developed, e.g., a turbulence injection model as employed by James et
+al. \cite{James2010} or a model that will ``turn'' the ALM body force vectors to
+approach the effects of leading and trailing edge vortex shedding during dynamic
+stall.
+
+The UNH-RVAT was simulated with the ALM embedded within a typical Smagorinsky
+LES, which thanks to its lower diffusion and/or dissipation was able to more
+accurately capture the large scale vortical flow structures shed by the rotor
+blades. Turbulence generated by the blade tip vortex shedding and dynamic stall
+region of the blade path was better resolved, but overall lower levels of
+turbulence were predicted, which is likely a consequence of the subgrid-scale
+model's influence on the stability of shed vortices. This effect was also
+apparent in the negative predictions of turbulent transport on the streamwise
+momentum recovery. Therefore, subgrid-scale modeling should be investigated
+further before applying the ALM LES to array analyses.
+
+The ALM provides a more physical flow description compared to momentum and
+vortex models, at a reasonable cost. The ALM has an important advantage over
+vortex methods in that the effort solving the flow field remains reasonably
+constant as a simulation progresses, whereas free-wake vortex methods typically
+increase in computational complexity each time step. This is especially
+important for array simulations, where for a given domain adding many turbines
+will be cheap in the ALM compared to the rapidly increasing number of vortex
+elements generated by each turbine. The ALM also drastically reduces
+computational effort compared to blade-resolved CFD, while maintaining the
+unsteadiness of the wake not resolved by a conventional actuator disk.
+Furthermore, its implementation within finite volume CFD provides a convenient
+way to study turbine array siting, since effects like terrain and buildings can
+be included, as can no-slip boundary conditions and boundary layer inflow
+profiles. Since the turbinesFoam library developed here uses OpenFOAM's modular
+\texttt{fvOptions} framework, the ALM body force term can be added as-is to
+compressible or multiphase solvers, for exploring more complicated physics.
+These features would be difficult or impossible to implement in a momentum or
+vortex models.
+
+Ultimately, the ALM retains some of the disadvantages of other blade element
+methods, but is a the next logical step bridging the gap between low- and
+high-fidelity flow modeling, and has been shown to be a method worthy of further
+advancement. The prospect of using ALM simulations to drive down computational
+cost of RANS by roughly four orders of magnitude---enabling 3-D unsteady
+Navier--Stokes simulations to be performed on a typical PC---itself justifies
+further development.
+
+At contemporary levels of computing power, the ALM is a useful tool for studying
+individual turbines when HPC is not feasible---requiring similar resources as
+2-D blade-resolved RANS, but with improved performance predictions. Furthermore,
+if arrays of CFTs advance to commercial scale, the ALM combined with LES
+represents one of the highest fidelity tools available, and will improve as
+turbulence modeling improves. It is therefore expected that this tool will prove
+valuable for both the engineering and research of wind and MHK turbines.
+
+\section{Future work}
+
+The actuator line model shows great potential for use in cross-flow turbine
+engineering, and this study has inspired many avenues for its improvement.
+
+Firstly, since the fundamental premise of the ALM relies on accurate static foil
+coefficient data, it is crucial that more be measured and published openly, to
+allow the exploration of new rotor designs with various profiles. There is some
+doubt regarding the veracity of the Sheldahl and Klimas NACA foil data
+\cite{Bedon2014}. However, competing datasets at large laboratory scale Reynolds
+numbers are rare, even for the standard symmetrical foils. CFD may play a role
+in generating new data, but models must be rigorously validated, given the
+difficulty in predicting stall and its importance on CFT blade loading.
+
+Profile coefficient databases could be expanded to include more dimensions, or
+independent variables. For example, since the ALM will almost always involve
+turbulence modeling, the effects of local turbulence levels on foil coefficient
+data could be included, since stall delay could affect loading significantly,
+especially for a cross-flow turbine.
+
+Dynamic stall model empirical constants should be further investigated for their
+optimal values within the cross-flow turbine context. Though the
+Leishman--Beddoes type models are formulated in terms of a nondimensional time,
+there may be a dependence of the various time constants on the pitch rate, as
+indicated by the improvement in predictions for the RM2 at optimal tip speed
+ratio when using a higher angle of attack lag time constant. Machine learning
+techniques could be employed to ``train'' the constants (and/or their
+$\lambda$-dependence) to one turbine's power curve, and they could be validated
+against the other turbine's data. A more data-intensive alternative could be to
+develop dynamic cyclic profile coefficient databases---either experimentally or
+numerically---for many reduced frequencies, though this would be a difficult
+task, even for a single profile.
+
+Flow curvature corrections should be examined in more detail, and it may be
+beneficial to develop models to transform entire foil datasets to match their
+virtual camber, as suggested by Migliore et al.~\cite{Migliore1980}. It
+also may be advantageous to implement a complementary model to adjust the
+direction of the resultant force vector to more accurately generate the blade's
+shed vorticity.
+
+Rotor angular speed control via generator models could be included to test the
+effectiveness of various control schemes, e.g., sinusoidal $\lambda$ set points,
+which have been shown in some cases to improve mean power coefficient
+\cite{Strom2015}. Free stream velocity sampling could be implemented to
+determine the optimal location for sensing reference velocity used to compute
+tip speed ratio and power coefficient, which can become difficult in sheared or
+otherwise nonuniform flows \cite{Forbush2015}.
+
+Since turbinesFoam is compatible with OpenFOAM's volume of fluid (VOF)
+multiphase flow solver interFoam, the ALM's effectiveness should be assessed for
+predicting the generation and interaction of turbines with surface waves. The
+ADV positioning system developed in Chapter~\ref{chap:exp-setup} could be
+retrofitted with a wave staff to measure and validate the free surface
+disturbance, which could be key to predicting the effects of placing turbines in
+channels at high blockage ratio.
+
+Possibly most importantly, the ALM should be validated with far-wake velocity
+measurements and/or performance data from two CFTs operating in proximity of
+each other. Next, the ALM should be evaluated for its use in modeling CFT arrays
+using both RANS and LES. If either can successfully postdict experimentally
+measured performance of closely-spaced physical model CFTs, they should then be
+coupled with optimization algorithms to automatically explore the design space,
+and ultimately drive down the cost of energy.
+
 \acks These are the acknowledgments

### 2016-04-28 scripts/getfigs.py

@@ -32,7 +32,8 @@ figdirs = {"RVAT-baseline": join(expdir, "RVAT baseline", "Figures"),
                                 "figures"),
            "end-effects": join(gdrive, "Research", "OpenFOAM", "turbinesFoam",
                                "End effects", "figures"),
-           "RM2-tow-tank": join(expdir, "RM2 tow tank", "Figures")}
+           "RM2-tow-tank": join(expdir, "RM2 tow tank", "Figures"),
+           "thesis": join(gdrive, "Research", "PhD thesis", "figures")}
 
 figlists = {"RVAT-baseline": ["meancontquiv.pdf", "kcont.pdf"],
             "CFT-vectors": ["cft-vectors.pdf", "alpha_deg_urel_geom.pdf"],
@@ -46,7 +47,8 @@ figlists = {"RVAT-baseline": ["meancontquiv.pdf", "kcont.pdf"],
                              "kcont.pdf", "recovery-bar-chart.pdf",
                              "vorticity-snapshot.png"],
             "end-effects": ["rvat-20-deg.pdf"],
-            "RM2-tow-tank": ["meancontquiv.pdf", "k_contours.pdf"]}
+            "RM2-tow-tank": ["meancontquiv.pdf", "k_contours.pdf"],
+            "thesis": ["alm-geometry.pdf", "alm-mesh.png"]}
 
 
 if not os.path.isdir("figures"):

### 2016-04-29 paper.tex
Start adding turbine geometry and CFT kinematics

@@
 Paraschivoiu~\cite{Para1988}. However, their poor performance for high solidity
-turbines \cite{Joo2015}, makes them less attractive, particular for marine
-hydrokinetic applications. Their description of the flow field is very crude as
-well, simply computing the momemtum contained within discretized streamtubes,
-which cannot resolve any of the complicated flow structures shed by the rotor.
+(or chord-to-radius ratio $c/R$) turbines \cite{Joo2015}, makes them less
+attractive, particular for marine hydrokinetic applications. Their description
+of the flow field is very crude as well, simply computing the momemtum contained
+within discretized streamtubes, which cannot resolve any of the complicated flow
+structures shed by the rotor.
 
@@
 
+\section{Turbines}
+
+Two turbines were modeled: the high solidity ($c/R=0.28$) UNH Reference
+Vertical-Axis Turbine (UNH-RVAT) and the medium solidity ($c/R=0.07$--$0.12$) US
+Department of Energy/Sandia National Labs Reference Model 2 (RM2) cross-flow
+turbine (at 1:6 scale), both of which are shown in Figure~\ref{fig:turbines}.
+Both rotors are of similar geometric size ($\sim 1$ m) and are constructed from
+symmetrical NACA foils mounted at mid-chord and mid-span.
+
+Open datasets for the performance and near-wake dynamics of the UNH-RVAT and
+RM2, available from \cite{Bachant2014-RVAT-baseline,Bachant2016-RVAT-Re-dep} and
+\cite{Bachant2016-RM2-data}, respectively. have been acquired in the UNH towing
+tank at Reynolds numbers based on turbine diameter $Re_D = 10^6$. Analysis of
+these datasets showed that turbine performance and near-wake characteristics
+were nearly Reynolds number independent in this
+regime~\cite{Bachant2016-Energies,Bachant2016-RM2-paper}.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.6\textwidth]{turbines}
+
+    \caption{Drawings of the turbines used in this study: the UNH-RVAT (left)
+    and DOE/SNL RM2 (right).}
+
+    \label{fig:turbines}
+\end{figure}
+
 \section{Theory}
@@
 of attack are computed by adding the vectors of relative blade motion and the
-local fluid velocity. The blade lift and drag forces are calculated as
+local fluid velocity, a diagram of which is shown in Figure~\ref{fig:vectors}.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[clip, trim=1in 1.5in 1in 0.5in,
+    width=0.4\textwidth]{figures/CFT-vectors_cft-vectors}
+
+    \caption{Vector diagram of velocity and forcing on a cross-flow turbine
+    blade element. Note that the free stream flow is from top to bottom, the
+    blade chord is in line with the tangential velocity (i.e., zero preset
+    pitch), and the drag vector is magnified by a factor of two (approximately,
+    relative to the lift vector) to enhance visibility.}
+
+    \label{fig:vectors}
+\end{figure}
+
+The kinematics of a CFT are parameterized by the tip speed ratio $\lambda =
+\omega R/U_\infty$, where $\omega$ is the shaft's angular velocity, $R$ is the
+maximum rotor radius, and $U_\infty$ is the free stream velocity. Using only
+geometric considerations---that is, ignoring the effect of the rotor forcing and
+assuming the inflow velocity to be equal to that of the free stream---the
+geometric angle of attack $\alpha$ and relative velocity magnitude
+$|\vec{U}_{\mathrm{rel}}|$ can be calculated as a function of the rotor azimuthal
+angle $\theta$, which is plotted for various tip speed ratios in
+Figure~\ref{fig:geom-alpha-urel}.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.75\textwidth]{CFT-vectors_alpha_deg_urel_geom}
+
+    \caption{Geometric angle of attack (left) and relative velocity (right)
+        versus azimuthal angle at various tip speed ratios.}
+
+    \label{fig:geom-alpha-urel}
+\end{figure}
+
+The blade lift and drag forces are calculated as
 \begin{equation}
@@
 
-    \includegraphics[width=0.9\textwidth]{thesis_alm-geometry}
+    \includegraphics[width=0.7\textwidth]{thesis_alm-geometry}
 
@@
 twice the local cell length $\Delta x$ in order to maintain numerical stability.
-
-Jha et al.~\cite{Jha2014} provides guidelines for choosing a projection
-width.
-
 Schito and Zasso \cite{Schito2014} found that a projection $\epsilon$ equal to
-the local mesh length was optimal.
+the local mesh length was optimal. Jha et al.~\cite{Jha2014} investigated the
+ideal projection width for HAWT blades, recommending an equivalent elliptic
+planform be constructed and used to calculate a spanwise $\epsilon$
+distribution.
 
@@
 
-    \includegraphics[width=0.75\textwidth]{end-effects_rvat-20-deg}
+    \includegraphics[width=0.65\textwidth]{end-effects_rvat-20-deg}
 

### 2016-04-29 paper.tex
Periods inside captions and make figs smaller to save space

@@
 
-    \includegraphics[width=0.85\textwidth]{RVAT-ALM_verification}
+    \includegraphics[width=0.7\textwidth]{RVAT-ALM_verification}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RM2-ALM_verification}
+    \includegraphics[width=0.7\textwidth]{RM2-ALM_verification}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RVAT-ALM_perf-curves}
+    \includegraphics[width=0.7\textwidth]{RVAT-ALM_perf-curves}
 
@@
 
-    \includegraphics[width=0.9\textwidth]{RVAT-ALM_meancontquiv}
+    \includegraphics[width=0.75\textwidth]{RVAT-ALM_meancontquiv}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RVAT-ALM_kcont}
+    \includegraphics[width=0.7\textwidth]{RVAT-ALM_kcont}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RVAT-ALM_wake-profiles}
+    \includegraphics[width=0.7\textwidth]{RVAT-ALM_wake-profiles}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RVAT-ALM_recovery-bar-chart}
+    \includegraphics[width=0.7\textwidth]{RVAT-ALM_recovery-bar-chart}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RM2-ALM_perf-curves}
+    \includegraphics[width=0.7\textwidth]{RM2-ALM_perf-curves}
 
@@
 
-    \includegraphics[width=0.9\textwidth]{RM2-ALM_meancontquiv}
+    \includegraphics[width=0.75\textwidth]{RM2-ALM_meancontquiv}
 
@@
 
-    \includegraphics[width=0.9\textwidth]{RM2-ALM_kcont}
+    \includegraphics[width=0.7\textwidth]{RM2-ALM_kcont}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RM2-ALM_wake-profiles}
+    \includegraphics[width=0.7\textwidth]{RM2-ALM_wake-profiles}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RM2-ALM_recovery-bar-chart}
+    \includegraphics[width=0.7\textwidth]{RM2-ALM_recovery-bar-chart}
 
@@
         actuator line model with a $k$--$\epsilon$ RANS closure and the experiments
-        described in Chapter~\ref{chap:RM2}}.
+        described in Chapter~\ref{chap:RM2}.}
 
@@
 
-    \includegraphics[width=\textwidth]{RVAT-ALM-LES_vorticity-snapshot}
+    \includegraphics[width=0.8\textwidth]{RVAT-ALM-LES_vorticity-snapshot}
 
@@
 
-    \includegraphics[width=0.9\textwidth]{RVAT-ALM-LES_meancontquiv}
+    \includegraphics[width=0.75\textwidth]{RVAT-ALM-LES_meancontquiv}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RVAT-ALM-LES_kcont}
+    \includegraphics[width=0.7\textwidth]{RVAT-ALM-LES_kcont}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RVAT-ALM-LES_wake-profiles}
+    \includegraphics[width=0.7\textwidth]{RVAT-ALM-LES_wake-profiles}
 
@@
 
-    \includegraphics[width=0.85\textwidth]{RVAT-ALM-LES_recovery-bar-chart}
+    \includegraphics[width=0.7\textwidth]{RVAT-ALM-LES_recovery-bar-chart}
 
     \caption{Weighted average momentum recovery terms at $x/D=1$ for the RVAT
-        ALM LES using the Smagorinsky SGS model}.
+        ALM LES using the Smagorinsky SGS model.}
 

### 2016-04-29 paper.tex
Add two introductory paragraphs; resolves #11

@@
 
-Despite its apparent---but not completely certain---effectiveness for predicting
-the performance and wake of a single turbine, blade-resolved CFD presents a huge
-computational expense since it must resolve fine details of the blade boundary
-layers, and must be performed in three dimensions, which will preclude its use
+Vertical-axis (cross-flow) turbines (VATs or CFTs) were the subject of
+significant research and development in the 1970s through the 1990s by groups
+like Sandia National Labs in the US~\cite{Sutherland2012} and the National
+Research Council of Canada~\cite{Para2002}. Despite minor commercial success for
+large scale onshore wind applications, vertical-axis turbines were virtually
+abandoned in favor of horizontal-axis (or axial-flow) turbines, as they
+generally are more efficient and don't encounter the high levels of fatigue
+loading that VATs do. Today, however, there is renewed interest in VATs for
+marine hydrokinetic (MHK) applications~\cite{ORPC2012}, urban
+wind~\cite{Awan2013}, and smaller scale, tightly spaced wind
+farms~\cite{Dabiri2011, Kinzel2012}, thanks to their relatively faster wake
+recovery.
+
+The mean near-wake structure of vertical-axis turbine has been shown to be
+largely dominated by the effects of tip vortex shedding, which induces levels of
+recovery due to vertical advection significantly larger than those from
+turbulent fluctuations~\cite{Bachant2015-JoT}---an effect not seen in HAT wakes.
+As interest shifts to designing and analyzing arrays of VATs, it is necessary to
+determine the effectiveness of various numerical modeling techniques to
+replicate VAT near-wake dynamics, such that wake recovery is accurately
+predicted, leading to accurate assessment of optimal array spacing.
+
+Reynolds-averaged Navier--Stokes (RANS) turbulence models computed on 3-D
+body-fitted (blade-resolved) grids can do a good job predicting the mean
+performance and near-wake structure of a VAT, though their effectiveness depends
+on the turbulence model applied~\cite{Bachant2016-BR-CFD}. However, 3-D
+blade-resolved RANS presents a huge computational expense---on the order of
+1,000 CPU hours per simulated second with contemporary hardware---since it must
+resolve fine details of the blade boundary layers, which will preclude its use
 for array analysis until the availability of computing power increases
@@
 (or chord-to-radius ratio $c/R$) turbines \cite{Joo2015}, makes them less
-attractive, particular for marine hydrokinetic applications. Their description
-of the flow field is very crude as well, simply computing the momemtum contained
-within discretized streamtubes, which cannot resolve any of the complicated flow
-structures shed by the rotor.
+attractive, particular for marine hydrokinetic applications, where rotors of the
+same size will encounter approximately an order of magnitude higher torque
+compared to wind turbines. Their description of the flow field is very crude as
+well, simply computing the momemtum contained within discretized streamtubes,
+which cannot resolve any of the complicated flow structures shed by the rotor.
 

### 2016-04-29 paper.tex
Text wrapping
Clarify BCs

@@
 geometric angle of attack $\alpha$ and relative velocity magnitude
-$|\vec{U}_{\mathrm{rel}}|$ can be calculated as a function of the rotor azimuthal
-angle $\theta$, which is plotted for various tip speed ratios in
+$|\vec{U}_{\mathrm{rel}}|$ can be calculated as a function of the rotor
+azimuthal angle $\theta$, which is plotted for various tip speed ratios in
 Figure~\ref{fig:geom-alpha-urel}.
@@
 in Chapter~\ref{chap:CFD} to mimic the tow tank (3.66 m wide, 2.44 m tall, 1.52
-m upstream and 2.16 m downstream), with similar boundary conditions (1 m/s
-inflow, no-slip walls and bottom, and a rigid slip for the top). Simulations
-were run for a total of 6 seconds, with the latter half used to calculate
-performance and wake statistics. Pressure--velocity coupling for the momentum
-equation was achieved using the PISO (pressure implicit splitting of operators)
-method. A slice of the mesh in the $x$--$y$ plane is shown in
-Figure~\ref{fig:ALM-mesh}.
+m upstream and 2.16 m downstream), with similar velocity boundary conditions: 1
+m/s inflow, no-slip (fixed to 1 m/s) walls and bottom, and a rigid slip
+condition for the top, to approximate some effects of the tow tank's free
+surface. Simulations were run for a total of 6 seconds, with the latter half
+used to calculate performance and wake statistics. Pressure--velocity coupling
+for the momentum equation was achieved using the PISO (pressure implicit
+splitting of operators) method. A slice of the mesh in the $x$--$y$ plane is
+shown in Figure~\ref{fig:ALM-mesh}.
 

### 2016-04-29 paper.tex
Add word 'experiments'

--- 2016-04-29 paper.tex
@@
 ALM. The simulations were performed inside a domain similar in size to that used
-in Chapter~\ref{chap:CFD} to mimic the tow tank (3.66 m wide, 2.44 m tall, 1.52
-m upstream and 2.16 m downstream), with similar boundary conditions (1 m/s
-inflow, no-slip walls and bottom, and a rigid slip for the top). Simulations
-were run for a total of 6 seconds, with the latter half used to calculate
-performance and wake statistics. Pressure--velocity coupling for the momentum
-equation was achieved using the PISO (pressure implicit splitting of operators)
-method. A slice of the mesh in the $x$--$y$ plane is shown in
-Figure~\ref{fig:ALM-mesh}.
+in Chapter~\ref{chap:CFD} to mimic the tow tank experiments (3.66 m wide, 2.44 m
+tall, 1.52 m upstream and 2.16 m downstream), with similar velocity boundary
+conditions: 1 m/s inflow, no-slip (fixed to 1 m/s) walls and bottom, and a rigid
+slip condition for the top, to approximate some effects of the tow tank's free
+surface. Simulations were run for a total of 6 seconds, with the latter half
+used to calculate performance and wake statistics. Pressure--velocity coupling
+for the momentum equation was achieved using the PISO (pressure implicit
+splitting of operators) method. A slice of the mesh in the $x$--$y$ plane is
+shown in Figure~\ref{fig:ALM-mesh}.
 

### 2016-04-29 paper.tex
Replace all chapter refs with papers; resolves #14

--- 2016-04-29 paper.tex
@@
 complicated meshes with many fine cells to resolve the boundary layers near the
-blade surfaces. As shown in Chapter~\ref{chap:RVAT-baseline}, the conventional
-uniform actuator disk is not a good candidate for a cross-flow turbine wake
-generator, never mind the fact that it does not typically compute performance
-predictions. There are then two actuator methods that use blade element theory
-to compute blade loading and therefore generate a wake that more closely
-resembles that of an actual turbine. These are the actuator cylinder or
-swept-surface model (ASSM) and the actuator line model (ALM). The ASSM solves
-for the average blade loading along its path and applies this as a constant body
-force term in the momentum equation. The ALM takes a similar approach but is an
-unsteady method, resolving the blade element locations in time.
+blade surfaces. As shown in \cite{Bachant2015-JoT}, the conventional uniform
+actuator disk is not a good candidate for a cross-flow turbine wake generator,
+never mind the fact that it does not typically compute performance predictions.
+There are then two actuator methods that use blade element theory to compute
+blade loading and therefore generate a wake that more closely resembles that of
+an actual turbine. These are the actuator cylinder or swept-surface model (ASSM)
+and the actuator line model (ALM). The ASSM solves for the average blade loading
+along its path and applies this as a constant body force term in the momentum
+equation. The ALM takes a similar approach but is an unsteady method, resolving
+the blade element locations in time.
 
@@
 ALM. The simulations were performed inside a domain similar in size to that used
-in Chapter~\ref{chap:CFD} to mimic the tow tank (3.66 m wide, 2.44 m tall, 1.52
-m upstream and 2.16 m downstream), with similar boundary conditions (1 m/s
-inflow, no-slip walls and bottom, and a rigid slip for the top). Simulations
-were run for a total of 6 seconds, with the latter half used to calculate
-performance and wake statistics. Pressure--velocity coupling for the momentum
-equation was achieved using the PISO (pressure implicit splitting of operators)
-method. A slice of the mesh in the $x$--$y$ plane is shown in
-Figure~\ref{fig:ALM-mesh}.
+in \cite{Bachant2016-BR-CFD} to mimic the tow tank experiments (3.66 m wide,
+2.44 m tall, 1.52 m upstream and 2.16 m downstream), with similar velocity
+boundary conditions: 1 m/s inflow, no-slip (fixed to 1 m/s) walls and bottom,
+and a rigid slip condition for the top, to approximate some effects of the tow
+tank's free surface. Simulations were run for a total of 6 seconds, with the
+latter half used to calculate performance and wake statistics.
+Pressure--velocity coupling for the momentum equation was achieved using the
+PISO (pressure implicit splitting of operators) method. A slice of the mesh in
+the $x$--$y$ plane is shown in Figure~\ref{fig:ALM-mesh}.
 
@@
 is, sinusoidally oscillating $\lambda$---would not be known a priori, and was
-excluded, unlike the 3-D blade-resolved cases in Chapter~\ref{chap:CFD}. Note
-that a systematic investigation of the effects of sinusoidal $\lambda$ was not
-undertaken, but for the UNH-RVAT a one to two percentage point increase in $C_P$
-was observed when running with similar parameters as the blade-resolved
-simulation.
+excluded, unlike the 3-D blade-resolved cases reported in
+\cite{Bachant2016-BR-CFD}. Note that a systematic investigation of the effects
+of sinusoidal $\lambda$ was not undertaken, but for the UNH-RVAT a one to two
+percentage point increase in $C_P$ was observed when running with similar
+parameters as the blade-resolved simulation.
 
 Compared to 3-D blade-resolved RANS, the ALM can solve a standalone turbine case
-on the order of CPU minutes per second of simulated time versus 1,000 CPU hours
-per second---a savings of 4 orders of magnitude.
+on the order of 0.1 CPU hours per second of simulated time versus 1,000 CPU
+hours per second---a savings of 4 orders of magnitude.
 
@@
 Figure~\ref{fig:RM2-ALM-verification}, respectively. Similar to the verification
-strategy employed in Chapter~\ref{chap:CFD}, the mesh topology was kept
+strategy employed in \cite{Bachant2016-BR-CFD}, the mesh topology was kept
 constant, and the resolution was scaled proportional to the number of cells in
@@
 the axis to have a large effect at high rotation rates, or an underestimation of
-the blade--strut connection corner drag coefficient. In Chapter~\ref{chap:RM2}
-we showed how these losses can be significant even with carefully smoothed
-struts and strut-blade connections. Overprediction of performance at high tip
-speed ratio could also be a consequence of the Leishman--Beddoes dynamic stall
-model, which can also be seen in the Darrieus VAWT momentum model results shown
-in Figure 6.70 of \cite{Para2002}.
+the blade--strut connection corner drag coefficient. In
+\cite{Rawlings2008,Bachant2016-RM2-paper} it was shown how these losses can be
+significant even with carefully smoothed struts and strut-blade connections.
+Overprediction of performance at high tip speed ratio could also be a
+consequence of the Leishman--Beddoes dynamic stall model, which can also be seen
+in the Darrieus VAWT momentum model results shown in Figure 6.70 of
+\cite{Para2002}.
 
@@
 
-Weighted averages for the momentum recovery terms were computed identically as
-they were in Chapter~\ref{chap:CFD}, and are plotted in
-Figure~\ref{fig:RVAT-ALM-recovery} along with the actuator disk (AD) results
-from Chapter~\ref{chap:RVAT-baseline}, 3-D blade-resolved RANS results from
-Chapter~\ref{chap:CFD} and experiments. The most glaring discrepancy is the
-ALM's prediction of positive cross-stream advection, which is caused by the lack
-of detail in the tip vortex shedding. The total for vertical advection, however,
-is close to that predicted by the 3-D blade-resolved Spalart--Allmaras model.
-Levels of turbulent transport due to eddy viscosity and deceleration due to the
-adverse pressure gradient are between those predicted by the 3-D blade-resolved
-$k$--$\omega$ SST and SA models. Overall, however, one might expect the total
-wake recovery rate to be comparable between all models except the actuator disk,
-which induces negative vertical advection, very little turbulent transport, and
-has a positive pressure gradient contribution. These results suggests the ALM
-would be an effective tool---much better than an actuator disk---for assessing
-downstream spacing of subsequent CFTs, though blade--vortex interaction for very
-tightly spaced rotors may not be captured, at least on relatively coarse meshes
-as used here.
+Weighted averages for the terms in the streamwise momentum equation were
+computed identically as they were in \cite{Bachant2015-JoT,Bachant2016-BR-CFD},
+and are plotted in Figure~\ref{fig:RVAT-ALM-recovery} along with the actuator
+disk (AD) results from \cite{Bachant2015-JoT}, 3-D blade-resolved RANS results
+from \cite{Bachant2016-BR-CFD} and experiments. The most glaring discrepancy is
+the ALM's prediction of positive cross-stream advection, which is caused by the
+lack of detail in the tip vortex shedding. The total for vertical advection,
+however, is close to that predicted by the 3-D blade-resolved Spalart--Allmaras
+model. Levels of turbulent transport due to eddy viscosity and deceleration due
+to the adverse pressure gradient are between those predicted by the 3-D
+blade-resolved $k$--$\omega$ SST and SA models. Overall, however, one might
+expect the total wake recovery rate to be comparable between all models except
+the actuator disk, which induces negative vertical advection, very little
+turbulent transport, and has a positive pressure gradient contribution. These
+results suggests the ALM would be an effective tool---much better than an
+actuator disk---for assessing downstream spacing of subsequent CFTs, though
+blade--vortex interaction for very tightly spaced rotors may not be captured, at
+least on relatively coarse meshes as used here.
 
@@
     \caption{Weighted average momentum recovery terms for the actuator disk (AD)
-    simulation from Chapter~\ref{chap:RVAT-baseline}, UNH-RVAT actuator line
-    model with a $k$--$\epsilon$ RANS closure, the two 3-D blade resolved RANS
-    models described in Chapter~\ref{chap:CFD}, and the experiments in
-    Chapter~\ref{chap:RVAT-baseline}.}
+        simulation from \cite{Bachant2015-JoT}, UNH-RVAT actuator line model with a
+        $k$--$\epsilon$ RANS closure, the two 3-D blade resolved RANS models
+        described in \cite{Bachant2016-BR-CFD}, and the experiments reported in
+        \cite{Bachant2015-JoT}.}
 
@@
         actuator line model with a $k$--$\epsilon$ RANS closure and the experiments
-        described in Chapter~\ref{chap:RM2}.}
+        described in \cite{Bachant2016-RM2-paper}.}
 
@@
 angle of the first peak at 1.4 radians---similar to the rotation presribed in
-the blade-resolved RANS simulations discussed in Chapter~\ref{chap:CFD}.
+the blade-resolved RANS simulations discussed in \cite{Bachant2016-BR-CFD}.
 
@@
 multiphase flow solver interFoam, the ALM's effectiveness should be assessed for
-predicting the generation and interaction of turbines with surface waves. The
-ADV positioning system developed in Chapter~\ref{chap:exp-setup} could be
-retrofitted with a wave staff to measure and validate the free surface
-disturbance, which could be key to predicting the effects of placing turbines in
-channels at high blockage ratio.
+predicting the generation and interaction of turbines with surface waves. Tow
+tank experiments could be performed with a wave gauge to measure and validate
+the free surface disturbance, which could be key to predicting the effects of
+placing turbines in channels at high blockage ratio.
 

### 2016-04-29 paper.tex
Mention validation against two turbines in the intro

@@
 combination of high-fidelity flow modeling that includes performance
-predictions, but with reduced computational expense. Here we will develop and
-evaluate the effectiveness of an ALM for cross-flow turbines inside both RANS
-and LES simulations, implemented as an extension for the OpenFOAM open-source
-CFD library.
+predictions, but with reduced computational expense. Here we have developed an
+ALM for cross-flow turbines inside both RANS and LES simulations, implemented as
+an extension for the OpenFOAM open-source CFD library. This model was validated
+against experimental data for both a high \cite{Bachant2015-JoT} and a medium
+\cite{Bachant2016-RM2-paper} solidity vertical-axis turbine.
 

### 2016-04-29 paper.tex
Used --> modeled
Add unsteadiness to theory section; resolves #10

@@
 
-    \caption{Drawings of the turbines used in this study: the UNH-RVAT (left)
+    \caption{Drawings of the turbines modeled in this study: the UNH-RVAT (left)
     and DOE/SNL RM2 (right).}
@@
 
-The blade lift and drag forces are calculated as
+Inspecting Figure~\ref{fig:geom-alpha-urel}, it is apparent that CFTs are
+inherently unsteady machines, encountering large amplitude and rapid
+oscillations in angle of attack, often exceeding the rotor blades' static stall
+angles in typical operation. The unsteadiness can be characterized by a reduced
+frequency~\cite{Leishman2006}
+\begin{equation}
+    k = \frac{\omega c}{2 U_\infty},
+\end{equation}
+which assumes the free stream velocity is constant. Unsteady effects begin to
+become significant for $k > 0.05$, and can become dominant for $k \ge 0.2$.
+
+For a VAT or CFT reduced frequency can be reformulated in terms of the tip speed
+ratio as
+\begin{equation}
+    k = \frac{\lambda c}{2R},
+\end{equation}
+which is then also a function of solidity or chord-to-radius ratio $c/R$. As an
+example, a large scale, relatively low solidity Darrieus wind turbine such as
+the Sandia 34 m diameter Test Bed, with an equatorial blade chord of 0.91 m
+\cite{Murray2011}, a reduced frequency $k=0.16$ is encountered based solely on
+angle of attack oscillations at $\lambda=6$. For a smaller scale CFT, e.g., with
+$c/R = 0.25$, operating at $\lambda = 2$, the reduced frequency is 0.25. It
+follows that unsteady effects will be significant for cross-flow turbines even
+in the absence of nonlinear effects such as stall.
+
+Assuming unsteady effects can be appropriately modeled, the blade element lift
+and drag forces are calculated as
 \begin{equation}

### 2016-04-29 paper.tex
Add acknowledgments

@@
 
-\acks These are the acknowledgments
+\acks The authors would like to acknowledge funding through a National Science
+Foundation CAREER award (principal investigator Martin Wosnik, NSF 1150797,
+Energy for Sustainability, program manager Gregory L. Rorrer). This work has
+been carried out as part of author P. Bachant's PhD research at the University
+of New Hampshire.
 

### 2016-05-02 paper.tex
Add offshore wind refs

@@
 loading that VATs do. Today, however, there is renewed interest in VATs for
-marine hydrokinetic (MHK) applications~\cite{ORPC2012}, urban
-wind~\cite{Awan2013}, and smaller scale, tightly spaced wind
-farms~\cite{Dabiri2011, Kinzel2012}, thanks to their relatively faster wake
-recovery.
+marine hydrokinetic (MHK) applications~\cite{ORPC2012}, offshore floating wind
+farms~\cite{Paulsen2011, Sandia2012, Dodd2014}, and smaller scale, tightly
+spaced wind farms~\cite{Dabiri2011, Kinzel2012}, thanks to their relatively
+faster wake recovery.
 

### 2016-05-02 paper.tex
Define solidity

@@
 models, such as the double multiple streamtube (DMST) approached developed by
-Paraschivoiu~\cite{Para1988}. However, their poor performance for high solidity
-(or chord-to-radius ratio $c/R$) turbines \cite{Joo2015}, makes them less
-attractive, particular for marine hydrokinetic applications, where rotors of the
-same size will encounter approximately an order of magnitude higher torque
-compared to wind turbines. Their description of the flow field is very crude as
-well, simply computing the momemtum contained within discretized streamtubes,
-which cannot resolve any of the complicated flow structures shed by the rotor.
+Paraschivoiu~\cite{Para1988}. However, their poor performance for rotors with
+high solidity~\cite{Joo2015} $Nc/R$ (and/or chord-to-radius ratio $c/R$), where
+$N$ is the number of blades, $c$ is the chord length, and $R$ is the maximum
+rotor radius, makes them less attractive, particular for marine hydrokinetic
+applications where rotors of the same size will encounter approximately an order
+of magnitude higher torque compared to wind turbines. Their description of the
+flow field is very crude as well, simply computing the momentum contained within
+discretized streamtubes, which cannot resolve any of the complicated flow
+structures shed by the rotor.
 

### 2016-05-02 paper.tex
Mention streamwise induction factor

--- 2016-05-02 paper.tex
@@
 models, such as the double multiple streamtube (DMST) approached developed by
-Paraschivoiu~\cite{Para1988}. However, their poor performance for high solidity
-(or chord-to-radius ratio $c/R$) turbines \cite{Joo2015}, makes them less
-attractive, particular for marine hydrokinetic applications, where rotors of the
-same size will encounter approximately an order of magnitude higher torque
-compared to wind turbines. Their description of the flow field is very crude as
-well, simply computing the momemtum contained within discretized streamtubes,
-which cannot resolve any of the complicated flow structures shed by the rotor.
+Paraschivoiu~\cite{Para1988}. However, their poor performance for high loading
+or streamwise induction factors, common for rotors with high
+solidity~\cite{Joo2015} $Nc/R$ (and/or chord-to-radius ratio $c/R$), where $N$
+is the number of blades, $c$ is the chord length, and $R$ is the maximum rotor
+radius, makes them less attractive, particular for marine hydrokinetic
+applications where rotors of the same size will encounter approximately an order
+of magnitude higher torque compared to wind turbines. Their description of the
+flow field is very crude as well, simply computing the momentum contained within
+discretized streamtubes, which cannot resolve any of the complicated flow
+structures shed by the rotor, limiting their applicability for array analysis.
 

### 2016-05-02 paper.tex
Clarify that we are talking about potential flow vortex methods

@@
 Blade element based vortex line methods can be solved with reasonable
-computational effort---ranging from an expense close to BEM up to that of 2-D
-blade-resolved CFD. The cost is a function of the number of vortex elements,
-which for free vortex methods increases at each time step, slowing down the
-calculation as it marches forward in time. Vortex methods can resolve more flow
-details than BEM. However, since the flow is governed by the Laplace equation,
-nonlinear effects such as the turbulent transport will not be included, which we
-have determined in experiments to be the same order of magnitude as the mean
-vertical advection.
+computational effort when coupled with potential flow models---ranging from an
+expense close to BEM up to that of 2-D blade-resolved CFD. The cost is a
+function of the number of vortex elements, which for free vortex methods may
+increase at each time step, slowing down the calculation as it marches forward
+in time, unless there is an additional model for vortex diffusion. Potential
+flow vortex methods can resolve more flow details than BEM. However, since the
+flow is governed by the Laplace equation, nonlinear effects such as the
+turbulent transport will not be included, which we have determined in
+experiments to be the same order of magnitude as the mean vertical advection.
 

### 2016-05-02 paper.tex
Mention NS based vortex method
Add clarification to BE diagram

@@
 for AFTs is in part due to the quasi-steady nature of the flow in the blade
-reference frame, and the relatively rare occurrence of stall.
+reference frame, and the relatively rare occurrence of stall. Note that a
+similar method can be used with the Navier--Stokes equations in
+vorticity--velocity form~\cite{Scheurich2011b}, which may be more efficient when
+vorticity is confined to a relatively small fraction of the domain, e.g., for a
+standalone turbine with a uniform laminar inflow.
 
@@
     \caption{Vector diagram of velocity and forcing on a cross-flow turbine
-    blade element. Note that the free stream flow is from top to bottom, the
-    blade chord is in line with the tangential velocity (i.e., zero preset
-    pitch), and the drag vector is magnified by a factor of two (approximately,
-    relative to the lift vector) to enhance visibility.}
+    blade element. Note that the free stream velocity $U_\infty$ is from top to
+    bottom (identical to $U_in$ for purely geometric calculations), the blade
+    chord is in line with the tangential velocity (i.e., zero preset pitch,
+    which would offset the geometric angle of attack $\alpha$), and the drag
+    vector is magnified by a factor of two (approximately, relative to the lift
+    vector) to enhance visibility.}
 

### 2016-05-02 paper.tex
Add to AL geometry diagram caption

@@
 
-    \caption{Actuator line geometry. Filled circles indicate geometry points
-        whereas squares indicate actuator line element locations.}
+    \caption{Actuator line geometry. Filled circles indicate points where
+    geometry is defined whereas squares indicate actuator line element
+    quarter-chord locations. Chord mount offset is defined as the distance from
+    an element's leading edge to each geometry point in chord lengths, which is
+    0.5 for both turbines in this study.}
 

### 2016-05-02 paper.tex
Note that U_rel is the projection onto the element profile cross-section

@@
 planform area (span $\times$ chord), $\vec{U}_\mathrm{rel}$ is the local
-relative velocity, and $C_l$ and $C_d$ are the sectional lift and drag
-coefficients, linearly interpolated from a table per the local angle of attack.
-The forces are then projected onto the rotor coordinate system to calculate
-torque, overall drag, etc. Forces from the turbine shaft and blade support
-struts are computed in a similar way. After the force on the actuator lines from
-the flow is computed, it is then added to the Navier--Stokes equations as a body
-force or momentum source (per unit density, assuming incompressible flow):
+relative velocity projected onto the plane of the element profile cross-section,
+and $C_l$ and $C_d$ are the sectional lift and drag coefficients, linearly
+interpolated from a table per the local angle of attack. The forces are then
+projected onto the rotor coordinate system to calculate torque, overall drag,
+etc. Forces from the turbine shaft and blade support struts are computed in a
+similar way. After the force on the actuator lines from the flow is computed, it
+is then added to the Navier--Stokes equations as a body force or momentum source
+(per unit density, assuming incompressible flow):
 \begin{equation}

### 2016-05-02 paper.tex
Note that the spanwise flow is neglected

--- 2016-05-02 paper.tex
@@
 planform area (span $\times$ chord), $\vec{U}_\mathrm{rel}$ is the local
-relative velocity, and $C_l$ and $C_d$ are the sectional lift and drag
-coefficients, linearly interpolated from a table per the local angle of attack.
-The forces are then projected onto the rotor coordinate system to calculate
-torque, overall drag, etc. Forces from the turbine shaft and blade support
-struts are computed in a similar way. After the force on the actuator lines from
-the flow is computed, it is then added to the Navier--Stokes equations as a body
-force or momentum source (per unit density, assuming incompressible flow):
+relative velocity projected onto the plane of the element profile cross-section
+(i.e., the spanwise component is neglected), and $C_l$ and $C_d$ are the
+sectional lift and drag coefficients, linearly interpolated from a table per the
+local angle of attack. The forces are then projected onto the rotor coordinate
+system to calculate torque, overall drag, etc. Forces from the turbine shaft and
+blade support struts are computed in a similar way. After the force on the
+actuator lines from the flow is computed, it is then added to the Navier--Stokes
+equations as a body force or momentum source (per unit density, assuming
+incompressible flow):
 \begin{equation}

### 2016-05-02 paper.tex
Note flow curvature correction was derived in Goude and not necessarily by Goude
Be more specific about potential flow vortex models

@@
 Furthermore, this effect is more pronounced for high solidity ($c/R$) turbines.
-Two different flow curvature corrections were considered: one by Goude
+Two different flow curvature corrections were considered: one derived in Goude
 \cite{Goude2012} and one by Mandal and Burton \cite{Mandal1994}.
@@
 The ALM provides a more physical flow description compared to momentum and
-vortex models, at a reasonable cost. The ALM has an important advantage over
-vortex methods in that the effort solving the flow field remains reasonably
-constant as a simulation progresses, whereas free-wake vortex methods typically
-increase in computational complexity each time step. This is especially
-important for array simulations, where for a given domain adding many turbines
-will be cheap in the ALM compared to the rapidly increasing number of vortex
-elements generated by each turbine. The ALM also drastically reduces
+potential flow vortex models, at a reasonable cost. The ALM has an important
+advantage over vortex methods in that the effort solving the flow field remains
+reasonably constant as a simulation progresses, whereas free-wake vortex methods
+typically increase in computational complexity each time step. This is
+especially important for array simulations, where for a given domain adding many
+turbines will be cheap in the ALM compared to the rapidly increasing number of
+vortex elements generated by each turbine. The ALM also drastically reduces
 computational effort compared to blade-resolved CFD, while maintaining the

### 2016-05-02 paper.tex
Remove comment about free-wake vortex method computational cost

@@
 The ALM provides a more physical flow description compared to momentum and
-potential flow vortex models, at a reasonable cost. The ALM has an important
-advantage over vortex methods in that the effort solving the flow field remains
-reasonably constant as a simulation progresses, whereas free-wake vortex methods
-typically increase in computational complexity each time step. This is
-especially important for array simulations, where for a given domain adding many
-turbines will be cheap in the ALM compared to the rapidly increasing number of
-vortex elements generated by each turbine. The ALM also drastically reduces
-computational effort compared to blade-resolved CFD, while maintaining the
-unsteadiness of the wake not resolved by a conventional actuator disk.
+potential flow vortex models, at a reasonable cost. The ALM also drastically
+reduces computational effort compared to blade-resolved CFD, while maintaining
+the unsteadiness of the wake not resolved by a conventional actuator disk.
 Furthermore, its implementation within finite volume CFD provides a convenient

### 2016-05-02 paper.tex
Use past tense

@@
     quarter-chord locations. Chord mount offset is defined as the distance from
-    an element's leading edge to each geometry point in chord lengths, which is
-    0.5 for both turbines in this study.}
+    an element's leading edge to each geometry point in chord lengths, which was
+    0.5 for both rotors modeled in this study.}
 

### 2016-05-02 paper.tex
Add refs for potential flow vortex models; resolves #19

@@
 
-Blade element based vortex line methods can be solved with reasonable
-computational effort when coupled with potential flow models---ranging from an
-expense close to BEM up to that of 2-D blade-resolved CFD. The cost is a
-function of the number of vortex elements, which for free vortex methods may
-increase at each time step, slowing down the calculation as it marches forward
-in time, unless there is an additional model for vortex diffusion. Potential
-flow vortex methods can resolve more flow details than BEM. However, since the
-flow is governed by the Laplace equation, nonlinear effects such as the
-turbulent transport will not be included, which we have determined in
-experiments to be the same order of magnitude as the mean vertical advection.
+Blade element based vortex line methods, e.g., \cite{Strickland1981,
+Murray2011}, can be solved with reasonable computational effort when coupled
+with potential flow models---ranging from an expense close to BEM up to that of
+2-D blade-resolved CFD. The cost is a function of the number of vortex elements,
+which for free vortex methods may increase at each time step, slowing down the
+calculation as it marches forward in time, unless there is an additional model
+for vortex diffusion. Potential flow vortex methods can resolve more flow
+details than BEM. However, since the flow is governed by the Laplace equation,
+nonlinear effects such as the turbulent transport will not be included, which we
+have determined in experiments to be the same order of magnitude as the mean
+vertical advection.
 

### 2016-05-02 paper.tex
Add pitching moment formula; resolves #17

@@
 
-Assuming unsteady effects can be appropriately modeled, the blade element lift
-and drag forces are calculated as
+Assuming unsteady effects can be appropriately modeled, the blade element lift force, drag forces, and pitching moment are calculated as
 \begin{equation}
@@
 \end{equation}
-and
 \begin{equation}
@@
 \end{equation}
-where $\rho$ is the fluid density, $A_\mathrm{elem}$ is the blade element
-planform area (span $\times$ chord), $\vec{U}_\mathrm{rel}$ is the local
+\begin{equation}
+    M = \frac{1}{2} \rho A_\mathrm{elem} c C_m |\vec{U}_\mathrm{rel}|^2,
+\end{equation}
+respectively, where $\rho$ is the fluid density, $A_\mathrm{elem}$ is the blade
+element planform area (span $\times$ chord), $\vec{U}_\mathrm{rel}$ is the local
 relative velocity projected onto the plane of the element profile cross-section
-(i.e., the spanwise component is neglected), and $C_l$ and $C_d$ are the
-sectional lift and drag coefficients, linearly interpolated from a table per the
-local angle of attack. The forces are then projected onto the rotor coordinate
-system to calculate torque, overall drag, etc. Forces from the turbine shaft and
-blade support struts are computed in a similar way. After the force on the
-actuator lines from the flow is computed, it is then added to the Navier--Stokes
-equations as a body force or momentum source (per unit density, assuming
-incompressible flow):
+(i.e., the spanwise component is neglected), and $C_l$, $C_d$, and $C_m$ are the
+sectional lift, drag, and pitching moment coefficients, respectively, which are
+linearly interpolated from a table per the local angle of attack. The forces are
+then projected onto the rotor coordinate system to calculate torque, overall
+drag, etc. Forces from the turbine shaft and blade support struts are computed
+in a similar way. After the force on the actuator lines from the flow is
+computed, it is then added to the Navier--Stokes equations as a body force or
+momentum source (per unit density, assuming incompressible flow):
 \begin{equation}

### 2016-05-02 paper.tex
Is --> was
Write abstract

@@
 \begin{abstract}
-    % TODO add abstract here
+    To bridge the gap between high and low fidelity numerical modeling tools for
+    vertical-axis (or cross-flow) turbines (VATs or CFTs), an actuator line
+    model (ALM) was developed and validated for both a high and a medium
+    solidity vertical-axis turbine at turbine diameter Reynolds numbers $Re_D
+    \sim 10^6$. The ALM is a hybridization of classical blade element theory
+    with Navier--Stokes based flow models, and in this study both
+    $k$--$\epsilon$ Reynolds-averaged Navier--Stokes (RANS) and Smagorinsky
+    large eddy simulation (LES) turbulence models were tested. The RANS models
+    were able to be run on coarse grids while still providing good convergence
+    behavior in terms of the mean power coefficient, and also approximately four
+    orders of magnitude reduction in computational expense compared with 3-D
+    blade-resolved RANS simulations. Submodels for dynamic stall, end effects,
+    added mass, and flow curvature were implemented, resulting in reasonable
+    performance predictions for the high solidity rotor, more discrepancies for
+    the medium solidity rotor, and overprediction for both cases at high tip
+    speed ratio. The wake results showed that the ALM was able to capture some
+    of the important flow features that contribute to VAT's relatively fast wake
+    recovery---a large improvement over the conventional actuator disk model.
+    The mean flow field was better realized with the LES, which still
+    represented a computational savings of two orders of magnitude compared with
+    3-D blade-resolved RANS, though vortex breakdown and subsequent turbulence
+    generation appeared to be underpredicted, which necessitates further
+    investigation of optimal subgrid scale modeling.
 \end{abstract}
@@
 quarter-chord location). A sampling line was chosen to be 5 times the local mesh
-cell length. Finally, an angle of attack correction is proposed
+cell length. Finally, an angle of attack correction was proposed
 \begin{equation}

### 2016-05-02 paper.tex
Add address for Goude

@@
 
-\author{P.~Bachant, A.~Goude, and M.~Wosnik}
-
-\address{Center for Ocean Renewable Energy, University of New Hampshire, 24
-Colovos Rd., Durham, NH, 03824, USA}
+\author{P.~Bachant\textsuperscript{1}, A.~Goude\textsuperscript{2}, and
+M.~Wosnik\textsuperscript{1}}
+
+\address{\textsuperscript{1}Center for Ocean Renewable Energy,
+University of New Hampshire, 24
+Colovos Rd., Durham, NH, 03824, USA \\
+\textsuperscript{2}Department of Engineering Sciences,
+Division of Electricity, Uppsala University, Uppsala 751 21, Sweden}
 

### 2016-05-04 paper.tex
Add blade chord line

@@
         bottom (identical to $U_in$ for purely geometric calculations), the blade
-    chord is in line with the tangential velocity (i.e., zero preset pitch,
-    which would offset the geometric angle of attack $\alpha$), and the drag
-    vector is magnified by a factor of two (approximately, relative to the lift
-    vector) to enhance visibility.}
+        chord (dash-dotted line) is coincident with the tangential velocity (i.e.,
+        zero preset pitch, which would offset the geometric angle of attack
+        $\alpha$), and the drag vector is magnified by a factor of two
+        (approximately, relative to the lift vector) to enhance visibility.}
 

### 2016-05-04 paper.tex
Make Bachant corresponding address
Add more BR CFD refs

@@
 
-\corraddr{M.~Wosnik, Chase Ocean Engineering Laboratory, University of New
+\corraddr{P.~Bachant, Chase Ocean Engineering Laboratory, University of New
 Hampshire, 24 Colovos Rd., Durham, NH, 03824, USA. \\
-E-mail: martin.wosnik@unh.edu}
+E-mail: pxL3@unh.edu}
 
@@
 performance and near-wake structure of a VAT, though their effectiveness depends
-on the turbulence model applied~\cite{Bachant2016-BR-CFD}. However, 3-D
+on the turbulence model
+applied~\cite{Bachant2016-BR-CFD,Lam2016,Alaimo2015,Boudreau2015}. However, 3-D
 blade-resolved RANS presents a huge computational expense---on the order of

### 2016-05-04 paper.tex
Do not speculate as to why DMST models do poorly for high solidity; resolves #18

@@
 models, such as the double multiple streamtube (DMST) approached developed by
-Paraschivoiu~\cite{Para1988}. However, their poor performance for high loading
-or streamwise induction factors, common for rotors with high
-solidity~\cite{Joo2015} $Nc/R$ (and/or chord-to-radius ratio $c/R$), where $N$
-is the number of blades, $c$ is the chord length, and $R$ is the maximum rotor
-radius, makes them less attractive, particular for marine hydrokinetic
+Paraschivoiu~\cite{Para1988}. However, their poor performance for rotors with
+high solidity~\cite{Joo2015} $Nc/R$ (and/or chord-to-radius ratio $c/R$), where
+$N$ is the number of blades, $c$ is the chord length, and $R$ is the maximum
+rotor radius, makes them less attractive, particular for marine hydrokinetic
 applications where rotors of the same size will encounter approximately an order
@@
     \caption{Vector diagram of velocity and forcing on a cross-flow turbine
-        blade element. Note that the free stream velocity $U_\infty$ is from top to
-        bottom (identical to $U_in$ for purely geometric calculations), the blade
-        chord (dash-dotted line) is coincident with the tangential velocity (i.e.,
-        zero preset pitch, which would offset the geometric angle of attack
-        $\alpha$), and the drag vector is magnified by a factor of two
+        blade element. Note that the free stream velocity $U_\infty$ is oriented
+        from top to bottom (identical to $U_in$ for purely geometric calculations),
+        the blade chord (dash-dotted line) is coincident with the tangential
+        velocity (i.e., zero preset pitch, which would offset the geometric angle of
+        attack $\alpha$), and the drag vector is magnified by a factor of two
         (approximately, relative to the lift vector) to enhance visibility.}

### 2016-05-04 paper.tex
Define NREL earlier

@@
 blade element's location in their actuator line simulation of a vertical-axis
-turbine using LES \cite{Shamsoddin2014}. NREL's SOWFA ALM for OpenFOAM also uses
-the velocity at the actuator line element location for computing inflow velocity
-and local angle of attack with no corrections \cite{Churchfield2013}.
+turbine using LES \cite{Shamsoddin2014}. The US National Renewable Energy
+Laboratory (NREL) SOWFA ALM for OpenFOAM also uses the velocity at the actuator
+line element location for computing inflow velocity and local angle of attack
+with no corrections \cite{Churchfield2013}.
 
@@
 
-The US National Renewable Energy Laboratory (NREL) has developed and released an
-actuator line modeling library, SOWFA~\cite{Churchfield2014b}, for simulating
-horizontal-axis wind turbine arrays using the OpenFOAM finite volume CFD
-library. OpenFOAM is free, open-source, widely used throughout industry and
-academia, and has grown a very active support community around itself
-\cite{Bachant2015-CFD-online}. Though SOWFA is also open-source, and has the
-ability to be coupled to NREL's FAST HAWT BEM code, its mostly procedural style
-would have required significant effort and duplicate code to adapt for
-cross-flow turbines. Thus, a new and more general ALM library was developed from
-the ground up that could model both cross- and axial-flow turbines, as well as
-standalone actuator lines. The actuator line model developed here, dubbed
-turbinesFoam \cite{Bachant2016-turbinesFoam-v0.0.7}, was also written as an
-extension library for OpenFOAM, and was developed freely and openly from its
-inception in order to increase community engagement and research efficiency.
+NREL has developed and released an actuator line modeling library,
+SOWFA~\cite{Churchfield2014b}, for simulating horizontal-axis wind turbine
+arrays using the OpenFOAM finite volume CFD library. OpenFOAM is free,
+open-source, widely used throughout industry and academia, and has grown a very
+active support community around itself \cite{Bachant2015-CFD-online}. Though
+SOWFA is also open-source, and has the ability to be coupled to NREL's FAST HAWT
+BEM code, its mostly procedural style would have required significant effort and
+duplicate code to adapt for cross-flow turbines. Thus, a new and more general
+ALM library was developed from the ground up that could model both cross- and
+axial-flow turbines, as well as standalone actuator lines. The actuator line
+model developed here, dubbed turbinesFoam
+\cite{Bachant2016-turbinesFoam-v0.0.7}, was also written as an extension library
+for OpenFOAM, and was developed freely and openly from its inception in order to
+increase community engagement and research efficiency.
 

### 2016-05-04 paper.tex
Add reference to RVAT baseline dataset
Add more ref to experiments, and always call RM2 medium solidity

@@
 
-Both the high solidity UNH-RVAT and medium-low solidity RM2 turbines were
-modeled using the ALM inside a Reynolds-averaged Navier--Stokes (RANS)
-simulation, closed with the standard $k$--$\epsilon$ turbulence model. These
-rotors provide diverse parameters, which helped evaluate the robustness of the
-ALM. The simulations were performed inside a domain similar in size to that used
-in \cite{Bachant2016-BR-CFD} to mimic the tow tank experiments (3.66 m wide,
-2.44 m tall, 1.52 m upstream and 2.16 m downstream), with similar velocity
-boundary conditions: 1 m/s inflow, no-slip (fixed to 1 m/s) walls and bottom,
-and a rigid slip condition for the top, to approximate some effects of the tow
-tank's free surface. Simulations were run for a total of 6 seconds, with the
-latter half used to calculate performance and wake statistics.
-Pressure--velocity coupling for the momentum equation was achieved using the
-PISO (pressure implicit splitting of operators) method. A slice of the mesh in
-the $x$--$y$ plane is shown in Figure~\ref{fig:ALM-mesh}.
+Both the high solidity UNH-RVAT and medium solidity RM2 turbines were modeled
+using the ALM inside a Reynolds-averaged Navier--Stokes (RANS) simulation,
+closed with the standard $k$--$\epsilon$ turbulence model. These rotors provide
+diverse parameters, which helped evaluate the robustness of the ALM. The
+simulations were performed inside a domain similar in size to that used in
+\cite{Bachant2016-BR-CFD} to mimic the tow tank experiments (3.66 m wide, 2.44 m
+tall, 1.52 m upstream and 2.16 m downstream), with similar velocity boundary
+conditions: 1 m/s inflow, no-slip (fixed to 1 m/s) walls and bottom, and a rigid
+slip condition for the top, to approximate some effects of the tow tank's free
+surface. Simulations were run for a total of 6 seconds, with the latter half
+used to calculate performance and wake statistics. Pressure--velocity coupling
+for the momentum equation was achieved using the PISO (pressure implicit
+splitting of operators) method. A slice of the mesh in the $x$--$y$ plane is
+shown in Figure~\ref{fig:ALM-mesh}.
 
@@
     \caption{Power and drag coefficient curves computed for the UNH-RVAT using
-        the actuator line model with RANS.}
+        the actuator line model with RANS, compared with experimental results
+        from~\cite{Bachant2016-RVAT-Re-dep}.}
 
@@
 Figure~\ref{fig:RVAT-ALM-meancontquiv} shows mean velocity field for the
-UNH-RVAT computed by the ALM RANS model. The asymmetry was captured well, along
-with some of the vertical flow due to blade tip vortex shedding, though the flow
-structure is missing the detail present in the experiments and blade-resolved
-RANS simulations. Overall, the wake appears to be over-diffused, which could be
-a consequence of the relatively coarse mesh. Note that with the DS and flow
+UNH-RVAT computed by the ALM RANS model. The asymmetry observed in the
+experiments \cite{Bachant2015-JoT} was captured well, along with some of the
+vertical flow due to blade tip vortex shedding, though the flow structure is
+missing the detail present in the experiments and blade-resolved RANS
+simulations. Overall, the wake appears to be over-diffused, which could be a
+consequence of the relatively coarse mesh. Note that with the DS and flow
 curvature corrections turned off, the direction of the mean swirling motion
@@
 shown in Figure~\ref{fig:RVAT-ALM-kcont}. The ALM was able to resolve the
-concentrated area of $k$ on the $+y$ side of the turbine, but the turbulence
-generated by the dynamic stall vortex shedding process is absent. This makes
-sense since in the ALM, the DS model only modulates the body force term in the
-momentum equation, which does not provide a mechanism for mimicking shed
-vortices or turbulence.
+concentrated area of $k$ on the $+y$ side of the turbine, present in the
+experiments~\cite{Bachant2015-JoT}, but the turbulence generated by the dynamic
+stall vortex shedding process is absent. This makes sense since in the ALM, the
+DS model only modulates the body force term in the momentum equation, which does
+not provide a mechanism for mimicking shed vortices or turbulence.
 
@@
     \caption{Mean streamwise velocity (left) and turbulence kinetic energy
-        (right) profiles at $z/H=0$ for the UNH-RVAT ALM.}
+        (right) profiles at $z/H=0$ for the UNH-RVAT ALM, compared with the
+        experimental data from~\cite{Bachant2016-RVAT-Re-dep}.}
 
@@
 Figure~\ref{fig:RM2-ALM-perf-curves} shows the performance curves computed for
-the RM2 by the ALM, and those from the tow tank experiments. As with the high
-solidity RVAT, $C_P$ is overpredicted at high $\lambda$. However, $\lambda_0$,
-the tip speed ratio of peak power coefficient, is also shifted to the right.
-This is indicative of inaccurate dynamic stall modeling, which could possibly be
-attributed to one of the models' tuning constants, e.g., the time constant for
-the lagged angle of attack. Limited ad hoc testing revealed that $C_P$ at
-$\lambda_0$ was more accurately predicted with a lagged angle of attack time
-constant roughly double the default value. This could be investigated further by
-looking at the phase of, e.g., maximum $C_P$ peaks compared to the experimental
-values (or possibly those from blade-resolved CFD), but for the present study
-only mean values were considered.
+the RM2 by the ALM, and those from the tow tank
+experiments~\cite{Bachant2016-RM2-paper}. As with the high solidity RVAT, $C_P$
+is overpredicted at high $\lambda$. However, $\lambda_0$, the tip speed ratio of
+peak power coefficient, is also shifted to the right. This is indicative of
+inaccurate dynamic stall modeling, which could possibly be attributed to one of
+the models' tuning constants, e.g., the time constant for the lagged angle of
+attack. Limited ad hoc testing revealed that $C_P$ at $\lambda_0$ was more
+accurately predicted with a lagged angle of attack time constant roughly double
+the default value. This could be investigated further by looking at the phase
+of, e.g., maximum $C_P$ peaks compared to the experimental values (or possibly
+those from blade-resolved CFD), but for the present study only mean values were
+considered.
 
@@
     \caption{Power and drag coefficient curves computed for the RM2 using the
-        ALM.}
+        ALM, compared with experimental data from~\cite{Bachant2016-RM2-data}.}
 
@@
 structure looks similar to the RVAT ALM RANS case, but for the RM2, the lack of
-detail from blade tip vortex shedding matches more closely with experiments.
-However, the vertical flow towards the $x$--$y$ center plane was captured, which
-is an important qualitative feature of both CFT near-wakes.
+detail from blade tip vortex shedding matches more closely with
+experiments~\cite{Bachant2016-RM2-paper}. However, the vertical flow towards the
+$x$--$y$ center plane was captured, which is an important qualitative feature of
+both CFT near-wakes.
 
@@
         (right) profiles at $x/D=0.93$ and $z/H=0$ computed for the RM2 using the
-    ALM inside a $k$--$\epsilon$ RANS model.}
+        ALM inside a $k$--$\epsilon$ RANS model, compared with experimental data
+        from~\cite{Bachant2016-RM2-data}.}
 
@@
 Figure~\ref{RVAT-ALM-LES-vorticity}. Compared with the RANS ALM results, the LES
-looks much more like the blade-resolved and experimental results, showing the
-clockwise and counterclockwise mean swirling motion on the $-y$ and $+y$ sides
-of the rotor, respectively.
+looks much more like the blade-resolved and experimental results
+\cite{Bachant2015-JoT}, showing the clockwise and counterclockwise mean swirling
+motion on the $-y$ and $+y$ sides of the rotor, respectively.
 
@@
     \caption{Mean velocity profiles in the UNH-RVAT near-wake at $x/D=1$ and
-        $z/H=0$ computed with the Smagorinsky LES model.}
+        $z/H=0$ computed with the Smagorinsky LES model, compared with experimental
+        data from~\cite{Bachant2016-RVAT-Re-dep}.}
 
@@
 effects corrections, was developed and validated against experimental datasets
-acquired for high and medium-low solidity rotors at scales where the performance
-and near-wake dynamics were essentially Reynolds number independent. When
-coupled to a $k$--$\epsilon$ RANS solver ALM simulations took $O(0.1)$ CPU hours
-per second of simulated time, while when coupled with a Smagorinsky LES model
-the computing time was $O(10)$ hours per second, which represent a four and two
-order of magnitude decrease in computational expense versus 3-D blade-resolved
-RANS, respectively.
+acquired for high and medium solidity rotors at scales where the performance and
+near-wake dynamics were essentially Reynolds number independent. When coupled to
+a $k$--$\epsilon$ RANS solver ALM simulations took $O(0.1)$ CPU hours per second
+of simulated time, while when coupled with a Smagorinsky LES model the computing
+time was $O(10)$ hours per second, which represent a four and two order of
+magnitude decrease in computational expense versus 3-D blade-resolved RANS,
+respectively.
 

### 2016-05-04 paper.tex
Turbine --> rotor

@@
     model (ALM) was developed and validated for both a high and a medium
-    solidity vertical-axis turbine at turbine diameter Reynolds numbers $Re_D
-    \sim 10^6$. The ALM is a hybridization of classical blade element theory
-    with Navier--Stokes based flow models, and in this study both
-    $k$--$\epsilon$ Reynolds-averaged Navier--Stokes (RANS) and Smagorinsky
-    large eddy simulation (LES) turbulence models were tested. The RANS models
-    were able to be run on coarse grids while still providing good convergence
-    behavior in terms of the mean power coefficient, and also approximately four
-    orders of magnitude reduction in computational expense compared with 3-D
+    solidity vertical-axis turbine at rotor diameter Reynolds numbers $Re_D \sim
+    10^6$. The ALM is a hybridization of classical blade element theory with
+    Navier--Stokes based flow models, and in this study both $k$--$\epsilon$
+    Reynolds-averaged Navier--Stokes (RANS) and Smagorinsky large eddy
+    simulation (LES) turbulence models were tested. The RANS models were able to
+    be run on coarse grids while still providing good convergence behavior in
+    terms of the mean power coefficient, and also approximately four orders of
+    magnitude reduction in computational expense compared with 3-D
     blade-resolved RANS simulations. Submodels for dynamic stall, end effects,

### 2016-05-04 paper.tex
Add BR CFD ref

@@
 on the turbulence model
-applied~\cite{Bachant2016-BR-CFD,Lam2016,Alaimo2015,Boudreau2015}. However, 3-D
-blade-resolved RANS presents a huge computational expense---on the order of
-1,000 CPU hours per simulated second with contemporary hardware---since it must
-resolve fine details of the blade boundary layers, which will preclude its use
-for array analysis until the availability of computing power increases
+applied~\cite{Bachant2016-BR-CFD,Lam2016,Alaimo2015,Boudreau2015,Marsh2015}.
+However, 3-D blade-resolved RANS presents a huge computational expense---on the
+order of 1,000 CPU hours per simulated second with contemporary hardware---since
+it must resolve fine details of the blade boundary layers, which will preclude
+its use for array analysis until the availability of computing power increases
 sufficiently. It is therefore necessary to explore simpler models that can

### 2016-05-04 scripts/zip.py
Add class and bst files to zip

@@ -11,7 +11,7 @@ version = check_output(["git", "describe"]).strip().decode()
 figures = ["figures/" + f for f in os.listdir("figures") \
            if f.endswith(".pdf") or f.endswith(".png")]
 
-files = ["paper.tex", "paper.bbl"]
+files = ["paper.tex", "paper.bbl", "weauth.cls", "wileyj.bst"]
 
 with ZipFile("archive/paper-{}.zip".format(version), "w") as f:
     for fig in figures:

### 2016-05-05 paper.tex
PhD --> doctoral
Fix typo forces --> force

@@
 
-Assuming unsteady effects can be appropriately modeled, the blade element lift force, drag forces, and pitching moment are calculated as
+Assuming unsteady effects can be appropriately modeled, the blade element lift
+force, drag force, and pitching moment are calculated as
 \begin{equation}
@@
 Energy for Sustainability, program manager Gregory L. Rorrer). This work has
-been carried out as part of author P. Bachant's PhD research at the University
-of New Hampshire.
+been carried out as part of author P. Bachant's doctoral research at the
+University of New Hampshire.
 

### 2016-05-12 paper.tex
Fix U_in in figure caption

@@
     blade element. Note that the free stream velocity $U_\infty$ is oriented
-        from top to bottom (identical to $U_in$ for purely geometric calculations),
-        the blade chord (dash-dotted line) is coincident with the tangential
-        velocity (i.e., zero preset pitch, which would offset the geometric angle of
-        attack $\alpha$), and the drag vector is magnified by a factor of two
-        (approximately, relative to the lift vector) to enhance visibility.}
+    from top to bottom (identical to $U_{\mathrm{in}}$ for purely geometric
+    calculations), the blade chord (dash-dotted line) is coincident with the
+    tangential velocity (i.e., zero preset pitch, which would offset the
+    geometric angle of attack $\alpha$), and the drag vector is magnified by a
+    factor of two (approximately, relative to the lift vector) to enhance
+    visibility.}
 

### 2016-05-12 paper.tex
Slightly tweak wording for DS model speculation

@@
 the models' tuning constants, e.g., the time constant for the lagged angle of
-attack. Limited ad hoc testing revealed that $C_P$ at $\lambda_0$ was more
-accurately predicted with a lagged angle of attack time constant roughly double
-the default value. This could be investigated further by looking at the phase
-of, e.g., maximum $C_P$ peaks compared to the experimental values (or possibly
-those from blade-resolved CFD), but for the present study only mean values were
-considered.
+attack. Limited ad hoc testing revealed that the mean $C_P$ at $\lambda_0$ more
+closely matched experimental measurements with $T_\alpha$ roughly double the
+default value given in~\cite{Sheng2008}. This could be investigated further by
+looking at the phase of, e.g., maximum $C_P$ peaks compared to the experimental
+values (or possibly those from blade-resolved CFD), but for the present study
+only mean values were considered.
 

### 2016-05-12 paper.tex
Define T_alpha earlier

@@
 inaccurate dynamic stall modeling, which could possibly be attributed to one of
-the models' tuning constants, e.g., the time constant for the lagged angle of
-attack. Limited ad hoc testing revealed that the mean $C_P$ at $\lambda_0$ more
-closely matched experimental measurements with $T_\alpha$ roughly double the
-default value given in~\cite{Sheng2008}. This could be investigated further by
-looking at the phase of, e.g., maximum $C_P$ peaks compared to the experimental
-values (or possibly those from blade-resolved CFD), but for the present study
-only mean values were considered.
+the models' tuning constants, e.g., the time constant $T_\alpha$ for the lagged
+angle of attack. Limited ad hoc testing revealed that the mean $C_P$ at
+$\lambda_0$ more closely matched experimental measurements with $T_\alpha$
+roughly double the default value given in~\cite{Sheng2008}. This could be
+investigated further by looking at the phase of, e.g., maximum $C_P$ peaks
+compared to the experimental values (or possibly those from blade-resolved CFD),
+but for the present study only mean values were considered.
 

### 2016-05-12 paper.tex
Discuss wake transport ratios; resolves #20
Update corresponding email; resolves #22
Note that this is a preprint; resolves #23
Remove 'hybridization' from abstract

@@
 
-\articletype{RESEARCH ARTICLE}
+\articletype{RESEARCH ARTICLE PREPRINT (submitted May 2016)}
 
@@
 Hampshire, 24 Colovos Rd., Durham, NH, 03824, USA. \\
-E-mail: pxL3@unh.edu}
+E-mail: pwbachant@gmail.com}
 
@@
     solidity vertical-axis turbine at rotor diameter Reynolds numbers $Re_D \sim
-    10^6$. The ALM is a hybridization of classical blade element theory with
+    10^6$. The ALM is a combination of classical blade element theory and
     Navier--Stokes based flow models, and in this study both $k$--$\epsilon$
@@
 negative, vertical advection was predicted reasonably well, and turbulent
-transport due to the eddy viscosity was also relatively large.
+transport due to the eddy viscosity was also relatively large. The ratio of wake
+transport compared with the UNH-RVAT RANS case (approximately 60\% to 70\% lower
+for the RM2) matches well with that computed from the experiments, which shows
+the ALM may successfully predict larger optimal array spacing for the RM2 versus
+UNH-RVAT.
 

### 2016-05-12 paper.tex
Remove a lot of the inflow velocity detection discussion

@@
 
-In momentum methods, the inflow velocity is determined by solving for the axial
-and angular induction factors \cite{Manwell2002}. However, using Navier--Stokes
-methods, it is somewhat unclear how to calculate the velocity vector used to
-compute the angle of attack and relative velocity, though we have access to much
-more information about the flow. Sorensen and Shen used an actuator line
-element's position to determine the inflow velocity for an axial-flow turbine
-\cite{Sorensen2002}. Similarly, Shamsoddin and Porte-Agel use the velocity at a
-blade element's location in their actuator line simulation of a vertical-axis
-turbine using LES \cite{Shamsoddin2014}. The US National Renewable Energy
-Laboratory (NREL) SOWFA ALM for OpenFOAM also uses the velocity at the actuator
-line element location for computing inflow velocity and local angle of attack
-with no corrections \cite{Churchfield2013}.
-
-Schito and Zasso developed an effective velocity model (EVM) for computing
-actuator forces in Navier--Stokes simulations \cite{Schito2014}. Their EVM
-proposes that inflow velocity should be sampled along a line perpendicular to
-the mean relative flow direction. They ultimately determined that the sampling
-line should be 1.5 chord lengths upstream of the actuator point (i.e.
-quarter-chord location). A sampling line was chosen to be 5 times the local mesh
-cell length. Finally, an angle of attack correction was proposed
-\begin{equation}
-    \Delta \alpha = \frac{c}{M} (1.2553 - 0.0552 C_d) C_l,
-    \label{eq:EVM-dalpha}
-\end{equation}
-where $c$ is the chord length, $M$ is the mesh size, and $C_l$ and $C_d$ are the
-lift and drag coefficients, respectively. Note that the constants in
-Equation~\ref{eq:EVM-dalpha} were determined from a calibration with 2-D
-blade-resolved CFD for a NACA 0012 foil and are not assumed to be universal.
-
-The EVM, despite showing robustness for its chosen validation case,
-unfortunately involves determination of two unknown tuning parameters. To avoid
-the additional effort and uncertainty in determining these, the inflow velocity
-was sampled at the element quarter-chord location using OpenFOAM's
-\texttt{interpolationCellPoint} class, which provides a linear weighted
-interpolation using cell values. This algorithm helps keep the sampled velocity
-``smooth'' compared with using the cell values themselves, especially when
-elements are moving in space as they are in a turbine, since meshes will likely
-have a cell size on the same order as the chord length, and will move on the
-order of one cell length per time step.
+Inflow velocity was sampled for each actuator line element at its quarter-chord
+location using OpenFOAM's \texttt{interpolationCellPoint} class, which provides
+a linear weighted interpolation using cell values. This algorithm helps keep the
+sampled velocity ``smooth'' compared with using the cell values themselves,
+especially when elements are moving in space as they are in a turbine, since
+meshes will likely have a cell size on the same order as the chord length, and
+will move on the order of one cell length per time step.
 

### 2016-05-12 paper.tex
Remove a bunch of DS model background info

@@
 
-DS models were first developed to improve predictive capability for helicopter
-rotors, on which DS has significant effects on maneuverability and operational
-envelope \cite{Bousman2000}. A summary of dynamic stall models developed for
-helicopter rotors is presented in \cite{Leishman2006}. The simplest dynamic
-stall models rely on semi-empirical correlations, e.g., the Gormont model
-\cite{Gormont1973}, developed at the Boeing--Vertol Company. Several variants of
-the Gormont model were developed for vertical-axis wind turbines, with varying
-degrees of success; a summary is presented in \cite{Para2002}.
-
-Leishman and Beddoes (LB) \cite{Leishman1989} developed a semi-empirical model
-for unsteady aerodynamics and dynamic stall, which is derived from the
-phenomenology of the physics instead rather than pure empiricism. Beddoes then
-updated the model to the so-called third generation or ``3G'' version
-\cite{Beddoes1993}. The LB DS models can be summarized conceptually based on the
-following principles:
-\begin{itemize}
-    \item Dynamic conditions cause a time lag in effective angle of attack and
-    lift force.
-
-    \item Separation is determined by the Kirchoff flow approximation, which is
-    also used to parameterize the normal force coefficient table based on the
-    trailing edge separation point. This separation point also encounters a time
-    lag.
-
-    \item The separation initiates a vortex shedding cycle that causes an
-    overshoot and subsequent undershoot in lift before returning to an attached
-    flow condition.
-\end{itemize}
-
-Sheng et al. \cite{Sheng2008} developed an LB DS model variant targeted at low
-Mach numbers. This model, along with the original and 3G LB DS model variants,
-was tested for its effectiveness in cross-flow turbine conditions by Dyachuk et
-al.~\cite{Dyachuk2014}, who concluded that the Sheng et al. variant results
-matched most closely with experiments. In a similar study, the Sheng et al.
-model also faired better than the Gormont model \cite{Dyachuk2015}, which
-inspired its adoption here for the ALM.
+In this study we employed a dynamic stall model developed for low mach numbers
+by Sheng et al. \cite{Sheng2008}---derived from the Leishman--Beddoes
+semi-empirical model \cite{Leishman1989}. This model, along with two other
+Leishman--Beddoes model variants, was tested for its effectiveness in cross-flow
+turbine conditions by Dyachuk et al.~\cite{Dyachuk2014}, who concluded that the
+Sheng et al. variant results matched most closely with experiments. In a similar
+study \cite{Dyachuk2015}, the Sheng et al. model also performed better than the
+Gormont model \cite{Gormont1973}, which inspired its adoption here for the ALM.
 

### 2016-05-12 paper.tex
Remove unused Mandal--Burton fc model details

@@
 Furthermore, this effect is more pronounced for high solidity ($c/R$) turbines.
-Two different flow curvature corrections were considered: one derived in Goude
-\cite{Goude2012} and one by Mandal and Burton \cite{Mandal1994}.
-
-The Goude correction is derived by considering a flat plat moving along a
-circular path in potential flow, for which the effective angle of attack
-including flow curvature effects is given by
+
+The flow curvature correction used here was derived in Goude \cite{Goude2012} by
+considering a flat plat moving along a circular path in potential flow, for
+which the effective angle of attack including flow curvature effects is given by
 \begin{equation}
@@
 negative.
-
-The Mandal--Burton flow curvature correction assumes that since the blade is
-encountering a curvilinear flow, it can be treated as having virtual camber.
-They introduce a factor to describe the variation of angle of attack from the
-leading to trailing edge
-\begin{equation}
-    \Delta \alpha = \alpha_\mathrm{TE} - \alpha_\mathrm{LE},
-    \label{eq:Mandal-Burton-alpha-diff}
-\end{equation}
-where TE and LE subscripts denote the values of angle of attack at the trailing
-and leading edge, respectively. Calculating these values for an actuator line
-element can be done by tracking the leading and trailing edge locations and
-velocities, then performing the same vector arithmetic used to calculate the
-quarter-chord angle of attack.
-
-An incidence correction factor
-\begin{equation}
-    \alpha_c = \arctan \left( \frac{1 - \cos (\Delta \alpha / 2)}{\sin (\Delta
-        \alpha / 2)} \right)
-    \label{eq:Mandal-Burton-alpha-corr}
-\end{equation}
-is introduced and added to the uncorrected angle of attack. Like the Goude
-model, $\alpha_c$ is positive on the upstream half of the turbine rotation and
-negative on the downstream half.
-
-It was determined that the Goude model performed best with respect to matching
-the near-wake characteristics measured in the experiments, while the performance
-predictions from both flow curvature models were relatively close to each other.
-Thus, the Goude model was used for all simulations reported here.
 

### 2016-05-12 paper.tex
Remove some unnecessary software details

@@
 in both axial-flow and cross-flow rotors, without having to copy code from one
-to the other. In contrast, the actuator line model in SOWFA uses a single
-software object to represent an entire array of turbines, which necessitates
-iterating through many nested lists down to the element level, which can be
-confusing to read.
-
-OpenFOAM's data structures are designed to be inherently parallel via message
-passing interface (MPI). By working within the library infrastructure the ALM
-code was easily parallelized, which will facilitate its deployment on high
-performance computing clusters for large flow simulations.
-
-Since all applications are run from a command line and all input data is text
-based, automation and integration with other tools is relatively
-straightforward. Future enhancements could include coupling with software for
-generating static foil data, e.g., XFOIL~\cite{Drela1989} or other OpenFOAM
-solvers, turbine controller models, structural analysis codes, and optimization
-tools, e.g., SNL's DAKOTA \cite{AdamsBaumanBohnhoffEtAl2009} or
-OpenMDAO~\cite{Gray2010}, for both individual turbines and array layouts.
+to the other.
 

### 2016-05-12 paper.tex
Remove emph on et al

@@
 A correction for added mass effects, or the effects due to accelerating the
-fluid, was taken from Strickland et al.~\cite{Strickland1981}, which was
-derived by considering a pitching flat plate in potential flow. In the blade
-element coordinate system, the normal and chordwise (pointing from trailing to
-leading edge, which is opposite the $x$-direction used by Strickland \emph{et
-    al.}) coefficients due to added mass are
+fluid, was taken from Strickland et al.~\cite{Strickland1981}, which was derived
+by considering a pitching flat plate in potential flow. In the blade element
+coordinate system, the normal and chordwise (pointing from trailing to leading
+edge, which is opposite the $x$-direction used by Strickland et al.)
+coefficients due to added mass are
 \begin{equation}

### 2016-05-12 paper.tex
Remove formulas for Glauert tip loss corrections

@@
 to develop a tip loss correction factor $F$ to be applied to the local lift of
-an axial-flow rotor blade element:
-\begin{equation}
-    F = \frac{2}{\pi} \cos^{-1} \left[ \exp \left( - \frac{N (R-r)}{2R \sin
-        \phi_R} \right) \right],
-\end{equation}
-where $F$ is the correction factor applied to the local lift, $N$ is number of
-blades, $R$ is the rotor radius, $r$ is the local blade element radius, and
-$\phi_R$ is the flow angle at the tip.
-
-The Glauert correction was further refined for horizontal-axis wind turbines by
-Shen et al. \cite{Shen2005a} as
-\begin{equation}
-    F_1 = \frac{2}{\pi} \cos^{-1} \left[ \exp \left( -g \frac{N (R-r)}{2R \sin
-    \phi_R} \right) \right],
-\end{equation}
-where $g$ is an additional function depending on the tip speed ratio and two
-tuning constants:
-\begin{equation}
-    g = \exp [ -c_1 (B \lambda - c_2) ].
-\end{equation}
-
-Despite their success for the blade element analysis of axial-flow rotors, these
-corrections both depend on rotor parameters---tip speed ratio, number of blades,
-element radius, tip flow angle---that do no necessarily translate directly to
-the geometry and flow environment of a cross-flow rotor. Therefore, a more
-general end effects model was sought.
+an axial-flow rotor blade element. The Glauert correction was further refined
+for horizontal-axis wind turbines by Shen et al. \cite{Shen2005a}. Despite their
+success for the blade element analysis of axial-flow rotors, these corrections
+both depend on rotor parameters---tip speed ratio, number of blades, element
+radius, tip flow angle---that do no necessarily translate directly to the
+geometry and flow environment of a cross-flow rotor. Therefore, a more general
+end effects model was sought.
 

### 2016-05-12 paper.tex
Remove code listing

@@
 turbinesFoam was written in OpenFOAM's style, using OpenFOAM's
-\texttt{fvOptions} framework for adding source terms to equations at
-runtime---see Listing~\ref{lst:fvOptions} for an example implementation within
-the Navier--Stokes' momentum equation. Using the \texttt{fvOptions} framework
-allows the CFT-ALM to be added to many of the solvers included in OpenFOAM,
-meaning it can be readily used with RANS or LES, multiphase models (e.g., for
-simulating the free surface in MHK installations), and even with heat transfer.
-This is in contrast to SOWFA's implementation, which requires custom flow
-solvers to be developed and maintained.
-
-\begin{lstlisting}[float,language=C++,caption=Adding source terms to the momentum equation in OpenFOAM.,label=lst:fvOptions]
-    tmp<fvVectorMatrix> UEqn
-    (
-        fvm::ddt(U)
-        + fvm::div(phi, U)
-        + turbulence->divDevReff(U)
-        ==
-        fvOptions(U)
-    );
-\end{lstlisting}
+\texttt{fvOptions} framework for adding source terms to equations at runtime.
+This implementation allows the CFT-ALM to be added to many of the standard
+solvers included in OpenFOAM without modification, i.e., it can be readily used
+with RANS or LES, multiphase models (e.g., for simulating the free surface in
+MHK installations), and even with heat transfer.
 

### 2016-05-12 paper.tex
Remove end effects fig

@@
 
-The correction functions for the standard Glauert and lifting line methods are
-plotted for comparison in Figure~\ref{fig:end-effects}. Function values were
-computed using the UNH-RVAT geometry, $\lambda = \lambda_0 = 1.9$, and a 20
-degree tip angle of attack. Both end effects corrections produce similar results
-for this specific case, which gives confidence that the lifting line method will
-produce reasonable results without the uncertainty of translating axial-flow to
-cross-flow rotor parameters.
-
-\begin{figure}
-    \centering
-
-    \includegraphics[width=0.65\textwidth]{end-effects_rvat-20-deg}
-
-    \caption{End effect correction function values for the Glauert and lifting
-        line methods using UNH-RVAT parameters at 20 degrees tip angle of attack and
-        $\lambda=1.9$.}
-
-    \label{fig:end-effects}
-\end{figure}
-
 \section{Software implementation}

### 2016-05-12 paper.tex
Remove RM2 wake figs

@@
 
-Figure~\ref{fig:RM2-ALM-meancontquiv} shows the mean velocity field at 1 m
-downstream or $x/D=0.93$ computed by the ALM for the RM2. The mean near-wake
-structure looks similar to the RVAT ALM RANS case, but for the RM2, the lack of
-detail from blade tip vortex shedding matches more closely with
-experiments~\cite{Bachant2016-RM2-paper}. However, the vertical flow towards the
-$x$--$y$ center plane was captured, which is an important qualitative feature of
-both CFT near-wakes.
-
-\begin{figure}
-    \centering
-
-    \includegraphics[width=0.75\textwidth]{RM2-ALM_meancontquiv}
-
-    \caption{Mean velocity field at $x/D=0.93$ for the RM2 predicted by the
-        ALM.}
-
-    \label{fig:RM2-ALM-meancontquiv}
-\end{figure}
-
-Figure~\ref{fig:RM2-ALM-kcont} shows the ALM's turbulence kinetic energy
-predictions in the near-wake of the RM2. Like for the UNH-RVAT, $k$ appears to
-be concentrated on the $+y$ side of the rotor. However, overall levels of
-turbulence are lower than for the UNH-RVAT, which is consistent with the
-experimental results. However, turbulence generated by the RM2's blade tip
-vortex shedding was not captured by the ALM.
-
-\begin{figure}
-    \centering
-
-    \includegraphics[width=0.7\textwidth]{RM2-ALM_kcont}
-
-    \caption{Turbulence kinetic energy contours at $x/D=0.93$ behind the RM2
-        predicted by the ALM.}
-
-    \label{fig:RM2-ALM-kcont}
-\end{figure}
-
-Wake profiles at turbine mid-height in the $x$--$y$ center plane are shown in
-Figure~\ref{fig:RM2-ALM-profiles}. Like the UNH-RVAT case, the mean velocity
-deficit appears to be recovering too quickly, which may similarly be due to
-coarseness of the grid or overprediction of the eddy viscosity. The peak in
-turbulence kinetic energy on the $-y$ side of the turbine was also
-underpredicted, though was much lower in magnitude compared to the UNH-RVAT,
-even in the experimental measurements.
-
-\begin{figure}
-    \centering
-
-    \includegraphics[width=0.7\textwidth]{RM2-ALM_wake-profiles}
-
-    \caption{Mean streamwise velocity (left) and turbulence kinetic energy
-        (right) profiles at $x/D=0.93$ and $z/H=0$ computed for the RM2 using the
-        ALM inside a $k$--$\epsilon$ RANS model, compared with experimental data
-        from~\cite{Bachant2016-RM2-data}.}
-
-    \label{fig:RM2-ALM-profiles}
-\end{figure}
-
-Finally, a similar mean streamwise momentum transport analysis was undertaken by
-computing weighted sums of each term across the entire domain in the $y$--$z$
-directions. Similar results as for the UNH-RVAT were obtained, i.e.,
-cross-stream advection was predicted to be positive where it should have been
-negative, vertical advection was predicted reasonably well, and turbulent
-transport due to the eddy viscosity was also relatively large. The ratio of wake
-transport compared with the UNH-RVAT RANS case (approximately 60\% to 70\% lower
-for the RM2) matches well with that computed from the experiments, which shows
-the ALM may successfully predict larger optimal array spacing for the RM2 versus
-UNH-RVAT.
-
-\begin{figure}
-    \centering
-
-    \includegraphics[width=0.7\textwidth]{RM2-ALM_recovery-bar-chart}
-
-    \caption{Weighted average momentum recovery terms at $x/D=0.93$ for the RM2
-        actuator line model with a $k$--$\epsilon$ RANS closure and the experiments
-        described in \cite{Bachant2016-RM2-paper}.}
-
-    \label{fig:RM2-ALM-recovery}
-\end{figure}
+The mean near-wake structure for the RM2 looked qualitatively similar to the
+RVAT ALM RANS case, but for the RM2, the effects of blade tip vortex shedding
+were weaker, which is consistent with experiments~\cite{Bachant2016-RM2-paper}.
+Nonetheless, the mean vertical flow towards the $x$--$y$ center plane was
+captured, which is an important qualitative feature of both CFT near-wakes.
+Like for the UNH-RVAT, predicted turbulence kinetic energy values were
+concentrated on the $+y$ side of the rotor. However, overall levels of
+turbulence are lower than for the UNH-RVAT, which is also consistent with the
+experimental results.
+
+As for the UNH-RVAT RANS case, a mean streamwise momentum transport analysis was
+undertaken for the RM2 near-wake by computing weighted sums of each term across
+the entire domain in the $y$--$z$ directions. Similar results as for the
+UNH-RVAT were obtained, i.e., cross-stream advection was predicted to be
+positive where it should have been negative, vertical advection was predicted
+reasonably well, and turbulent transport due to the eddy viscosity was also
+relatively large. The ratio of wake transport compared with the UNH-RVAT RANS
+case (approximately 60\% to 70\% lower for the RM2) matches well with that
+computed from the experiments, which shows the ALM may successfully predict
+larger optimal array spacing for the RM2 versus UNH-RVAT.
 

### 2016-05-12 paper.tex
Remove future work

@@
 
-\section{Future work}
-
-The actuator line model shows great potential for use in cross-flow turbine
-engineering, and this study has inspired many avenues for its improvement.
-
-Firstly, since the fundamental premise of the ALM relies on accurate static foil
-coefficient data, it is crucial that more be measured and published openly, to
-allow the exploration of new rotor designs with various profiles. There is some
-doubt regarding the veracity of the Sheldahl and Klimas NACA foil data
-\cite{Bedon2014}. However, competing datasets at large laboratory scale Reynolds
-numbers are rare, even for the standard symmetrical foils. CFD may play a role
-in generating new data, but models must be rigorously validated, given the
-difficulty in predicting stall and its importance on CFT blade loading.
-
-Profile coefficient databases could be expanded to include more dimensions, or
-independent variables. For example, since the ALM will almost always involve
-turbulence modeling, the effects of local turbulence levels on foil coefficient
-data could be included, since stall delay could affect loading significantly,
-especially for a cross-flow turbine.
-
-Dynamic stall model empirical constants should be further investigated for their
-optimal values within the cross-flow turbine context. Though the
-Leishman--Beddoes type models are formulated in terms of a nondimensional time,
-there may be a dependence of the various time constants on the pitch rate, as
-indicated by the improvement in predictions for the RM2 at optimal tip speed
-ratio when using a higher angle of attack lag time constant. Machine learning
-techniques could be employed to ``train'' the constants (and/or their
-$\lambda$-dependence) to one turbine's power curve, and they could be validated
-against the other turbine's data. A more data-intensive alternative could be to
-develop dynamic cyclic profile coefficient databases---either experimentally or
-numerically---for many reduced frequencies, though this would be a difficult
-task, even for a single profile.
-
-Flow curvature corrections should be examined in more detail, and it may be
-beneficial to develop models to transform entire foil datasets to match their
-virtual camber, as suggested by Migliore et al.~\cite{Migliore1980}. It
-also may be advantageous to implement a complementary model to adjust the
-direction of the resultant force vector to more accurately generate the blade's
-shed vorticity.
-
-Rotor angular speed control via generator models could be included to test the
-effectiveness of various control schemes, e.g., sinusoidal $\lambda$ set points,
-which have been shown in some cases to improve mean power coefficient
-\cite{Strom2015}. Free stream velocity sampling could be implemented to
-determine the optimal location for sensing reference velocity used to compute
-tip speed ratio and power coefficient, which can become difficult in sheared or
-otherwise nonuniform flows \cite{Forbush2015}.
-
-Since turbinesFoam is compatible with OpenFOAM's volume of fluid (VOF)
-multiphase flow solver interFoam, the ALM's effectiveness should be assessed for
-predicting the generation and interaction of turbines with surface waves. Tow
-tank experiments could be performed with a wave gauge to measure and validate
-the free surface disturbance, which could be key to predicting the effects of
-placing turbines in channels at high blockage ratio.
-
-Possibly most importantly, the ALM should be validated with far-wake velocity
-measurements and/or performance data from two CFTs operating in proximity of
-each other. Next, the ALM should be evaluated for its use in modeling CFT arrays
-using both RANS and LES. If either can successfully postdict experimentally
-measured performance of closely-spaced physical model CFTs, they should then be
-coupled with optimization algorithms to automatically explore the design space,
-and ultimately drive down the cost of energy.
-
 \acks The authors would like to acknowledge funding through a National Science

### 2016-05-12 paper.tex
Add abbreviations for LB DS model

@@
 
-Dynamic stall is encountered when the blade angle of attack changes rapidly in
-time and exceeds a certain threshold, often near the static stall angle
+Dynamic stall (DS) is encountered when the blade angle of attack changes rapidly
+in time and exceeds a certain threshold, often near the static stall angle
 \cite{McCroskey1981}. The stall is characterized by an initial increase in lift
@@
 In this study we employed a dynamic stall model developed for low mach numbers
-by Sheng et al. \cite{Sheng2008}---derived from the Leishman--Beddoes
+by Sheng et al. \cite{Sheng2008}---derived from the Leishman--Beddoes (LB)
 semi-empirical model \cite{Leishman1989}. This model, along with two other

### 2016-05-12 paper.tex
Remove citation; removes about 100 words

@@
 open-source, widely used throughout industry and academia, and has grown a very
-active support community around itself \cite{Bachant2015-CFD-online}. Though
-SOWFA is also open-source, and has the ability to be coupled to NREL's FAST HAWT
-BEM code, its mostly procedural style would have required significant effort and
-duplicate code to adapt for cross-flow turbines. Thus, a new and more general
-ALM library was developed from the ground up that could model both cross- and
-axial-flow turbines, as well as standalone actuator lines. The actuator line
-model developed here, dubbed turbinesFoam
-\cite{Bachant2016-turbinesFoam-v0.0.7}, was also written as an extension library
-for OpenFOAM, and was developed freely and openly from its inception in order to
-increase community engagement and research efficiency.
+active support community around itself. Though SOWFA is also open-source, and
+has the ability to be coupled to NREL's FAST HAWT BEM code, its mostly
+procedural style would have required significant effort and duplicate code to
+adapt for cross-flow turbines. Thus, a new and more general ALM library was
+developed from the ground up that could model both cross- and axial-flow
+turbines, as well as standalone actuator lines. The actuator line model
+developed here, dubbed turbinesFoam \cite{Bachant2016-turbinesFoam-v0.0.7}, was
+also written as an extension library for OpenFOAM, and was developed freely and
+openly from its inception in order to increase community engagement and research
+efficiency.
 

### 2016-05-13 paper.tex
Remove paragraphs about momentum and vortex methods

@@
 
-Very fast solution times can be obtained with blade element momentum (BEM)
-models, such as the double multiple streamtube (DMST) approached developed by
-Paraschivoiu~\cite{Para1988}. However, their poor performance for rotors with
-high solidity~\cite{Joo2015} $Nc/R$ (and/or chord-to-radius ratio $c/R$), where
-$N$ is the number of blades, $c$ is the chord length, and $R$ is the maximum
-rotor radius, makes them less attractive, particular for marine hydrokinetic
-applications where rotors of the same size will encounter approximately an order
-of magnitude higher torque compared to wind turbines. Their description of the
-flow field is very crude as well, simply computing the momentum contained within
-discretized streamtubes, which cannot resolve any of the complicated flow
-structures shed by the rotor, limiting their applicability for array analysis.
-
-Blade element based vortex line methods, e.g., \cite{Strickland1981,
-Murray2011}, can be solved with reasonable computational effort when coupled
-with potential flow models---ranging from an expense close to BEM up to that of
-2-D blade-resolved CFD. The cost is a function of the number of vortex elements,
-which for free vortex methods may increase at each time step, slowing down the
-calculation as it marches forward in time, unless there is an additional model
-for vortex diffusion. Potential flow vortex methods can resolve more flow
-details than BEM. However, since the flow is governed by the Laplace equation,
-nonlinear effects such as the turbulent transport will not be included, which we
-have determined in experiments to be the same order of magnitude as the mean
-vertical advection.
-
-It is therefore desirable to retain a Navier--Stokes description of the flow
-field and use an actuator-type model for parameterizing the turbine loading,
-which dramatically drives down computational expense by removing the need for
-complicated meshes with many fine cells to resolve the boundary layers near the
-blade surfaces. As shown in \cite{Bachant2015-JoT}, the conventional uniform
-actuator disk is not a good candidate for a cross-flow turbine wake generator,
-never mind the fact that it does not typically compute performance predictions.
-There are then two actuator methods that use blade element theory to compute
-blade loading and therefore generate a wake that more closely resembles that of
-an actual turbine. These are the actuator cylinder or swept-surface model (ASSM)
-and the actuator line model (ALM). The ASSM solves for the average blade loading
-along its path and applies this as a constant body force term in the momentum
-equation. The ALM takes a similar approach but is an unsteady method, resolving
-the blade element locations in time.
+For analyzing turbine arrays, it is desirable to retain a Navier--Stokes
+description of the flow field---in contrast to, e.g., momentum or potential flow
+vortex models---to capture the effects of nonlinear advection and turbulent
+transport. However, rather than resolving the fine details in the blade boundary
+layers, an actuator-type model for parameterizing the turbine loading may be
+employed, which dramatically drives down computational expense. As shown in
+\cite{Bachant2015-JoT}, the conventional uniform actuator disk is not a good
+candidate for a cross-flow turbine wake generator, never mind the fact that it
+does not typically compute performance predictions. There are then two actuator
+methods that use blade element theory to compute blade loading and therefore
+generate a wake that more closely resembles that of an actual turbine. These are
+the actuator cylinder or swept-surface model (ASSM) and the actuator line model
+(ALM). The ASSM solves for the average blade loading along its path and applies
+this as a constant body force term in the momentum equation. The ALM takes a
+similar approach but is an unsteady method, resolving the blade element
+locations in time.
 

### 2016-05-13 paper.tex
Past tense

@@
 NACA 0021 coefficients were used for both turbines, despite the fact that the
-UNH-RVAT is constructed from NACA 0020 foils, as a NACA 0020 dataset is not
+UNH-RVAT is constructed from NACA 0020 foils, as a NACA 0020 dataset was not
 available---it is assumed the small difference in foil thickness is negligible.

### 2016-05-13 paper.tex
Remove more programming details

@@
 developed here, dubbed turbinesFoam \cite{Bachant2016-turbinesFoam-v0.0.7}, was
-also written as an extension library for OpenFOAM, and was developed freely and
-openly from its inception in order to increase community engagement and research
-efficiency.
-
-turbinesFoam was written in OpenFOAM's style, using OpenFOAM's
-\texttt{fvOptions} framework for adding source terms to equations at runtime.
-This implementation allows the CFT-ALM to be added to many of the standard
-solvers included in OpenFOAM without modification, i.e., it can be readily used
-with RANS or LES, multiphase models (e.g., for simulating the free surface in
-MHK installations), and even with heat transfer.
-
-OpenFOAM and turbinesFoam are written in the C++ programming language, which
-follows the object oriented programming paradigm. This characteristic helped
-modularize the ALM code for increased readability and reuse. In turbinesFoam, a
-turbine is a software object that is composed of actuator line objects, which
-themselves are composed of actuator line element objects. Structuring the code
-this way allows isolation and reuse of the functionality of individual
-components. For example, the actuator line object was written such that is could
-be used outside the turbine context to ensure it produces the correct forcing,
-without adding the complexity of rotation, other actuator lines, etc. that would
-be present in a turbine rotor. The very same actuator line objects can be used
-in both axial-flow and cross-flow rotors, without having to copy code from one
-to the other.
+also written as an extension library for OpenFOAM, using its \texttt{fvOptions}
+framework for adding source terms to equations at runtime. This implementation
+allows the CFT-ALM to be added to many of the standard solvers included in
+OpenFOAM without modification, i.e., it can be readily used with RANS or LES,
+multiphase models (e.g., for simulating the free surface in MHK installations),
+and even with heat transfer.
 

### 2016-05-13 paper.tex
Add OpenFOAM version

@@
 developed here, dubbed turbinesFoam \cite{Bachant2016-turbinesFoam-v0.0.7}, was
-also written as an extension library for OpenFOAM, using its \texttt{fvOptions}
-framework for adding source terms to equations at runtime. This implementation
-allows the CFT-ALM to be added to many of the standard solvers included in
-OpenFOAM without modification, i.e., it can be readily used with RANS or LES,
-multiphase models (e.g., for simulating the free surface in MHK installations),
-and even with heat transfer.
+also written as an extension library for OpenFOAM (version 3.0.x), using its
+\texttt{fvOptions} framework for adding source terms to equations at runtime.
+This implementation allows the CFT-ALM to be added to many of the standard
+solvers included in OpenFOAM without modification, i.e., it can be readily used
+with RANS or LES, multiphase models (e.g., for simulating the free surface in
+MHK installations), and even with heat transfer.
 

### 2016-05-13 paper.tex
Don't end sentence with two periods
Remove some intermediate equations from end effects section

@@
 direction of positive moment is ``nose-up,'' which is opposite that used by
-Strickland et al..
+Strickland et al.
 
@@
 rearranged into a matrix equation to solve for the unknown Fourier coefficients
-$A_n$,
-\begin{equation}
-    [\alpha_m ] - \alpha_{L=0} = [D_{mn}][A_n],
-\end{equation}
-where
-\begin{equation}
-    D_{mn} = \sum_1^N \left[ \frac{2b}{\pi c_m} \sin n \theta_m + n \frac{\sin n
-        \theta_m}{\sin \theta_m} \right].
-\end{equation}
-
-With the Fourier coefficients, the circulation distribution can be calculated as
+$A_n$, after which the circulation distribution can be calculated as
 \begin{equation}

### 2016-05-13 paper.tex
Slim down end effects intro

@@
 the lift distribution due to the bound vortex from foils of finite span must
-drop to zero at the tips.
-
-Glauert~\cite{Glauert1935} used Prandtl's lifting line theory \cite{Prandtl1927}
-to develop a tip loss correction factor $F$ to be applied to the local lift of
-an axial-flow rotor blade element. The Glauert correction was further refined
-for horizontal-axis wind turbines by Shen et al. \cite{Shen2005a}. Despite their
-success for the blade element analysis of axial-flow rotors, these corrections
-both depend on rotor parameters---tip speed ratio, number of blades, element
-radius, tip flow angle---that do no necessarily translate directly to the
-geometry and flow environment of a cross-flow rotor. Therefore, a more general
-end effects model was sought.
+drop to zero at the tips. One popular end effects correction was developed by
+Glauert~\cite{Glauert1935} for the blade element analysis of axial-flow rotors.
+However this correction depends on rotor parameters---tip speed ratio, number of
+blades, element radius, tip flow angle---that do no necessarily translate
+directly to the geometry and flow environment of a cross-flow rotor. Therefore,
+a more general end effects model was sought.
 
@@
 
-We can therefore compute a correction function based on the normalized spanwise
-lift coefficient distribution
+We can therefore compute a correction function $F$ to be applied to the ALM lift
+coefficient, based on the normalized spanwise lift coefficient distribution
 \begin{equation}

### 2016-05-13 paper.tex
Remove DS background info

@@
 
-Dynamic stall (DS) is encountered when the blade angle of attack changes rapidly
-in time and exceeds a certain threshold, often near the static stall angle
-\cite{McCroskey1981}. The stall is characterized by an initial increase in lift
-beyond static values as a vortex is shed from the foil's leading edge, after
-which a drop in lift and large nose-down pitching moment occurs as the vortex is
-advected downstream. As the angle of attack drops below the critical value, flow
-reattaches, closing the so-called hysteresis loop. Dynamic stall has been shown
-to be a significant positive contributor to performance in CFTs \cite{Para2002,
-    Urbina2013}, therefore an accurate model is key.
-
 In this study we employed a dynamic stall model developed for low mach numbers

### 2016-05-13 paper.tex
Remove comment about FAST

@@
 open-source, widely used throughout industry and academia, and has grown a very
-active support community around itself. Though SOWFA is also open-source, and
-has the ability to be coupled to NREL's FAST HAWT BEM code, its mostly
-procedural style would have required significant effort and duplicate code to
-adapt for cross-flow turbines. Thus, a new and more general ALM library was
-developed from the ground up that could model both cross- and axial-flow
+active support community around itself. Though SOWFA is also open-source, its
+mostly procedural style would have required significant effort and duplicate
+code to adapt for cross-flow turbines. Thus, a new and more general ALM library
+was developed from the ground up that could model both cross- and axial-flow
 turbines, as well as standalone actuator lines. The actuator line model

### 2016-05-13 paper.tex
Move comments about shaft foil data up
Remove sinusoidal tsr comments and save computational expense for the conclusions

@@
 
-\section{Static foil coefficient data}
+\section{Static profile coefficient data}
 
@@
 coefficients are then interpolated linearly within the database.
+
+Each rotor's shaft was assumed to have a drag coefficient $C_d = 1.1$, and the
+blade support strut end element drag coefficients were set to 0.05, to
+approximate the effects of separation in the corners of the blade--strut
+connections.
 
@@
 Bachant2016-UNH-RVAT-turbinesFoam-v1.0.0-LES}.
-
-The same foil coefficient data were used for all simulations---those for a NACA
-0021 as reported by Sheldahl and Klimas \cite{Sheldahl1981}. Each rotor's shaft
-was assumed to have a drag coefficient $C_d = 1.1$, and the blade support strut
-end element drag coefficients were set to 0.05, to approximate the effects of
-separation in the corners of the blade--strut connections.
-
-Since the ALM is intended to be an engineering tool when coupled with RANS, it
-was assumed that information about tip speed ratio due to control details---that
-is, sinusoidally oscillating $\lambda$---would not be known a priori, and was
-excluded, unlike the 3-D blade-resolved cases reported in
-\cite{Bachant2016-BR-CFD}. Note that a systematic investigation of the effects
-of sinusoidal $\lambda$ was not undertaken, but for the UNH-RVAT a one to two
-percentage point increase in $C_P$ was observed when running with similar
-parameters as the blade-resolved simulation.
-
-Compared to 3-D blade-resolved RANS, the ALM can solve a standalone turbine case
-on the order of 0.1 CPU hours per second of simulated time versus 1,000 CPU
-hours per second---a savings of 4 orders of magnitude.
 
@@
 time was $O(10)$ hours per second, which represent a four and two order of
-magnitude decrease in computational expense versus 3-D blade-resolved RANS,
-respectively.
+magnitude decrease in computational expense versus 3-D blade-resolved
+RANS~\cite{Bachant2016-BR-CFD}, respectively.
 

### 2016-05-13 paper.tex
Slim down conclusions

@@
 reduces computational effort compared to blade-resolved CFD, while maintaining
-the unsteadiness of the wake not resolved by a conventional actuator disk.
-Furthermore, its implementation within finite volume CFD provides a convenient
-way to study turbine array siting, since effects like terrain and buildings can
-be included, as can no-slip boundary conditions and boundary layer inflow
-profiles. Since the turbinesFoam library developed here uses OpenFOAM's modular
-\texttt{fvOptions} framework, the ALM body force term can be added as-is to
-compressible or multiphase solvers, for exploring more complicated physics.
-These features would be difficult or impossible to implement in a momentum or
-vortex models.
-
-Ultimately, the ALM retains some of the disadvantages of other blade element
-methods, but is a the next logical step bridging the gap between low- and
-high-fidelity flow modeling, and has been shown to be a method worthy of further
-advancement. The prospect of using ALM simulations to drive down computational
-cost of RANS by roughly four orders of magnitude---enabling 3-D unsteady
-Navier--Stokes simulations to be performed on a typical PC---itself justifies
-further development.
-
-At contemporary levels of computing power, the ALM is a useful tool for studying
-individual turbines when HPC is not feasible---requiring similar resources as
-2-D blade-resolved RANS, but with improved performance predictions. Furthermore,
-if arrays of CFTs advance to commercial scale, the ALM combined with LES
-represents one of the highest fidelity tools available, and will improve as
-turbulence modeling improves. It is therefore expected that this tool will prove
-valuable for both the engineering and research of wind and MHK turbines.
+the unsteadiness of the wake not resolved by a conventional actuator disk. When
+combined with RANS the ALM will allow VAT array simulations on individual PCs,
+and with high performance computing and LES the ALM is one of the highest
+fidelity array modeling tools available.
 

### 2016-05-13 paper.tex
Add concluding sentence
Move inflow velocity detection up and slim down theory

@@
 recovery due to vertical advection significantly larger than those from
-turbulent fluctuations~\cite{Bachant2015-JoT}---an effect not seen in HAT wakes.
-As interest shifts to designing and analyzing arrays of VATs, it is necessary to
-determine the effectiveness of various numerical modeling techniques to
-replicate VAT near-wake dynamics, such that wake recovery is accurately
-predicted, leading to accurate assessment of optimal array spacing.
+turbulent fluctuations~\cite{Bachant2015-JoT}---an effect not seen in axial-flow
+or horizontal-axis turbine (AFT or HAT) wakes. As interest shifts to designing
+and analyzing arrays of VATs, it is necessary to determine the effectiveness of
+various numerical modeling techniques to replicate VAT near-wake dynamics, such
+that wake recovery is accurately predicted, leading to accurate assessment of
+optimal array spacing.
 
@@
 blades as lines of blade elements, for which 2-D profile lift and drag
-coefficients are given. For each blade element, relative flow velocity and angle
-of attack are computed by adding the vectors of relative blade motion and the
-local fluid velocity, a diagram of which is shown in Figure~\ref{fig:vectors}.
+coefficients are given. For each blade element, relative flow velocity
+$\vec{U}_\mathrm{rel}$ and angle of attack $\alpha$ are computed by adding the
+vectors of relative blade motion $-\omega r$, where $\omega$ is the rotor
+angular velocity and $r$ is the blade element radius, and the local inflow
+velocity $\vec{U}_\mathrm{in}$, a diagram of which is shown in
+Figure~\ref{fig:vectors}.
 
@@
 
+Inflow velocity was sampled for each actuator line element at its quarter-chord
+location using OpenFOAM's \texttt{interpolationCellPoint} class, which provides
+a linear weighted interpolation using cell values. This algorithm helps keep the
+sampled velocity ``smooth'' compared with using the cell values themselves,
+especially when elements are moving in space as they are in a turbine, since
+meshes will likely have a cell size on the same order as the chord length, and
+will move on the order of one cell length per time step.
+
 The kinematics of a CFT are parameterized by the tip speed ratio $\lambda =
-\omega R/U_\infty$, where $\omega$ is the shaft's angular velocity, $R$ is the
-maximum rotor radius, and $U_\infty$ is the free stream velocity. Using only
-geometric considerations---that is, ignoring the effect of the rotor forcing and
-assuming the inflow velocity to be equal to that of the free stream---the
-geometric angle of attack $\alpha$ and relative velocity magnitude
-$|\vec{U}_{\mathrm{rel}}|$ can be calculated as a function of the rotor
-azimuthal angle $\theta$, which is plotted for various tip speed ratios in
-Figure~\ref{fig:geom-alpha-urel}.
-
-\begin{figure}
-    \centering
-
-    \includegraphics[width=0.75\textwidth]{CFT-vectors_alpha_deg_urel_geom}
-
-    \caption{Geometric angle of attack (left) and relative velocity (right)
-        versus azimuthal angle at various tip speed ratios.}
-
-    \label{fig:geom-alpha-urel}
-\end{figure}
-
-Inspecting Figure~\ref{fig:geom-alpha-urel}, it is apparent that CFTs are
-inherently unsteady machines, encountering large amplitude and rapid
+\omega R/U_\infty$, where $R$ is the maximum rotor radius, and $U_\infty$ is the
+free stream velocity. In contrast to an AFT, which under ideal conditions can be
+considered a steady machine, CFT blades encounter large amplitude and rapid
 oscillations in angle of attack, often exceeding the rotor blades' static stall
-angles in typical operation. The unsteadiness can be characterized by a reduced
+angles in typical operation, which makes their blade loading and performance
+more difficult to predict. The unsteadiness can be characterized by a reduced
 frequency~\cite{Leishman2006}
@@
 \end{figure}
-
-\section{Determining inflow velocity}
-
-Inflow velocity was sampled for each actuator line element at its quarter-chord
-location using OpenFOAM's \texttt{interpolationCellPoint} class, which provides
-a linear weighted interpolation using cell values. This algorithm helps keep the
-sampled velocity ``smooth'' compared with using the cell values themselves,
-especially when elements are moving in space as they are in a turbine, since
-meshes will likely have a cell size on the same order as the chord length, and
-will move on the order of one cell length per time step.
 
@@
 and with high performance computing and LES the ALM is one of the highest
-fidelity array modeling tools available.
+fidelity array modeling tools available. Ultimately, the ALM will help reduce
+dependence on expensive physical modeling to optimize VAT array layouts.
 

### 2016-05-13 paper.tex
Remove blade element discretization section

@@
 with a Navier--Stokes description of the flow field. The ALM treats turbine
-blades as lines of blade elements, for which 2-D profile lift and drag
-coefficients are given. For each blade element, relative flow velocity
-$\vec{U}_\mathrm{rel}$ and angle of attack $\alpha$ are computed by adding the
-vectors of relative blade motion $-\omega r$, where $\omega$ is the rotor
-angular velocity and $r$ is the blade element radius, and the local inflow
-velocity $\vec{U}_\mathrm{in}$, a diagram of which is shown in
-Figure~\ref{fig:vectors}.
+blades as lines of blade or actuator line elements, defined by their
+quarter-chord location, and for which 2-D profile lift and drag coefficients are
+known. For each blade element, relative flow velocity $\vec{U}_\mathrm{rel}$ and
+angle of attack $\alpha$ are computed by adding the vectors of relative blade
+motion $-\omega r$, where $\omega$ is the rotor angular velocity and $r$ is the
+blade element radius, and the local inflow velocity $\vec{U}_\mathrm{in}$, a
+diagram of which is shown in Figure~\ref{fig:vectors}.
 
@@
 \end{equation}
-
-\section{Blade element discretization}
-
-In the ALM, a turbine is a collection of actuator lines, which themselves are
-collections of actuator line elements (ALEs). The position of each ALE is a
-point in space indicating its quarter-chord location. The element is further
-defined by its chord direction vector, chord length, span direction vector, span
-length, and velocity vector.
-
-An actuator line is created from defined geometry points, between which ALE
-parameters are interpolated linearly. This way, an actuator line can be defined
-by fewer geometry points than element locations. For example, an AL with
-straight planform boundaries---e.g. a straight or tapered wing---only needs two
-geometry points to be fully defined. Figure~\ref{fig:AL-geom} shows and example
-schematic of a tapered actuator line with three geometry points at the
-half-chord locations and six total elements. Note that the middle geometry point
-is technically redundant, but is shown for illustrative purposes. For actuator
-lines that represent bluff bodies, e.g., shafts, the chord mount offset is set
-to 1/4, such that the element location is centered along the line of geometry
-points.
-
-\begin{figure}
-    \centering
-
-    \includegraphics[width=0.7\textwidth]{thesis_alm-geometry}
-
-    \caption{Actuator line geometry. Filled circles indicate points where
-    geometry is defined whereas squares indicate actuator line element
-    quarter-chord locations. Chord mount offset is defined as the distance from
-    an element's leading edge to each geometry point in chord lengths, which was
-    0.5 for both rotors modeled in this study.}
-
-    \label{fig:AL-geom}
-\end{figure}
 

### 2016-05-13 paper.tex
Combine turbines section with last intro paragraph

@@
 
-The ASSM and ALM were implemented to model a very low Reynolds number 2-D
-cross-flow turbine experiment in a flume using large eddy simulation
-(LES)~\cite{Shamsoddin2014}. Performance predictions for this case were not
-reported, but the ALM was shown to be more effective at postdicting the wake
-characteristics measured in the experiments by Brochier et
+The ASSM and ALM were implemented to model a very low Reynolds number (based on
+rotor diameter, $Re_D \sim 10^4$) 2-D cross-flow turbine experiment in a flume
+using large eddy simulation (LES)~\cite{Shamsoddin2014}. Performance predictions
+for this case were not reported, but the ALM was shown to be more effective at
+postdicting the wake characteristics measured in the experiments by Brochier et
 al.~\cite{Brochier1986}.
@@
 an extension for the OpenFOAM open-source CFD library. This model was validated
-against experimental data for both a high \cite{Bachant2015-JoT} and a medium
-\cite{Bachant2016-RM2-paper} solidity vertical-axis turbine.
-
-\section{Turbines}
-
-Two turbines were modeled: the high solidity ($c/R=0.28$) UNH Reference
+against experimental data for the high solidity ($c/R=0.28$) UNH Reference
 Vertical-Axis Turbine (UNH-RVAT) and the medium solidity ($c/R=0.07$--$0.12$) US
@@
 turbine (at 1:6 scale), both of which are shown in Figure~\ref{fig:turbines}.
-Both rotors are of similar geometric size ($\sim 1$ m) and are constructed from
-symmetrical NACA foils mounted at mid-chord and mid-span.
-
-Open datasets for the performance and near-wake dynamics of the UNH-RVAT and
-RM2, available from \cite{Bachant2014-RVAT-baseline,Bachant2016-RVAT-Re-dep} and
-\cite{Bachant2016-RM2-data}, respectively. have been acquired in the UNH towing
-tank at Reynolds numbers based on turbine diameter $Re_D = 10^6$. Analysis of
-these datasets showed that turbine performance and near-wake characteristics
-were nearly Reynolds number independent in this
-regime~\cite{Bachant2016-Energies,Bachant2016-RM2-paper}.
+Validation datasets for both turbines were taken from
+\cite{Bachant2016-RVAT-Re-dep} and \cite{Bachant2016-RM2-data}, respectively.
+The turbines were modeled at $Re_D \sim 10^6$, at which experimentally measured
+performance and near-wake characteristics were nearly Reynolds number
+independent~\cite{Bachant2016-Energies,Bachant2016-RM2-paper}.
 

### 2016-05-13 paper.tex
Remove unused syntax highlighting config

@@
 \usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
-
-%% Code syntax highlighting
-
-\usepackage{listings}
-\usepackage{color}
-
-\definecolor{mygreen}{rgb}{0,0.6,0}
-\definecolor{mygray}{rgb}{0.5,0.5,0.5}
-\definecolor{mymauve}{rgb}{0.58,0,0.82}
-\definecolor{mylightgray}{rgb}{0.95,0.95,0.95}
-
-\lstset{frame=single,
-    % language=C++,
-    aboveskip=3mm,
-    belowskip=3mm,
-    showstringspaces=false,
-    columns=flexible,
-    basicstyle=\ttfamily,
-    numbers=none,
-    numberstyle=\tiny\color{mygray},
-    keywordstyle=\bfseries\color{blue!40!black},
-    commentstyle=\color{gray},
-    stringstyle=\color{mygreen},
-    % otherkeywords={for,if,else},
-    breaklines=true,
-    breakatwhitespace=true,
-    tabsize=4,
-    escapeinside={\%*}{*)},
-    keepspaces=true,
-    captionpos=b,
-    backgroundcolor=\color{mylightgray}
-}
 

### 2016-08-27 reviews/reviewer1-response.md
Start reviewer 1 response--address novelty; towards #31

# Response to reviewer 1

We thank the reviewer for his or her detailed comments, which we have addressed
to improve the manuscript significantly.

>General remarks:
- I understand that the implementation of this work in OpenFOAM has demanded a lot of effort but could you please elaborate on what is new/novel in this work? An issue that has not been attacked/discovered/raised before or a numerical implementation which overcomes a known problem.

As noted in the "Novelty" file, this work presents some results that have either never been simulated or reported before:

1. Validating with Reynolds numbers more representative of a full-scale turbine, and two orders of magnitude higher than those used by Shamsoddin and Porte-Agel.
2. Reporting performance predictions, which has also not yet been done for a VAT in an ALM simulation.
3. Embedding the ALM inside of a RANS model for additional computational affordability.

The reviewer is correct that we had not sufficiently described this novelty in
the paper itself, and we have updated the introduction accordingly.

>- Minor comment: I'd rather use nondimentional units for variables such as time step
>
>Pages 1-6: Again, very good introduction and gathering of introductory note, but in my opinion, it
should be shortened for a journal.
>
Page 7, Fig 3: It is taking a lot of room, perhaps you can make a smaller schematic
>
Page 9 fig 6: The Cp figure is an interesting one. But first of all, but is there a particular reason
why you have not shown Ct next to it (it's more common this way than having Cd). Also, would it not be possible to have LES results on top of both figures 6 and 11 for the cases you have simulated?
>
Page 11, Line 42: I'm not quite sure if what you have stated about Smagorinsky model is correct. It
certainly is very easy to implement and gives nice results in some cases, but we are all aware of its
deficiencies. We set a fix Cs value which means we add numerical dissipation to the whole
computatiponal domain, but this is not really accurate when dealing with cases with strong directional
homogeneity and anisotropy. Why didn't you choose a dynamic Smagorinsky or even dynamic Lagrangian model? Besides, what is the Cs value you have used?
>
Page 13, Fig 15: Following my last comment, figure to the right may very well explain that the
Smagorinsly model has been over dissipative thereby hampering turbulent fluctuations which may have otherwise resulted in higher TKEs.
>
Page 14, Line 27: Matching mean velocities is usually rather straightforward (in non-presudospectral
codes, sometimes even shutting down the whole SGS modeling wouln'd make a difference as long as no solid walls are involved). But higher order moments such as Reynolds stress tensor components are usually more sensitive. Perhaps you can go more in depth with that part -- possibly compare your
results with other experiments or CFD simulations. Besides, please try and overcome the relatively large discrepancies seen in figures 6 (high TSR), 9, 11.

### 2016-08-27 paper.tex
Start reviewer 1 response--address novelty; towards #31

@@
 postdicting the wake characteristics measured in the experiments by Brochier et
-al.~\cite{Brochier1986}.
+al.~\cite{Brochier1986}. In this study we validated the ALM with experimental
+data taken at Reynolds number two orders of magnitude higher, reporting its
+performance predictions, neither of which have been done before. Additionally,
+assessed the use of the ALM in more computationally affordable unsteady RANS
+simulations, which also has not yet been reported for VATs.
 

### 2016-08-27 paper.tex
Shrink mesh fig in light of review 1
Note that rotor drag is often called thrust
Add line numbers to manuscript
Fix incomplete sentence

@@
 \usepackage{moreverb}
+
+\usepackage{lineno}
 
@@
 E-mail: pwbachant@gmail.com}
+
+\linenumbers
 
@@
 
-    \includegraphics[width=0.8\textwidth]{thesis_alm-mesh}
+    \includegraphics[width=0.65\textwidth]{thesis_alm-mesh}
 
@@
 
-Power and drag coefficient curves are plotted for the UNH-RVAT in
-Figure~\ref{fig:RVAT-ALM-perf-curves}. The ALM was successful at predicting the
-performance tip speed ratios up to $\lambda_0$, which suggests that dynamic
-stall was being modeled accurately, but $C_P$ was overpredicted at high
-$\lambda$. This may have been caused by the omission of additional parasitic
-drag sources such as roughness from exposed bolt heads located far enough from
-the axis to have a large effect at high rotation rates, or an underestimation of
-the blade--strut connection corner drag coefficient. In
+Power and overall rotor drag (a.k.a. thrust) coefficient curves are plotted for
+the UNH-RVAT in Figure~\ref{fig:RVAT-ALM-perf-curves}. The ALM was successful at
+predicting the performance tip speed ratios up to $\lambda_0$, which suggests
+that dynamic stall was being modeled accurately, but $C_P$ was overpredicted at
+high $\lambda$. This may have been caused by the omission of additional
+parasitic drag sources such as roughness from exposed bolt heads located far
+enough from the axis to have a large effect at high rotation rates, or an
+underestimation of the blade--strut connection corner drag coefficient. In
 \cite{Rawlings2008,Bachant2016-RM2-paper} it was shown how these losses can be
@@
 serves as a good standard for LES modeling since its behavior is well-reported
-in the literature. Default Smagorinsky model coefficients were used, LES filter
-width was set as the cube root of the local cell volume. The tip speed ratio was
-set to oscillate sinusoidally about $\lambda_0$ with a 0.19 magnitude and the
-angle of the first peak at 1.4 radians---similar to the rotation presribed in
-the blade-resolved RANS simulations discussed in \cite{Bachant2016-BR-CFD}.
+in the literature. Default Smagorinsky model coefficients were used, and the LES
+filter width was set as the cube root of the local cell volume. The tip speed
+ratio was set to oscillate sinusoidally about $\lambda_0$ with a 0.19 magnitude
+and the angle of the first peak at 1.4 radians---similar to the rotation
+presribed in the blade-resolved RANS simulations discussed in
+\cite{Bachant2016-BR-CFD}.
 

### 2016-08-27 paper.tex
Add default Smagorinsky coeff value

@@
 serves as a good standard for LES modeling since its behavior is well-reported
-in the literature. Default Smagorinsky model coefficients were used, and the LES
-filter width was set as the cube root of the local cell volume. The tip speed
-ratio was set to oscillate sinusoidally about $\lambda_0$ with a 0.19 magnitude
-and the angle of the first peak at 1.4 radians---similar to the rotation
-presribed in the blade-resolved RANS simulations discussed in
+in the literature. OpenFOAM's default Smagorinsky model coefficients were used
+(giving an approximate equivalent Smagorinsky coefficient $C_S = 0.17$), and the
+LES filter width was set as the cube root of the local cell volume. The tip
+speed ratio was set to oscillate sinusoidally about $\lambda_0$ with a 0.19
+magnitude and the angle of the first peak at 1.4 radians---similar to the
+rotation presribed in the blade-resolved RANS simulations discussed in
 \cite{Bachant2016-BR-CFD}.

### 2016-08-27 reviews/reviewer1-response.md
Note how the Smagorinsky model is a good baseline, not necessarily the most accurate SGS model

@@
 >- Minor comment: I'd rather use nondimentional units for variables such as time step
->
->Pages 1-6: Again, very good introduction and gathering of introductory note, but in my opinion, it
-should be shortened for a journal.
->
-Page 7, Fig 3: It is taking a lot of room, perhaps you can make a smaller schematic
->
-Page 9 fig 6: The Cp figure is an interesting one. But first of all, but is there a particular reason
-why you have not shown Ct next to it (it's more common this way than having Cd). Also, would it not be possible to have LES results on top of both figures 6 and 11 for the cases you have simulated?
->
-Page 11, Line 42: I'm not quite sure if what you have stated about Smagorinsky model is correct. It
-certainly is very easy to implement and gives nice results in some cases, but we are all aware of its
-deficiencies. We set a fix Cs value which means we add numerical dissipation to the whole
-computatiponal domain, but this is not really accurate when dealing with cases with strong directional
-homogeneity and anisotropy. Why didn't you choose a dynamic Smagorinsky or even dynamic Lagrangian model? Besides, what is the Cs value you have used?
->
-Page 13, Fig 15: Following my last comment, figure to the right may very well explain that the
+
+Figure 6 does use a time step that is technically nondimensional (time steps per rotor revolution).
+
+>Pages 1-6: Again, very good introduction and gathering of introductory note,
+but in my opinion, it should be shortened for a journal.
+
+We agree and have shortened the introduction significantly.
+
+TODO: Shorten introduction
+
+>Page 7, Fig 3: It is taking a lot of room, perhaps you can make a smaller
+schematic
+
+We have made this figure smaller in the manuscript, a change that may be
+retained by the typesetter.
+
+>Page 9 fig 6: The Cp figure is an interesting one. But first of all, but is
+there a particular reason why you have not shown Ct next to it (it's more common
+this way than having Cd). Also, would it not be possible to have LES results on
+top of both figures 6 and 11 for the cases you have simulated?
+
+We use the overal rotor drag coefficient nomenclature ($C_D$) since this was
+adopted by Paraschivoiu (2001). As noted, this is commonly called the thrust
+coefficient for horizontal-axis turbines. We have added a note to the manuscript
+to clarify this.
+
+We have not plotted the LES results, since we only used LES for one tip speed
+ratio, and for one turbine---it was mainly employed to assess wake predictions.
+
+>Page 11, Line 42: I'm not quite sure if what you have stated about Smagorinsky
+model is correct. It certainly is very easy to implement and gives nice results
+in some cases, but we are all aware of its deficiencies. We set a fix Cs value
+which means we add numerical dissipation to the whole computatiponal domain, but
+this is not really accurate when dealing with cases with strong directional
+homogeneity and anisotropy. Why didn't you choose a dynamic Smagorinsky or even
+dynamic Lagrangian model? Besides, what is the Cs value you have used?
+
+Firstly, the $C_S$ value used was OpenFOAM's default value of approximately
+0.17. This detail has been added to the text.
+
+Secondly, we are not trying to claim that the Smagorinsky model is the most
+accurate "standard" model, but that is has been used in many other ALM papers,
+and therefore a good baseline for comparison. The wording of the text has been
+adjusted to reflect the notion that we are interested in this comparison to
+baseline for turbine ALM modeling.
+
+>Page 13, Fig 15: Following my last comment, figure to the right may very well explain that the
 Smagorinsly model has been over dissipative thereby hampering turbulent fluctuations which may have otherwise resulted in higher TKEs.

### 2016-08-27 paper.tex
Note how the Smagorinsky model is a good baseline, not necessarily the most accurate SGS model

@@
 turbulence model \cite{Smagorinsky1963}, which was the first of its kind, and
-serves as a good standard for LES modeling since its behavior is well-reported
-in the literature. OpenFOAM's default Smagorinsky model coefficients were used
-(giving an approximate equivalent Smagorinsky coefficient $C_S = 0.17$), and the
-LES filter width was set as the cube root of the local cell volume. The tip
-speed ratio was set to oscillate sinusoidally about $\lambda_0$ with a 0.19
-magnitude and the angle of the first peak at 1.4 radians---similar to the
-rotation presribed in the blade-resolved RANS simulations discussed in
-\cite{Bachant2016-BR-CFD}.
+serves as a good baseline for LES modeling since its behavior is well-reported
+in the literature, especially for ALM simulations. OpenFOAM's default
+Smagorinsky model coefficients were used (giving an approximate equivalent
+Smagorinsky coefficient $C_S = 0.17$), and the LES filter width was set as the
+cube root of the local cell volume. The tip speed ratio was set to oscillate
+sinusoidally about $\lambda_0$ with a 0.19 magnitude and the angle of the first
+peak at 1.4 radians---similar to the rotation presribed in the blade-resolved
+RANS simulations discussed in \cite{Bachant2016-BR-CFD}.
 

### 2016-08-27 reviews/reviewer1-response.md
Address comment about over-dissipative nature of Smagorinsky model

@@
 adjusted to reflect the notion that we are interested in this comparison to
-baseline for turbine ALM modeling.
+baseline for turbine ALM modeling, and are by no means implying that this SGS
+model is the most accurate or appropriate for this case. Certainly testing
+additional SGS models will be part of future work, as is mentioned in the
+conclusions.
 
->Page 13, Fig 15: Following my last comment, figure to the right may very well explain that the
-Smagorinsly model has been over dissipative thereby hampering turbulent fluctuations which may have otherwise resulted in higher TKEs.
->
-Page 14, Line 27: Matching mean velocities is usually rather straightforward (in non-presudospectral
+>Page 13, Fig 15: Following my last comment, figure to the right may very well
+explain that the Smagorinsly model has been over dissipative thereby hampering
+turbulent fluctuations which may have otherwise resulted in higher TKEs.
+
+This is a valid point, and we have added a comment to the text accordingly. It
+is important to note, however, that since we don't have near-wall boundary
+layers, it may be that turbulence is only generated by larger-scale shear and
+vortex breakdown, rather than by boundary layer separation or transition, which
+would occur with a body-fitted mesh.
+
+>Page 14, Line 27: Matching mean velocities is usually rather straightforward (in non-presudospectral
 codes, sometimes even shutting down the whole SGS modeling wouln'd make a difference as long as no solid walls are involved).

### 2016-08-27 paper.tex
Address comment about over-dissipative nature of Smagorinsky model

@@
 generated in the large eddy simulation was approximately an order of magnitude
-too low. Once again this was probably a result of the subgrid-scale modeling and
-its effect on the stability of vortex structures.
+too low. Once again this was likely a result of the subgrid-scale modeling and
+its effect on the stability of vortex structures, or its over-dissipative nature
+removing energy from the flow before it is advected to the plane of interest.
 

### 2016-08-27 reviews/reviewer1-response.md
Text wrapping in reviewer 1 response
Tweak r1 response

@@
 
-Firstly, the $C_S$ value used was OpenFOAM's default value of approximately
-0.17. This detail has been added to the text.
-
-Secondly, we are not trying to claim that the Smagorinsky model is the most
-accurate "standard" model, but that is has been used in many other ALM papers,
-and therefore a good baseline for comparison. The wording of the text has been
+We are not trying to claim that the Smagorinsky model is the most accurate or
+"the standard" model, but that is has been used in many other ALM papers, and
+therefore a good baseline for comparison. The wording of the text has been
 adjusted to reflect the notion that we are interested in this comparison to
@@
 conclusions.
+
+The $C_S$ value used was OpenFOAM's default value of approximately 0.17. This
+detail has been added to the text.
 
@@
 
->Page 14, Line 27: Matching mean velocities is usually rather straightforward (in non-presudospectral
-codes, sometimes even shutting down the whole SGS modeling wouln'd make a difference as long as no solid walls are involved).
-But higher order moments such as Reynolds stress tensor components are usually more sensitive.
-Perhaps you can go more in depth with that part -- possibly compare your
-results with other experiments or CFD simulations. Besides, please try and overcome the relatively large discrepancies seen in figures 6 (high TSR), 9, 11.
+>Page 14, Line 27: Matching mean velocities is usually rather straightforward
+(in non-presudospectral codes, sometimes even shutting down the whole SGS
+modeling wouln'd make a difference as long as no solid walls are involved). But
+higher order moments such as Reynolds stress tensor components are usually more
+sensitive. Perhaps you can go more in depth with that part -- possibly compare
+your results with other experiments or CFD simulations. Besides, please try and
+overcome the relatively large discrepancies seen in figures 6 (high TSR), 9, 11.

### 2016-08-27 reviews/reviewer1-response.md
Finish up R1 response

@@
 We agree and have shortened the introduction significantly.
-
-TODO: Shorten introduction
 
@@
 overcome the relatively large discrepancies seen in figures 6 (high TSR), 9, 11.
+
+We think it is unlikely that the discrepancies seen in predicted loading are
+related to higher order statistics of the flow field, as simplier blade element
+methods, e.g., momentum and potential flow vortex methods, also often produce
+similar overpredictions at higher TSR. It should also be noted that neither the
+static nor dynamic blade element loading is computed as a function of the local
+flow unsteadiness or TKE, though this would be an interesting avenue for future
+investigation, which has never been attempted to our knowledge.

### 2016-08-27 paper.tex
Slim down intro a little; resolves #31

@@
 candidate for a cross-flow turbine wake generator, never mind the fact that it
-does not typically compute performance predictions. There are then two actuator
-methods that use blade element theory to compute blade loading and therefore
-generate a wake that more closely resembles that of an actual turbine. These are
-the actuator cylinder or swept-surface model (ASSM) and the actuator line model
-(ALM). The ASSM solves for the average blade loading along its path and applies
-this as a constant body force term in the momentum equation. The ALM takes a
-similar approach but is an unsteady method, resolving the blade element
-locations in time.
-
-The ALM, originally developed by Sorensen and Shen \cite{Sorensen2002}, has
-become popular for modeling axial-flow or horizontal-axis turbines, and has been
-shown in blind tests to be competitive with blade-resolved
+does not typically compute performance predictions.
+
+Actuator line modeling (ALM), originally developed by Sorensen and Shen
+\cite{Sorensen2002}, is an unsteady method that tracks blade element locations,
+and has become popular for modeling axial-flow or horizontal-axis turbines, and
+has been shown in blind tests to be competitive with blade-resolved
 CFD~\cite{Krogstad2013, Pierella2014}. The ALM combined with large eddy
@@
 postdicting the wake characteristics measured in the experiments by Brochier et
-al.~\cite{Brochier1986}. In this study we validated the ALM with experimental
-data taken at Reynolds number two orders of magnitude higher, reporting its
-performance predictions, neither of which have been done before. Additionally,
-assessed the use of the ALM in more computationally affordable unsteady RANS
-simulations, which also has not yet been reported for VATs.
-
-It is therefore proposed that an actuator line model may be the optimal
-combination of high-fidelity flow modeling that includes performance
-predictions, but with reduced computational expense. Here we have developed an
-ALM for cross-flow turbines inside both RANS and LES simulations, implemented as
-an extension for the OpenFOAM open-source CFD library. This model was validated
-against experimental data for the high solidity ($c/R=0.28$) UNH Reference
-Vertical-Axis Turbine (UNH-RVAT) and the medium solidity ($c/R=0.07$--$0.12$) US
-Department of Energy/Sandia National Labs Reference Model 2 (RM2) cross-flow
-turbine (at 1:6 scale), both of which are shown in Figure~\ref{fig:turbines}.
-Validation datasets for both turbines were taken from
-\cite{Bachant2016-RVAT-Re-dep} and \cite{Bachant2016-RM2-data}, respectively.
-The turbines were modeled at $Re_D \sim 10^6$, at which experimentally measured
-performance and near-wake characteristics were nearly Reynolds number
-independent~\cite{Bachant2016-Energies,Bachant2016-RM2-paper}.
+al.~\cite{Brochier1986}.
+
+In this study we have developed an ALM for cross-flow turbines and embedded the
+model inside LES and unsteady RANS simulations, the latter of which has not yet
+been reported in the literature. This model was validated against experimental
+data for the high solidity ($c/R=0.28$) UNH Reference Vertical-Axis Turbine
+(UNH-RVAT) and the medium solidity ($c/R=0.07$--$0.12$) US Department of
+Energy/Sandia National Labs Reference Model 2 (RM2) cross-flow turbine (at 1:6
+scale), both of which are shown in Figure~\ref{fig:turbines}. Validation
+datasets for both turbines were taken from \cite{Bachant2016-RVAT-Re-dep} and
+\cite{Bachant2016-RM2-data}, respectively. The turbines were modeled at $Re_D
+\sim 10^6$, at which experimentally measured performance and near-wake
+characteristics were nearly Reynolds number
+independent~\cite{Bachant2016-Energies,Bachant2016-RM2-paper}, and two orders of
+magnitude higher than previous ALM investigations for VATs.
 

### 2016-08-27 reviews/reviewer2-response.md
Start response to reviewer 2

# Response to reviewer 2

>1. Pag1. Please insert the list of Symbols since there are a lot of
abbreviations in the text, a quick reference could be useful.

>2. Pag. Pag3. Sect. 2. The parameter reduced frequency "k" is the key parameter
for unsteadiness considerations concerning VAT: some more detailed comments
should be given concerning the actual value of this parameter in the case study
numerically modelled. At the same time a comment should be given concerning the
ALM capabilities (and/or future improvements) in terms of unsteady aerodynamics.

>3. Pag4. Sect.4.  Is the force smoothing with the Gaussian function really
necessary? In P.-E. Réthoré and N. N. Sørensen, A discrete force allocation
algorithm, Wind Energy, 2012. Rethore and Sorensen showed how this spatial force
filter could be avoided with modified Rhie–Chow interpolation algorithm.

In our experience, especially with the finer LES grids, the force smoothing is
necessary. Without smoothing, the blade loading becomes quite "noisy" due to the
subsequent instability from sharper momentum source gradients. In this study we
did not investigate changes to the numerical algorithms used to solve the flow
field, but certainly future research may prove the Gaussian width can be
narrowed or eliminated if, for example, modified Rhie--Chow interpolation is
implemented in OpenFOAM.

>4. Pag6. Sect.7. Does the authors took into consideration that the ALM could be
in theory able to catch the Tip vortex effect without a correction model? i.e.
the vortex generated by the lifting of the blade line is already influencing the
angle of attack on the blade.

>5. Pag7. Sect.8-9 . Add a sketch of the domain boundaries with the position of
the turbine, and the flow direction. Figure 3 is not very effective in terms of
quantitative information on the actual implementation of the code.

>6. Sect. 8-9 The details on the domain boundaries of the implemented case study
should be an opportunity for clarify if any blockage effect was present in the
reference experiments and if the same blockage conditions were adopted in the
numerical implementation.

>7. Pag12. Sect.9.4 "The near-wake's mean velocity field at x/D=1 is shown in
Figure 12..." should be actually Figure 13

### 2016-08-27 paper.tex
Add nomenclature per reviewer 2 request

@@
 \usepackage{lineno}
+
+\usepackage{nomencl}
+\makenomenclature
 
@@
 
+\printnomenclature
+
 \section{Introduction}
@@
 faster wake recovery.
+
+\nomenclature{CFT}{Cross-flow turbine}
+\nomenclature{VAT}{Vertical-axis turbine}
+\nomenclature{AFT}{Axial-flow turbine}
+\nomenclature{HAT}{Horizontal-axis turbine}
 
@@
 
+\nomenclature{RANS}{Reynolds-averaged Navier--Stokes}
+
 For analyzing turbine arrays, it is desirable to retain a Navier--Stokes
@@
 CFD~\cite{Krogstad2013, Pierella2014}. The ALM combined with large eddy
-simulation has become the state-of-the-art for modeling entire wind
+simulation (LES) has become the state-of-the-art for modeling entire wind
 farms~\cite{Archer2013, Churchfield2012, Sorensen2015, Fleming2013,
@@
 
-The ASSM and ALM were implemented to model a very low Reynolds number (based on
+\nomenclature{ALM}{Actuator line model}
+\nomenclature{LES}{Large eddy simulation}
+
+The ALM has been previously used to model a very low Reynolds number (based on
 rotor diameter, $Re_D \sim 10^4$) 2-D cross-flow turbine experiment in a flume
@@
 
+\nomenclature{$c$}{Blade chord length}
+\nomenclature{$R$}{Maximum rotor radius}
+\nomenclature{$Re_D$}{Rotor diameter Reynolds number}
+\nomenclature{DOE}{US Department of Energy}
+\nomenclature{SNL}{Sandia National Laboratories}
+
 \begin{figure}
@@
 
+\nomenclature{$\vec{U}_\mathrm{rel}$}{Blade element local relative velocity}
+\nomenclature{$\alpha$}{Blade element angle of attack}
+\nomenclature{$\omega$}{Rotor angular velocity}
+\nomenclature{$r$}{Blade element radius}
+\nomenclature{$\vec{U}_\mathrm{in}$}{Blade element local inflow velocity}
+
 \begin{figure}
@@
 
+\nomenclature{$U_\infty$}{Free stream velocity}
+
 Inflow velocity was sampled for each actuator line element at its quarter-chord
@@
 become significant for $k > 0.05$, and can become dominant for $k \ge 0.2$.
+
+\nomenclature{$\lambda$}{Rotor tip speed ratio}
+\nomenclature{$k$}{Reduced frequency or turbulence kinetic energy}
 
@@
 
+\nomenclature{$C_l$}{Blade element lift coefficient}
+\nomenclature{$C_d$}{Blade element drag coefficient}
+\nomenclature{$C_m$}{Blade element moment coefficient}
+
 \section{Static profile coefficient data}
@@
 $Re_c \sim 10^5$.
+
+\nomenclature{$Re_c$}{Blade chord Reynolds number}
 
@@
 
+\nomenclature{$\epsilon$}{Gaussion force smoothing function width}
+
 \subsection{Verification}
@@
 
+\nomenclature{$C_P$}{Rotor power coefficient}
+\nomenclature{$C_D$}{Overall rotor drag (a.k.a. thrust) coefficient}
+\nomenclature{$D$}{Rotor diameter}
+
 \begin{figure}
@@
 \end{figure}
+
+\nomenclature{$H$}{Rotor height}
 
@@
 
+\nomenclature{$\lambda_0$}{Tip speed ratio of maximum power coefficient}
+
 \begin{figure}

### 2016-08-27 reviews/reviewer2-response.md
Add nomenclature per reviewer 2 request
Add comment on tip vortices to reviewer 1 response

@@
 abbreviations in the text, a quick reference could be useful.
+
+We have updated the manuscript with a nomenclature section.
 
@@
 
+While tip vortices are produced by the ALM, and these do affect the local angle
+of attack, we are not sure how easy it would be to create a ALM that resolves
+tip effects totally accurately without any correction model. The issue may be
+that since we do not enforce the no peneration BC through the blade, the tip
+vortex can leave the blade too early, which causes incorrect angles of attack
+close to the tip.
+
 >5. Pag7. Sect.8-9 . Add a sketch of the domain boundaries with the position of

### 2016-08-27 reviews/reviewer2-response.md
Fix cross-referencing error
Add comment about blockage

@@
 
+The cross-section of the domain was set equal to match blockage, though the free
+surface was not modeled. However, we did use a slip BC for the top as opposed to
+a fixed tow speed condition, to get closer to the free surface condition. We
+have added a small clarification regarding blockage matching to the text.
+
 >7. Pag12. Sect.9.4 "The near-wake's mean velocity field at x/D=1 is shown in
 Figure 12..." should be actually Figure 13
+
+Thank you for catching this error. The manuscript has been corrected.

### 2016-08-27 paper.tex
Fix cross-referencing error
Add comment about blockage

@@
 simulations were performed inside a domain similar in size to that used in
-\cite{Bachant2016-BR-CFD} to mimic the tow tank experiments (3.66 m wide, 2.44 m
-tall, 1.52 m upstream and 2.16 m downstream), with similar velocity boundary
-conditions: 1 m/s inflow, no-slip (fixed to 1 m/s) walls and bottom, and a rigid
-slip condition for the top, to approximate some effects of the tow tank's free
-surface. Simulations were run for a total of 6 seconds, with the latter half
-used to calculate performance and wake statistics. Pressure--velocity coupling
-for the momentum equation was achieved using the PISO (pressure implicit
-splitting of operators) method. A slice of the mesh in the $x$--$y$ plane is
-shown in Figure~\ref{fig:ALM-mesh}.
+\cite{Bachant2016-BR-CFD} to approximately (including blockage) the tow tank
+experiments (3.66 m wide, 2.44 m tall, 1.52 m upstream and 2.16 m downstream),
+with similar velocity boundary conditions: 1 m/s inflow, no-slip (fixed to 1
+m/s) walls and bottom, and a rigid slip condition for the top, to approximate
+some effects of the tow tank's free surface. Simulations were run for a total of
+6 seconds, with the latter half used to calculate performance and wake
+statistics. Pressure--velocity coupling for the momentum equation was achieved
+using the PISO (pressure implicit splitting of operators) method. A slice of the
+mesh in the $x$--$y$ plane is shown in Figure~\ref{fig:ALM-mesh}.
 
@@
 The near-wake's mean velocity field at $x/D=1$ is shown in
-Figure~\ref{RVAT-ALM-LES-vorticity}. Compared with the RANS ALM results, the LES
-looks much more like the blade-resolved and experimental results
+Figure~\ref{RVAT-ALM-LES-meancontquiv}. Compared with the RANS ALM results, the
+LES looks much more like the blade-resolved and experimental results
 \cite{Bachant2015-JoT}, showing the clockwise and counterclockwise mean swirling

### 2016-08-27 paper.tex
Make new figure to show computational domain

@@
 
-    \includegraphics[width=0.65\textwidth]{thesis_alm-mesh}
-
-    \caption{$x$--$y$ planar slice of the mesh used for the ALM RANS
-        simulations.}
+    \includegraphics[width=0.8\textwidth]{alm-domain}
+
+    \caption{Computational domain used for the RANS ALM simulations. All
+    dimensions are in meters, flow is oriented in the $+x$-direction, and the
+    cylindrical region represents the swept area of the UNH-RVAT rotor. Note
+    that increased mesh refinrement was used for the LES simulations.}
 

### 2016-08-27 reviews/reviewer2-response.md
Make new figure to show computational domain
Add note about reduced frequencies encountered; resolves #32

@@
 ALM capabilities (and/or future improvements) in terms of unsteady aerodynamics.
+
+We have added values for the reduced frequencies of the rotors tested here, and
+a note that the LB dynamic stall models do contain corrections for unsteady
+aerodynamics even in the absence of stall. This implies that these models should
+work better than simply using static foil data, e.g., at high TSR, where angles
+of attack do not exceed the static stall angle.
 
@@
 
+This is a valid point. We have created a new figure to better describe the
+computational domain and dimensions.
+
 >6. Sect. 8-9 The details on the domain boundaries of the implemented case study

### 2016-08-27 paper.tex
Add note about reduced frequencies encountered; resolves #32

@@
 For a VAT or CFT reduced frequency can be reformulated in terms of the tip speed
-ratio as
-\begin{equation}
-    k = \frac{\lambda c}{2R},
-\end{equation}
-which is then also a function of solidity or chord-to-radius ratio $c/R$. As an
-example, a large scale, relatively low solidity Darrieus wind turbine such as
-the Sandia 34 m diameter Test Bed, with an equatorial blade chord of 0.91 m
-\cite{Murray2011}, a reduced frequency $k=0.16$ is encountered based solely on
-angle of attack oscillations at $\lambda=6$. For a smaller scale CFT, e.g., with
-$c/R = 0.25$, operating at $\lambda = 2$, the reduced frequency is 0.25. It
-follows that unsteady effects will be significant for cross-flow turbines even
-in the absence of nonlinear effects such as stall.
+ratio as \begin{equation} k = \frac{\lambda c}{2R}, \end{equation} which is then
+also a function of solidity or chord-to-radius ratio $c/R$. As an example, a
+large scale, relatively low solidity Darrieus wind turbine such as the Sandia 34
+m diameter Test Bed, with an equatorial blade chord of 0.91 m \cite{Murray2011},
+a reduced frequency $k=0.16$ is encountered based solely on angle of attack
+oscillations at $\lambda=6$. For the two turbines studied here, reduced
+frequencies at $\lambda_0$ are 0.27 and 0.17 (based on root chord length) for
+the UNH-RVAT and RM2, respectively. Therefore, both cases will be inherently
+unsteady, with these effects dominating the UNH-RVAT's behavior. Hence, an
+accurate unsteady aerodynamics model is important to predicting blade loading.
 
@@
 Gormont model \cite{Gormont1973}, which inspired its adoption here for the ALM.
+Note that even in the absence of stall, the LB DS models still modulate foil
+force and moment coefficients to account for unsteadiness, which is important
+for the reduced frequencies of the rotors simulated here.
 

### 2016-08-27 scripts/diff.sh
Add diff script

@@ -1,0 +1,11 @@
+#!/usr/bin/env bash
+# This script creates a diff version comparing the recent with whatever revision
+# is supplied as the first argument. This can be any valid Git identifier, e.g.,
+# a commit hash, tag, etc.
+
+rev=$1
+git show $rev:paper.tex > archive/paper-$rev.tex
+latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$rev.tex paper.tex > paper-$rev-diff.tex
+latexmk paper-$rev-diff.tex -pdf -f
+mv paper-$rev-diff.pdf archive
+rm paper-$rev*

### 2016-08-27 paper.tex
Fix broken crossref
Fix typo

@@
 simulations were performed inside a domain similar in size to that used in
-\cite{Bachant2016-BR-CFD} to approximately (including blockage) the tow tank
+\cite{Bachant2016-BR-CFD} to approximatel (including blockage) the tow tank
 experiments (3.66 m wide, 2.44 m tall, 1.52 m upstream and 2.16 m downstream),
@@
 The near-wake's mean velocity field at $x/D=1$ is shown in
-Figure~\ref{RVAT-ALM-LES-meancontquiv}. Compared with the RANS ALM results, the
-LES looks much more like the blade-resolved and experimental results
+Figure~\ref{fig:RVAT-ALM-LES-meancontquiv}. Compared with the RANS ALM results,
+the LES looks much more like the blade-resolved and experimental results
 \cite{Bachant2015-JoT}, showing the clockwise and counterclockwise mean swirling

### 2016-09-11 scripts/diff.sh
Add diff target to Makefile and always compare to last tag

@@ -1,11 +1,11 @@
 #!/usr/bin/env bash
-# This script creates a diff version comparing the recent with whatever revision
-# is supplied as the first argument. This can be any valid Git identifier, e.g.,
-# a commit hash, tag, etc.
+# This script creates a diff version comparing the current version with the
+# most recent tagged version
 
-rev=$1
-git show $rev:paper.tex > archive/paper-$rev.tex
-latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$rev.tex paper.tex > paper-$rev-diff.tex
-latexmk paper-$rev-diff.tex -pdf -f
-mv paper-$rev-diff.pdf archive
-rm paper-$rev*
+PREVIOUS=$(git tag | tail -n1)
+CURRENT=$(git describe --always --dirty)
+git show $PREVIOUS:paper.tex > archive/paper-$PREVIOUS.tex
+latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$PREVIOUS.tex paper.tex > paper-$PREVIOUS-$CURRENT-diff.tex
+latexmk paper-$PREVIOUS-$CURRENT-diff.tex -pdf -f
+mv paper-$PREVIOUS-$CURRENT-diff.pdf archive
+rm paper-$PREVIOUS-$CURRENT*

### 2016-09-11 scripts/zip.py
Use --always and --dirty with zip script

@@ -6,7 +6,8 @@ from subprocess import check_output
 import os
 
 # Describe version via git
-version = check_output(["git", "describe"]).strip().decode()
+version = check_output(["git", "describe", "--always",
+                        "--dirty"]).strip().decode()
 
 figures = ["figures/" + f for f in os.listdir("figures") \
            if f.endswith(".pdf") or f.endswith(".png")]

### 2016-09-11 scripts/diff.sh
Fix diff if latest version is tagged

@@ -4,6 +4,13 @@
 
 PREVIOUS=$(git tag | tail -n1)
 CURRENT=$(git describe --always --dirty)
+
+# If current version is a tag, use one tag prior
+if [ $CURRENT = $PREVIOUS ]
+then
+    PREVIOUS=$(git tag | tail -n2 | head -n1)
+fi
+
 git show $PREVIOUS:paper.tex > archive/paper-$PREVIOUS.tex
 latexdiff --exclude-textcmd="section,subsection,subsubsection" --math-markup=off archive/paper-$PREVIOUS.tex paper.tex > paper-$PREVIOUS-$CURRENT-diff.tex
 latexmk paper-$PREVIOUS-$CURRENT-diff.tex -pdf -f

### 2018-03-29 reviews/reviewer1-response.md
Update review responses

--- 2016-08-27 reviews/reviewer1-response.md
@@
 to improve the manuscript significantly.
-
->General remarks:
-- I understand that the implementation of this work in OpenFOAM has demanded a lot of effort but could you please elaborate on what is new/novel in this work?
-An issue that has not been attacked/discovered/raised before or a numerical implementation which overcomes a known problem.
-
-As noted in the "Novelty" file, this work presents some results that have either never been simulated or reported before:
-
-1. Validating with Reynolds numbers more representative of a full-scale turbine, and two orders of magnitude higher than those used by Shamsoddin and Porte-Agel.
-2. Reporting performance predictions, which has also not yet been done for a VAT in an ALM simulation.
-3. Embedding the ALM inside of a RANS model for additional computational affordability.
-
-The reviewer is correct that we had not sufficiently described this novelty in
-the paper itself, and we have updated the introduction accordingly.
-
->- Minor comment: I'd rather use nondimentional units for variables such as time step
-
-Figure 6 does use a time step that is technically nondimensional (time steps per rotor revolution).
-
->Pages 1-6: Again, very good introduction and gathering of introductory note,
-but in my opinion, it should be shortened for a journal.
-
-We agree and have shortened the introduction significantly.
-
-TODO: Shorten introduction
-
->Page 7, Fig 3: It is taking a lot of room, perhaps you can make a smaller
-schematic
-
-We have made this figure smaller in the manuscript, a change that may be
-retained by the typesetter.
-
->Page 9 fig 6: The Cp figure is an interesting one. But first of all, but is
-there a particular reason why you have not shown Ct next to it (it's more common
-this way than having Cd). Also, would it not be possible to have LES results on
-top of both figures 6 and 11 for the cases you have simulated?
-
-We use the overal rotor drag coefficient nomenclature ($C_D$) since this was
-adopted by Paraschivoiu (2001). As noted, this is commonly called the thrust
-coefficient for horizontal-axis turbines. We have added a note to the manuscript
-to clarify this.
-
-We have not plotted the LES results, since we only used LES for one tip speed
-ratio, and for one turbine---it was mainly employed to assess wake predictions.
-
->Page 11, Line 42: I'm not quite sure if what you have stated about Smagorinsky
-model is correct. It certainly is very easy to implement and gives nice results
-in some cases, but we are all aware of its deficiencies. We set a fix Cs value
-which means we add numerical dissipation to the whole computatiponal domain, but
-this is not really accurate when dealing with cases with strong directional
-homogeneity and anisotropy. Why didn't you choose a dynamic Smagorinsky or even
-dynamic Lagrangian model? Besides, what is the Cs value you have used?
-
-We are not trying to claim that the Smagorinsky model is the most accurate or
-"the standard" model, but that is has been used in many other ALM papers, and
-therefore a good baseline for comparison. The wording of the text has been
-adjusted to reflect the notion that we are interested in this comparison to
-baseline for turbine ALM modeling, and are by no means implying that this SGS
-model is the most accurate or appropriate for this case. Certainly testing
-additional SGS models will be part of future work, as is mentioned in the
-conclusions.
-
-The $C_S$ value used was OpenFOAM's default value of approximately 0.17. This
-detail has been added to the text.
-
->Page 13, Fig 15: Following my last comment, figure to the right may very well
-explain that the Smagorinsly model has been over dissipative thereby hampering
-turbulent fluctuations which may have otherwise resulted in higher TKEs.
-
-This is a valid point, and we have added a comment to the text accordingly. It
-is important to note, however, that since we don't have near-wall boundary
-layers, it may be that turbulence is only generated by larger-scale shear and
-vortex breakdown, rather than by boundary layer separation or transition, which
-would occur with a body-fitted mesh.
-
->Page 14, Line 27: Matching mean velocities is usually rather straightforward
-(in non-presudospectral codes, sometimes even shutting down the whole SGS
-modeling wouln'd make a difference as long as no solid walls are involved). But
-higher order moments such as Reynolds stress tensor components are usually more
-sensitive. Perhaps you can go more in depth with that part -- possibly compare
-your results with other experiments or CFD simulations. Besides, please try and
-overcome the relatively large discrepancies seen in figures 6 (high TSR), 9, 11.

### 2018-03-29 reviews/reviewer2-response.md
Update review responses

--- 2016-08-27 reviews/reviewer2-response.md
@@
 
->1. Pag1. Please insert the list of Symbols since there are a lot of
-abbreviations in the text, a quick reference could be useful.
-
-We have updated the manuscript with a nomenclature section.
-
->2. Pag. Pag3. Sect. 2. The parameter reduced frequency "k" is the key parameter
-for unsteadiness considerations concerning VAT: some more detailed comments
-should be given concerning the actual value of this parameter in the case study
-numerically modelled. At the same time a comment should be given concerning the
-ALM capabilities (and/or future improvements) in terms of unsteady aerodynamics.
-
->3. Pag4. Sect.4.  Is the force smoothing with the Gaussian function really
-necessary? In P.-E. Réthoré and N. N. Sørensen, A discrete force allocation
-algorithm, Wind Energy, 2012. Rethore and Sorensen showed how this spatial force
-filter could be avoided with modified Rhie–Chow interpolation algorithm.
-
-In our experience, especially with the finer LES grids, the force smoothing is
-necessary. Without smoothing, the blade loading becomes quite "noisy" due to the
-subsequent instability from sharper momentum source gradients. In this study we
-did not investigate changes to the numerical algorithms used to solve the flow
-field, but certainly future research may prove the Gaussian width can be
-narrowed or eliminated if, for example, modified Rhie--Chow interpolation is
-implemented in OpenFOAM.
-
->4. Pag6. Sect.7. Does the authors took into consideration that the ALM could be
-in theory able to catch the Tip vortex effect without a correction model? i.e.
-the vortex generated by the lifting of the blade line is already influencing the
-angle of attack on the blade.
-
-While tip vortices are produced by the ALM, and these do affect the local angle
-of attack, we are not sure how easy it would be to create a ALM that resolves
-tip effects totally accurately without any correction model. The issue may be
-that since we do not enforce the no peneration BC through the blade, the tip
-vortex can leave the blade too early, which causes incorrect angles of attack
-close to the tip.
-
->5. Pag7. Sect.8-9 . Add a sketch of the domain boundaries with the position of
-the turbine, and the flow direction. Figure 3 is not very effective in terms of
-quantitative information on the actual implementation of the code.
-
->6. Sect. 8-9 The details on the domain boundaries of the implemented case study
-should be an opportunity for clarify if any blockage effect was present in the
-reference experiments and if the same blockage conditions were adopted in the
-numerical implementation.
-
->7. Pag12. Sect.9.4 "The near-wake's mean velocity field at x/D=1 is shown in
-Figure 12..." should be actually Figure 13
+We thank the reviewer for his or her insightful comments, and feel that these
+have helped improve the manuscript significantly.

### 2018-03-29 paper.tex
Add revised preprint date
Fix typo
Switch to Energies format

@@
-% wedoc.tex V1.0, 17 June 2010
-
-\documentclass[times]{weauth}
-
-\usepackage{moreverb}
-
-\usepackage{lineno}
-
-\usepackage{nomencl}
-\makenomenclature
-
-\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+%  LaTeX support: latex@mdpi.com
+%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).
+
+% You need to save the "mdpi.cls" and "mdpi.bst" files into the same folder as this template file.
+
+%=================================================================
+\documentclass[journal,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
+
+% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'.
+The document class line would be, e.g.
+\documentclass[preprints,article,accept,moreauthors,pdftex,10pt,a4paper]{mdpi}.
+This is especially recommended for submission to arXiv, where line numbers should be removed before posting.
+For preprints.org, the editorial staff will make this change immediately prior to posting.
 
@@
 
-\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
-T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-
-\def\volumeyear{2016}
-
-\begin{document}
-
-\runningheads{P.~Bachant \emph{et al.}}{Actuator line modeling of vertical-axis
-turbines}
-
-\articletype{RESEARCH ARTICLE PREPRINT (submitted May 2016)}
-
-\title{Actuator line modeling of vertical-axis turbines}
-
-\author{P.~Bachant\textsuperscript{1}, A.~Goude\textsuperscript{2}, and
-M.~Wosnik\textsuperscript{1}}
-
-\address{\textsuperscript{1}Center for Ocean Renewable Energy,
-University of New Hampshire, 24
-Colovos Rd., Durham, NH, 03824, USA \\
-\textsuperscript{2}Department of Engineering Sciences,
-Division of Electricity, Uppsala University, Uppsala 751 21, Sweden}
-
-\corraddr{P.~Bachant, Chase Ocean Engineering Laboratory, University of New
-Hampshire, 24 Colovos Rd., Durham, NH, 03824, USA. \\
-E-mail: pwbachant@gmail.com}
-
-\linenumbers
-
-\begin{abstract}
+%
+%--------------------
+% Class Options:
+%--------------------
+% journal
+%----------
+% Choose between the following MDPI journals:
+% actuators, addictions, admsci, aerospace, agriculture, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci, beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci, buildings, carbon, cancers, catalysts, cells, ceramics, challenges, chemengineering, chemosensors, children, chromatography, cleantechnol, climate, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, cybersecurity, data, dentistry, designs, diagnostics, diseases, diversity, econometrics, economies, education, electrochem, electrochemistry, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, galaxies, games, gastrointestdisord, gels, genealogy, genes, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, informatics, information, infrastructures, inorganics, insects, instruments, ijerph, ijfs, ijms, ijgi, ijtpp, inventions, j, jcdd, jcm, jcs, jdb, jfb, jfmk, jimaging, jof, jintelligence, jlpea, jmmp, jmse, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, mti, medsci, medicina, medicines, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, nanomaterials, ncrna, neonatalscreening, neuroglia, nitrogen, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, plants, plasma, polymers, polysaccharides, preprints, proceedings, processes, proteomes, publications, quaternary, qubs, reactions, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, scipharm, sensors, separations, sexes, sinusitis, socsci, societies, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem
+%---------
+% article
+%---------
+% The default type of manuscript is article, but can be replaced by:
+% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, correction, conferenceproceedings, conferencereport, expressionofconcern, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, opinion, obituary, projectreport, reply, reprint, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
+% supfile = supplementary materials
+% protocol: If you are preparing a "Protocol" paper, please refer to http://www.mdpi.com/journal/mps/instructions for details on its expected structure and content.
+%----------
+% submit
+%----------
+% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted.
+This will only make changes to the frontpage (e.g.
+the logo of the journal will get visible), the headings, and the copyright information.
+Also, line numbering will be removed.
+Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
+%------------------
+% moreauthors
+%------------------
+% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
+%---------
+% pdftex
+%---------
+% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.
+
+%=================================================================
+\firstpage{1}
+\makeatletter
+\setcounter{page}{\@firstpage}
+\makeatother
+\pubvolume{xx}
+\issuenum{1}
+\articlenumber{1}
+\pubyear{2018}
+\copyrightyear{2018}
+%\externaleditor{Academic Editor: name}
+\history{
+    Preprint uploaded XXXXXX
+    Received: date;
+    Accepted: date;
+    Published: date
+}
+
+%\updates{yes} % If there is an update available, un-comment this line
+
+%------------------------------------------------------------------
+% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
+%\pdfoutput=1
+
+%=================================================================
+% Add packages and commands here.
+The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz, siunitx
+
+%=================================================================
+%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
+%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).
+
+%=================================================================
+% Full title of the paper (Capitalized)
+\Title{Actuator line modeling of vertical-axis turbines}
+
+% Author Orchid ID: enter ID or remove command
+\newcommand{\orcidauthorA}{0000-0003-4039-6954} % Add \orcidA{} behind the author's name
+%\newcommand{\orcidauthorB}{0000-0000-000-000X} % Add \orcidB{} behind the author's name
+
+% Authors, for the paper (add full first names)
+\Author{
+    Peter Bachant $^{1,\dagger,\ddagger}$\orcidA{},
+    Anders Goude $^{2,\ddagger}$
+    and Martin Wosnik $^{1,}$*
+}
+
+% Authors, for metadata in PDF
+\AuthorNames{Peter Bachant, Anders Goude and Martin Wosnik}
+
+% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
+\address{%
+$^{1}$ \quad University of New Hampshire, USA\\
+$^{2}$ \quad Uppsala University, Sweden}
+
+% Contact information of the corresponding author
+\corres{Correspondence: pwbachant@gmail.com}
+
+% Current address and/or shared authorship
+% \firstnote{Current address: Affiliation 3}
+\secondnote{These authors contributed equally to this work.}
+% The commands \thirdnote{} till \eighthnote{} are available for further notes
+
+% Simple summary
+%\simplesumm{}
+
+% Abstract (Do not insert blank lines, i.e. \\)
+\abstract{
     To bridge the gap between high and low fidelity numerical modeling tools for
@@
     investigation of optimal subgrid scale modeling.
-\end{abstract}
-
-\keywords{wind turbine; VAWT; cross-flow turbine; blade element theory; dynamic
+}
+
+% Keywords
+\keyword{wind turbine; VAWT; cross-flow turbine; blade element theory; dynamic
 stall; lifting line}
 
-\maketitle
-
-\printnomenclature
+% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
+%\PACS{J0101}
+%\MSC{}
+%\JEL{}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Only for the journal Applied Sciences:
+%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Only for the journal Data:
+%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately.
+If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal.
+In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}
+
+%\datasetlicense{license under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Only for the journal Toxins
+%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}
+
+%\setcounter{secnumdepth}{4}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Only for the journal Gels: Please place the Experimental Section after the Conclusions
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%The order of the section titles is: Introduction, Materials and Methods, Results, Discussion, Conclusions for these journals: aerospace,algorithms,antibodies,antioxidants,atmosphere,axioms,biomedicines,carbon,crystals,designs,diagnostics,environments,fermentation,fluids,forests,fractalfract,informatics,information,inventions,jfmk,jrfm,lubricants,neonatalscreening,neuroglia,particles,pharmaceutics,polymers,processes,technologies,viruses,vision
 
@@
 
-\nomenclature{CFT}{Cross-flow turbine}
-\nomenclature{VAT}{Vertical-axis turbine}
-\nomenclature{AFT}{Axial-flow turbine}
-\nomenclature{HAT}{Horizontal-axis turbine}
+% \nomenclature{CFT}{Cross-flow turbine}
+% \nomenclature{VAT}{Vertical-axis turbine}
+% \nomenclature{AFT}{Axial-flow turbine}
+% \nomenclature{HAT}{Horizontal-axis turbine}
 
@@
 
-\nomenclature{RANS}{Reynolds-averaged Navier--Stokes}
+% \nomenclature{RANS}{Reynolds-averaged Navier--Stokes}
 
@@
 
-\nomenclature{ALM}{Actuator line model}
-\nomenclature{LES}{Large eddy simulation}
+% \nomenclature{ALM}{Actuator line model}
+% \nomenclature{LES}{Large eddy simulation}
 
@@
 
-\nomenclature{$c$}{Blade chord length}
-\nomenclature{$R$}{Maximum rotor radius}
-\nomenclature{$Re_D$}{Rotor diameter Reynolds number}
-\nomenclature{DOE}{US Department of Energy}
-\nomenclature{SNL}{Sandia National Laboratories}
+% \nomenclature{$c$}{Blade chord length}
+% \nomenclature{$R$}{Maximum rotor radius}
+% \nomenclature{$Re_D$}{Rotor diameter Reynolds number}
+% \nomenclature{DOE}{US Department of Energy}
+% \nomenclature{SNL}{Sandia National Laboratories}
 
@@
 
-\nomenclature{$\vec{U}_\mathrm{rel}$}{Blade element local relative velocity}
-\nomenclature{$\alpha$}{Blade element angle of attack}
-\nomenclature{$\omega$}{Rotor angular velocity}
-\nomenclature{$r$}{Blade element radius}
-\nomenclature{$\vec{U}_\mathrm{in}$}{Blade element local inflow velocity}
+% \nomenclature{$\vec{U}_\mathrm{rel}$}{Blade element local relative velocity}
+% \nomenclature{$\alpha$}{Blade element angle of attack}
+% \nomenclature{$\omega$}{Rotor angular velocity}
+% \nomenclature{$r$}{Blade element radius}
+% \nomenclature{$\vec{U}_\mathrm{in}$}{Blade element local inflow velocity}
 
@@
 
-\nomenclature{$U_\infty$}{Free stream velocity}
+% \nomenclature{$U_\infty$}{Free stream velocity}
 
@@
 
-\nomenclature{$\lambda$}{Rotor tip speed ratio}
-\nomenclature{$k$}{Reduced frequency or turbulence kinetic energy}
+% \nomenclature{$\lambda$}{Rotor tip speed ratio}
+% \nomenclature{$k$}{Reduced frequency or turbulence kinetic energy}
 
@@
 
-\nomenclature{$C_l$}{Blade element lift coefficient}
-\nomenclature{$C_d$}{Blade element drag coefficient}
-\nomenclature{$C_m$}{Blade element moment coefficient}
+% \nomenclature{$C_l$}{Blade element lift coefficient}
+% \nomenclature{$C_d$}{Blade element drag coefficient}
+% \nomenclature{$C_m$}{Blade element moment coefficient}
 
@@
 
-\nomenclature{$Re_c$}{Blade chord Reynolds number}
+% \nomenclature{$Re_c$}{Blade chord Reynolds number}
 
@@
     cylindrical region represents the swept area of the UNH-RVAT rotor. Note
-    that increased mesh refinrement was used for the LES simulations.}
+    that increased mesh refinement was used for the LES simulations.}
 
@@
 
-\nomenclature{$\epsilon$}{Gaussion force smoothing function width}
+% \nomenclature{$\epsilon$}{Gaussion force smoothing function width}
 
@@
 
-\nomenclature{$C_P$}{Rotor power coefficient}
-\nomenclature{$C_D$}{Overall rotor drag (a.k.a. thrust) coefficient}
-\nomenclature{$D$}{Rotor diameter}
+% \nomenclature{$C_P$}{Rotor power coefficient}
+% \nomenclature{$C_D$}{Overall rotor drag (a.k.a. thrust) coefficient}
+% \nomenclature{$D$}{Rotor diameter}
 
@@
 
-\nomenclature{$H$}{Rotor height}
+% \nomenclature{$H$}{Rotor height}
 
@@
 
-\nomenclature{$\lambda_0$}{Tip speed ratio of maximum power coefficient}
+% \nomenclature{$\lambda_0$}{Tip speed ratio of maximum power coefficient}
 
@@
 
-\acks The authors would like to acknowledge funding through a National Science
+\acknowledgments{
+    The authors would like to acknowledge funding through a National Science
     Foundation CAREER award (principal investigator Martin Wosnik, NSF 1150797,
@@
     University of New Hampshire.
-
-\bibliographystyle{wileyj}
+}
+
+\authorcontributions{
+    P.B. wrote the computer code, ran the simulations, post-processed the
+    results, and wrote the paper.
+    A.G. contributed to the computer code and the manuscript.
+    M.W. contributed to writing and editing of the manuscript.
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\conflictsofinterest{The authors declare no conflicts of interest.}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.
+
+% The following MDPI journals use author-date citation: Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws, Religions, Risks, Social Sciences.
+For those journals, please follow the formatting guidelines on http://www.mdpi.com/authors/references
+% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
+% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p.
+328; \citeyear{ref-journal-3b}, p.475).
+This produces: Wong (1999, p.
+328; 2000, p.
+475)
+
+%=====================================
+% References, variant B: external bibliography
+%=====================================
+\externalbibliography{yes}
 \bibliography{library}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{document}

### 2018-03-31 paper.tex
Add 'the'
Update OF and TF versions

@@
 turbines, as well as standalone actuator lines. The actuator line model
-developed here, dubbed turbinesFoam \cite{Bachant2016-turbinesFoam-v0.0.7}, was
-also written as an extension library for OpenFOAM (version 3.0.x), using its
+developed here, dubbed turbinesFoam \cite{Bachant2018-turbinesFoam-v0.0.8}, was
+also written as an extension library for OpenFOAM (v5.0), using its
 \texttt{fvOptions} framework for adding source terms to equations at runtime.
@@
     A.G. contributed to the computer code and the manuscript.
-    M.W. contributed to writing and editing of the manuscript.
+    M.W. contributed to the writing and editing of the manuscript.
 }

### 2018-03-31 paper.tex
Fix journal name and correspondence

@@
 %=================================================================
-\documentclass[journal,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
+\documentclass[energies,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
 
@@
 \Author{
-    Peter Bachant $^{1,\dagger,\ddagger}$\orcidA{},
-    Anders Goude $^{2,\ddagger}$
+    Peter Bachant $^{1}$\orcidA{},
+    Anders Goude $^{2,}$
     and Martin Wosnik $^{1,}$*
@@
 % Contact information of the corresponding author
-\corres{Correspondence: pwbachant@gmail.com}
+\corres{Correspondence: martin.wosnik@unh.edu}
 
@@
 % \firstnote{Current address: Affiliation 3}
-\secondnote{These authors contributed equally to this work.}
+% \secondnote{These authors contributed equally to this work.}
 % The commands \thirdnote{} till \eighthnote{} are available for further notes

### 2018-04-08 paper.tex
Remove comma from superscript
Update case file citations

@@
     Peter Bachant $^{1}$\orcidA{},
-    Anders Goude $^{2,}$
+    Anders Goude $^{2}$
     and Martin Wosnik $^{1,}$*
@@
 force projection width $\epsilon$. Case files for running all the simulations
-presented here with OpenFOAM 3.0.x are available from
-\cite{Bachant2016-UNH-RVAT-turbinesFoam-v1.0.0,
-Bachant2016-RM2-turbinesFoam-v1.0.0,
-Bachant2016-UNH-RVAT-turbinesFoam-v1.0.0-LES}.
+presented here with OpenFOAM v5.0 are available from
+\cite{Bachant2018-UNH-RVAT-turbinesFoam-v1.1.0,
+Bachant2018-RM2-turbinesFoam-v1.1.0}.
 

### 2018-04-08 scripts/zip.py
Fix zip script

@@ -26,7 +26,8 @@ for fig in figures:
         if f in fig:
             figures.remove(fig)
 
-files = ["paper.tex", "paper.bbl", "weauth.cls", "wileyj.bst"]
+files = ["paper.tex", "paper.bbl", "logo-ccby.eps", "logo-mdpi.eps",
+         "logo-updates.pdf", "chicago2.bst", "mdpi.cls", "mdpi.bst"]
 
 with ZipFile("archive/paper-{}.zip".format(version), "w") as f:
     for fig in figures:

### 2018-04-13 scripts/zip.py
Zip converted PDF

@@ -26,11 +26,15 @@ for fig in figures:
         if f in fig:
             figures.remove(fig)
 
-files = ["paper.tex", "paper.bbl", "logo-ccby.eps", "logo-mdpi.eps",
+files = ["paper.tex", "paper.bbl", "logo-mdpi-eps-converted-to.pdf",
          "logo-updates.pdf", "chicago2.bst", "mdpi.cls", "mdpi.bst"]
 
 with ZipFile("archive/paper-{}.zip".format(version), "w") as f:
     for fig in figures:
         f.write(fig)
     for file in files:
-        f.write(file)
+        if "-eps-converted-to" in file:
+            arcname = file.replace("-eps-converted-to", "")
+        else:
+            arcname = None
+        f.write(file, arcname=arcname)

### 2018-05-03 paper.tex
Uncomment pdfoutput for arXiv
Set as preprint
Add 2017 Boudreau article; resolves #39

@@
 % The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
-%\pdfoutput=1
+\pdfoutput=1
 
@@
 on the turbulence model
-applied~\cite{Bachant2016-BR-CFD,Lam2016,Alaimo2015,Boudreau2015,Marsh2015}.
+applied~\cite{Bachant2016-BR-CFD,Lam2016,Alaimo2015,Boudreau2017,Marsh2015}.
 However, 3-D blade-resolved RANS presents a huge computational expense---on the

### 2018-05-04 reviews/reviewer1-response.md
Start responses to reviewer 1

@@
 to improve the manuscript significantly.
+
+>L. 111. Can you define lambda0?
+
+This definition has been added to the theory section of the manuscript.
+
+>Fig. 4 and 5. Can you include a line with the experimental value of Cp to appreciate how the model converges to it?
+
+This is a reasonable suggestion. However, we feel that this may confuse the
+readers with the difference between checking the independence of the model
+parameters with the validation reported later.
+
+>L. 228. Is the mesh uniform in the three directions? It can be quite relevant to state how the vertical resolution of the turbine is.
+
+Yes, the mesh was uniform in all directions. This detail has been added to the
+manuscript.
+
+>L. 233. Define zero-subscript quantities.
+
+Thanks for catching this. Is it now noted that $\Delta t_0$ is the time step
+chosed to run the final simulations.
+
+>L.
+268.
+The drag and thrust can be considered differently to CFTs.
+Although for the whole rotor the drag can be considered as the thrust, i.e.
+as typically in a cylinder, it can bring to confusion when drag coefficient calculations on a blade are performed, as these are dependent on x- and y- forces, and not just on the former.
+
+This is a valid point. Both are defined in different coordinate systems.
+The clarify, we have attempted to separate blade element
+and rotor drag using lowercase and uppercase subscripts, respectively.
+
+>Fig.
+6.
+The results showed in this figure are very promising.
+The overestimation at high tip speed ratios have been commonly found in literature as the authors pointed out.
+However, the reviewer questions if such effects can arise from the use of k-epsilon turbulence model.
+This model is usually not considered when wall effects are relevant in the analysed flow.
+Thus, the k-omega model is more used as in the author’s work from reference 10.
+Why did you not consider k-omega which would allow to a closer comparison to those blade-resolved results?
+At high tip speed ratios, the blades might undergo quite light dynamic stall, and maybe your DS model is over-predicting this value and triggering the Cd and Cp curves.
+Nevertheless, this is an almost unavoidable consequence of actuator line models.
+The reviewer suggest that the turbulence model may contribute more than expected by the authors in such Cp prediction.
+
+We have doubts that the turbulence model is causing much difference, since
+near actuator line elements, there is not a wall model. There could be an
+effect from a difference in turbulence production in the model from shear
+caused by the momentum source. We hypothesize that these discrepancies arise
+from secondary effects, or potentially deficiencies in the input foil data,
+both of which affect the simpler momentum and vortex models.
+
+>Fig.
+11.
+The results of power coefficient from the second case show that the model is quite case dependent.
+The results show that the model is not able to represent the physics about this turbine and thus the actual accuracy of the ALM cannot be verified without any additional figure that shows how the flow field is.
+Can you provide transversal plane contours showing the actual flow field generated by the ALM?
+Could this inaccuracy be again due to the k-epsilon model?
+
+We have added a snapshot of the internal flow field and force vectors to give
+and idea of the level of detail, or lack thereof, regarding local flow near
+the blades. Unfortunately there is no mechanism for the ALM to generate flow
+features like the pair of counter rotating vortices induced by dynamic stall.
+This is a compromise inherent to the ALM, which we have clarified. It is
+advantageous when attempting to simulate array-scale flows, but not so much
+for individual device optimization, despite our validation efforts focusing
+on the latter.
+
+>Section 9.4. The presented LES results are quite disappointing due to the large differences against the RANS results which were very good. The reviewer has few concerns about this section:
+>
+Can you specify more details about the number of processors used?
+It is true that LES demands more computational resources than RANS but with the meshes used for the ALM it should be affordable to run LES.
+
+We ran the LES simulations on 4 processors, a note that has been added to the
+paper. It was still relatively affordable compared to the blade-resolved RANS.
+We have noted a 2 order of magnitude drop in computational cost.
+
+>Overall LES results are poor.
+The predicted Cp dropped to 20% which seems to be derived from the resolution of the flow field instead of the ALM itself.
+While a good agreement is found for the velocity wake as seen in Fig.
+13 and 15(left), the turbulent kinetic energy results show an unexpected over damping of the velocity fluctuations.
+The use of LES is herein motivated to achieve a better resolution of the turbulent fluctuations by explicitly resolving the large-scale flow structures, and in the present manuscript the TKE has dropped dramatically.
+Fig.
+14 and 15 (right) show a lack of flow unsteadiness attributed by the authors to the influence of the SGS model.
+The spatial discretisation is not described and it may be possible that depending on it, the scheme introduced an extra damping typically used in implicit LES wherein no SGS is used.
+Nevertheless, the reviewer would recommend to double-check the kinetic energy results as is quite confusing that the velocity wake prediction is greatly improved while the velocity fluctuations are almost gone.
+
+One important note is that the SGS TKE is not tracked by the Smagorinsky
+model, so there will be a lack of energy there.
+
+It is not likely that the numerical schemes are dissipating the energy, since
+all 2nd order schemes are used.
+
+TODO: Check on TKE calculation from LES model and finish this.
+
+>For further validation, the authors may consider to plot RANS vs LES at closer distance to the turbine, e.g.
+x/D=0.5 and study how the turbulent kinetic energy decay is in both models, or a transversal plane showing the generation of the vortices.

### 2018-05-04 paper.tex
Start responses to reviewer 1

@@
 oscillations at $\lambda=6$. For the two turbines studied here, reduced
-frequencies at $\lambda_0$ are 0.27 and 0.17 (based on root chord length) for
+frequencies at the tip speed ratio of peak power coefficient, $\lambda_0$,
+are 0.27 and 0.17 (based on root chord length) for
 the UNH-RVAT and RM2, respectively. Therefore, both cases will be inherently
@@
 constant, and the resolution was scaled proportional to the number of cells in
-the $x$-direction $N_x$ for the base hexahedral mesh. Both models displayed low
+the $x$-direction $N_x$ for the base hexahedral mesh, which was essentially
+uniform in resolution in all directions. Both models displayed low
 sensitivity to the number of time steps per revolution. Spatial grid dependence,
@@
 Final spatial grid resolutions were chosen as $N_x=48$ for both the UNH-RVAT and
-RM2 cases. Time steps were chosen as $\Delta t = 0.01$ and $\Delta t = 0.005$
+RM2 cases. Time steps were chosen as $\Delta t_0 = 0.01$ and
+$\Delta t_0 = 0.005$
 seconds for the UNH-RVAT and RM2 respectively, which correspond to approximately
@@
 performance curves, the number of steps per revolution was kept constant, i.e.,
-the time step was adjusted to $\Delta t = \Delta t_0 \lambda_0 / \lambda$.
+the time step was adjusted to $\Delta t = \Delta t_0 \lambda_0 / \lambda$,
+where $\lambda_0$ was the tip speed ratio roughly corresponding to peak $C_P$
+as observed in experiments, which was used for verifying the numerical
+sensitivity to spatial and temporal resolution.
 
@@
 
+The internal rotor flow about the horizontal center plane, along with
+the directions and relative magnitudes of the actuator line force vectors
+are shown in Figure~\ref{fig:RM2-internal}.
+It can be seen how flow features like dynamic stall vortices are not
+captured with the ALM, despite forces themselves being modulated by the
+dynamic stall model. This combined with the diffusive nature of the
+RANS model helps explain discrepancies in predicted power coefficient.
+With the computational affordability of the ALM coupled with RANS comes a
+compromise in flow detail, which should be acceptable for array-level
+simulation, but for individual device design a blade-resolved
+simulation would be more advisable.
+
+\begin{figure}
+    \centering
+
+    \includegraphics[width=0.8\textwidth]{rm2-vel-force-6-seconds}
+
+    \caption{Velocity magnitude and actuator line force field vectors for the
+    RM2 RANS simulation at 6 seconds simulated time.}
+
+    \label{fig:RM2-internal}
+\end{figure}
+
 The mean near-wake structure for the RM2 looked qualitatively similar to the
@@
 With these resolutions computation times were $O(10)$ CPU hours per second of
-simulated time, which is approximately two orders of magnitude lower than for
+simulated time (on 4 processors),
+which was approximately two orders of magnitude lower than for
 blade-resolved RANS.

### 2018-05-05 reviews/reviewer1-response.md
Finish addressing reviewer 1 comments

@@
 paper. It was still relatively affordable compared to the blade-resolved RANS.
-We have noted a 2 order of magnitude drop in computational cost.
+We have noted a 2 order of magnitude drop in computational cost as well.
 
@@
 Nevertheless, the reviewer would recommend to double-check the kinetic energy results as is quite confusing that the velocity wake prediction is greatly improved while the velocity fluctuations are almost gone.
-
-One important note is that the SGS TKE is not tracked by the Smagorinsky
-model, so there will be a lack of energy there.
+>
+>For further validation, the authors may consider to plot RANS vs LES at closer distance to the turbine, e.g.
+x/D=0.5 and study how the turbulent kinetic energy decay is in both models, or a transversal plane showing the generation of the vortices.
 
 It is not likely that the numerical schemes are dissipating the energy, since
-all 2nd order schemes are used.
-
-TODO: Check on TKE calculation from LES model and finish this.
-
->For further validation, the authors may consider to plot RANS vs LES at closer distance to the turbine, e.g.
-x/D=0.5 and study how the turbulent kinetic energy decay is in both models, or a transversal plane showing the generation of the vortices.
+all 2nd order schemes are used. We went back and looked at contours of the
+resolved and SGS TKE, and interestingly, the total TKE is significantly
+increased after the transition in the grid resolution in the near-wake.
+This does imply that the vortices created by the actuator lines do remain
+artificially stable, and the grid transition helps break them down and generate
+resaonably levels of TKE. This would not be much of an issue in array
+simulations with spacings greater than a few diameters. A note has been added
+to the manuscript. We considered adding the contour plots, but the paper is
+starting to get quite long. We can include them if they will contribute to
+the understanding, however.

### 2018-05-05 paper.tex
Finish addressing reviewer 1 comments

@@
 RANS, LES is more able to predict the turbulence generated by the blade tip
-vortex shedding and dynamic stall effects, though the overall level of
-unsteadiness generated was much lower, especially on the $+y$ side of the rotor.
+vortex shedding and dynamic stall effects, though the total turbulence kinetic
+energy was lower, especially on the $+y$ side of the rotor.
 This is likely a consequence of the SGS modeling, where the vortical structures
@@
 generated in the large eddy simulation was approximately an order of magnitude
-too low. Once again this was likely a result of the subgrid-scale modeling and
-its effect on the stability of vortex structures, or its over-dissipative nature
-removing energy from the flow before it is advected to the plane of interest.
+too low.
+Note that $k$ (both resolved and SGS) does increase significantly downstream
+after the transition in grid resolution, which implies the grid can help
+compensate for the artificial stability of the vortex structures caused by
+the Smagorinsky model.
+It is hypothesized that the turbulence kinetic energy would better match
+experimental data taken further downstream, meaning the model should perform
+well for array simulations with spacing as close as $2D$.
 

### 2018-05-05 paper.tex
Address reviewer 2 comments

@@
 \end{equation}
-which assumes the free stream velocity is constant. Unsteady effects begin to
+which assumes the free stream velocity is constant.
+Unsteady effects begin to
 become significant for $k > 0.05$, and can become dominant for $k \ge 0.2$.
+Note $\omega$ in the reduced frequency equation refers to the frequency of
+angle of attack oscillation, which is normally sinusoidal in pitching foil
+experiments, but is analogous to cross-flow turbine blade kinematics, i.e.,
+the angle of attack cycle encounted by a CFT blade has a period equal to one
+turbine revolution.
 
@@
 For a VAT or CFT reduced frequency can be reformulated in terms of the tip speed
-ratio as \begin{equation} k = \frac{\lambda c}{2R}, \end{equation} which is then
+ratio as
+\begin{equation}
+    k = \frac{\lambda c}{2R},
+\end{equation}
+which is then
 also a function of solidity or chord-to-radius ratio $c/R$. As an example, a

### 2018-05-05 paper.tex
Address reviewer 3 comments

@@
 Before the dynamic stall subroutine is executed, the static profile data for
-each element is interpolated linearly based on local chord Reynolds number. The
+each element is interpolated linearly based on local chord Reynolds number,
+which is calculated based on the detected inflow velocity for each actuator
+line element and the simulation's kinematic viscosity. The
 profile data characteristics---static stall angle, zero-lift drag coefficient,
@@
 active support community around itself. Though SOWFA is also open-source, its
-mostly procedural style would have required significant effort and duplicate
-code to adapt for cross-flow turbines. Thus, a new and more general ALM library
-was developed from the ground up that could model both cross- and axial-flow
+architecture is focused on simulating horizontal-axis wind turbine arrays
+in the atmospheric boundary layer and would have required significant effort to
+adapt for cross-flow turbines. Thus, a new and more general ALM library
+was developed that could model both cross- and axial-flow
 turbines, as well as standalone actuator lines. The actuator line model
-developed here, dubbed turbinesFoam \cite{Bachant2018-turbinesFoam-v0.0.8}, was
+developed here, turbinesFoam \cite{Bachant2018-turbinesFoam-v0.0.8}, was
 also written as an extension library for OpenFOAM (v5.0), using its

### 2018-05-05 paper.tex
Add OpenFOAM to abstract, intro, keywords per reviewer 4

@@
     Reynolds-averaged Navier--Stokes (RANS) and Smagorinsky large eddy
-    simulation (LES) turbulence models were tested. The RANS models were able to
-    be run on coarse grids while still providing good convergence behavior in
+    simulation (LES) turbulence models were tested using the open-source
+    OpenFOAM computational fluid dynamics framework. The RANS models were able
+    to be run on coarse grids while still providing good convergence behavior in
     terms of the mean power coefficient, and also approximately four orders of
@@
 \keyword{wind turbine; VAWT; cross-flow turbine; blade element theory; dynamic
-stall; lifting line}
+stall; lifting line; OpenFOAM}
 
@@
 model inside LES and unsteady RANS simulations, the latter of which has not yet
-been reported in the literature. This model was validated against experimental
-data for the high solidity ($c/R=0.28$) UNH Reference Vertical-Axis Turbine
-(UNH-RVAT) and the medium solidity ($c/R=0.07$--$0.12$) US Department of
-Energy/Sandia National Labs Reference Model 2 (RM2) cross-flow turbine (at 1:6
-scale), both of which are shown in Figure~\ref{fig:turbines}. Validation
-datasets for both turbines were taken from \cite{Bachant2016-RVAT-Re-dep} and
+been reported in the literature. This model was implemented within the
+open-source OpenFOAM CFD framework and validated against experimental data for
+the high solidity ($c/R=0.28$) UNH Reference Vertical-Axis Turbine (UNH-RVAT)
+and the medium solidity ($c/R=0.07$--$0.12$) US Department of Energy/Sandia
+National Labs Reference Model 2 (RM2) cross-flow turbine (at 1:6 scale), both of
+which are shown in Figure~\ref{fig:turbines}. Validation datasets for both
+turbines were taken from \cite{Bachant2016-RVAT-Re-dep} and
 \cite{Bachant2016-RM2-data}, respectively. The turbines were modeled at $Re_D

### 2018-05-05 paper.tex
Don't --> do not

@@
 abandoned in favor of horizontal-axis (or axial-flow) turbines, as they
-generally are more efficient and don't encounter the high levels of fatigue
+generally are more efficient and do not encounter the high levels of fatigue
 loading that VATs do. Today, however, there is renewed interest in VATs for

### 2018-05-06 paper.tex
Finish addressing reviewer 4

@@
 
-After the force on the ALE from the flow is calculated, it is then projected
+After the force on the actuator line element from the flow is calculated,
+it is then projected
 back onto the flow field as a source term in the momentum equation. To avoid
@@
 resolved response of the attached flow loading and effects of flow acceleration,
-also know as added mass. Furthermore, the angles of attack encountered by a CFT
+also known as added mass. Furthermore, the angles of attack encountered by a CFT
 blade will often be high enough to encounter dynamic stall (DS). It is therefore
@@
 However this correction depends on rotor parameters---tip speed ratio, number of
-blades, element radius, tip flow angle---that do no necessarily translate
+blades, element radius, tip flow angle---that do not necessarily translate
 directly to the geometry and flow environment of a cross-flow rotor. Therefore,
@@
 
-    \caption{Weighted average momentum recovery terms for the actuator disk (AD)
-        simulation from \cite{Bachant2015-JoT}, UNH-RVAT actuator line model with a
-        $k$--$\epsilon$ RANS closure, the two 3-D blade resolved RANS models
-        described in \cite{Bachant2016-BR-CFD}, and the experiments reported in
-        \cite{Bachant2015-JoT}.}
+    \caption{
+        Weighted average momentum recovery terms for the actuator disk (AD)
+        simulation from \cite{Bachant2015-JoT}, UNH-RVAT actuator line model
+        with a $k$--$\epsilon$ RANS closure, the two 3-D blade resolved RANS
+        models described in \cite{Bachant2016-BR-CFD} ($k$--$\omega$ SST and
+        Spalart--Allmaras, SA), and the experiments reported in
+        \cite{Bachant2015-JoT}.
+    }
 
